% =============================================================================
% Section 4: Architecture and Formal Semantics (RESTRUCTURED)
% =============================================================================

\subsection{Document Model: L1--L4 as Conceptual Modelling Layers}\label{subsec:document-model}

ADL Lite organises capability documents into four semantic layers, each with distinct syntax, role, and corresponding event types. The layering separates identity metadata (L1), narrative content (L2), typed assertions (L3), and executable actions (L4), enabling both human readability and machine processing within a single Markdown file. Table~\ref{tab:document-model} summarises the document model.

\begin{table}[htbp]
\centering
\small
\caption{The L1--L4 Document Model. Each layer contributes distinct semantics to the capability document.}
\label{tab:document-model}
\begin{tabular}{@{}p{1.2cm}p{3.2cm}p{4cm}p{4.5cm}@{}}
\toprule
\textbf{Layer} & \textbf{Syntax} & \textbf{Role} & \textbf{Event Type} \\
\midrule
L1 & YAML front matter & Identity metadata (derived snapshot) & SNAPSHOT \\
L2 & Markdown body & Human/LLM narrative prose & --- \\
L3 & \texttt{adl:relation}, \texttt{adl:evidence}, \texttt{adl:seal} & Typed semantic assertions & RELATE, EVIDENCE, SEAL \\
L4 & \texttt{adl:action} & Typed actions with preconditions & REGISTER, VALIDATE, DEPRECATE, \ldots \\
\bottomrule
\end{tabular}
\end{table}

The L1 layer comprises YAML front matter containing identity metadata; critically, these fields are \emph{derived snapshots} recomputed from the EventChain-record (the serialized information content entity, see \S\ref{subsec:two-level-account}). The L2 layer consists of Markdown body prose subject to the Singular Subjectivity Assertion (SSA).

\textbf{Definition (SSA).} The Singular Subjectivity Assertion is a document-level constraint stating that every L2 Markdown body is authored by a single actor at a single time. It is enforced by the parser: any L2 section containing multiple contradictory subjective claims without explicit attribution is flagged as an SSA violation. The SSA ensures that the L2 narrative layer remains traceable to a single source of subjectivity, preventing ``ghost-authored'' reasoning. Contradictory subjective assertions must be recorded as separate L4 events (e.g., \textsc{Evidence} or \textsc{Relate}) by different actors. The SSA interacts with lifecycle audit as follows: (i)~a concept in \texttt{provisional} status may contain SSA-violating prose (it is unvalidated); (ii)~a \textsc{Validate} action's precondition requires that the L2 body pass SSA checking; (iii)~\textsc{Fork} events inherit the parent's L2 body but must satisfy SSA for the new actor's rationale. Basic SHACL shapes are implemented (Appendix~\ref{app:implementation}); full SSA-specific SHACL enforcement and epistemic context learning remain future work (FW6).

The L3 layer introduces typed semantic assertions through fenced blocks supporting predicates such as \texttt{isomorphic-to}, \texttt{specialisation-of}, and \texttt{related-to}. If two agents independently register ``gradient-explosion'' and ``exploding-gradients'', the recommended workflow detects the near-duplicate through string similarity and links them with the strongest applicable predicate (\texttt{isomorphic-to} for structural equivalence). Automated canonicalisation via embedding clustering is deferred to future work (FW7).

\textbf{Invariant 2 (L3 Relation Reconciliation).} When either the source or target concept of a \texttt{RELATE} event undergoes a state transition to \texttt{deprecated} or \texttt{archived}, the relation's validity is determined by the status of both endpoint concepts. A relation is valid only if at least one endpoint is in \texttt{validated} status and neither endpoint is in \texttt{archived} or \texttt{deprecated} status. Let $r$ be a relation with predicate $p$ from concept $C_1$ to concept $C_2$, and let $S(C)$ denote the current lifecycle status of concept $C$. Then:
\begin{align}
    \text{valid}(r) \iff{}& S(C_1) \notin \{\text{archived}, \text{deprecated}\} \land S(C_2) \notin \{\text{archived}, \text{deprecated}\} \nonumber \\
    &{}\land \bigl(S(C_1) = \text{validated} \lor S(C_2) = \text{validated}\bigr).
\end{align}
Relations with at least one endpoint in \texttt{validated} status and no endpoint in \texttt{archived} or \texttt{deprecated} status remain valid for query and traversal. Conversely, relations between two \texttt{provisional} concepts, or where either endpoint is \texttt{deprecated} or \texttt{archived}, are excluded from the warm index. If a forked concept $C'$ is created via a \texttt{FORK} event from $C$, the parent concept retains all existing relations, whereas $C'$ inherits \texttt{isomorphic-to} and \texttt{specialisation-of} links from $C$. It does not inherit \texttt{analogical-to} links, which are re-evaluated during the forking agent's validation workflow. This invariant ensures that the relation graph remains navigable during multi-agent reorganisation and that stale or superseded edges do not pollute semantic queries.

\subsection{EventChain: Core Data Structure}\label{subsec:eventchain}

The EventChain-record (the serialized information content entity, see \S\ref{subsec:two-level-account}) is the fundamental data structure of ADL Lite: an append-only sequence of hash-linked events. An event comprises ten fields: \texttt{event\_id}, \texttt{concept\_id}, \texttt{event\_type}, \texttt{actor}, \texttt{reasoning}, \texttt{timestamp}, \texttt{payload}, \texttt{previous\_event\_id}, \texttt{hash}, and an optional \texttt{signature} (for Ed25519 signature verification, \S\ref{subsec:trust-model}). The event type system is partitioned into lifecycle events $\Sigma_{\text{life}}$ and communication events $\Sigma_{\text{comm}}$:

\begin{align}
    \Sigma_{\text{life}} &= \{\text{REGISTER},\; \text{VALIDATE},\; \text{DEPRECATE},\; \text{FORK},\; \text{ARCHIVE}\} \\
    \Sigma_{\text{comm}} &= \{\text{RELATE},\; \text{EVIDENCE},\; \text{SEAL},\; \text{ANNOUNCE},\; \text{PUBLISH},\; \text{SNAPSHOT}, \nonumber \\
    &\qquad\; \text{CALIBRATE},\; \text{SYNC\_DASHBOARD},\; \text{LISTEN},\; \text{REVOKE}\}
\end{align}

\subsection{Canonical Serialization and Cryptographic Integrity}\label{subsec:crypto}

Each event's hash is computed from canonical JSON serialization with lexicographically sorted keys, six-decimal rounding, and the timestamp. The canonicalization policy is version-locked: a schema version identifier (\texttt{canon-v1}) is included in the hash input, ensuring that future changes to rounding precision or key ordering do not retroactively invalidate existing hashes. Integrity verification $\text{VerifyIntegrity}(C)$ checks four conditions for a chain $C = [e_1, \ldots, e_n]$: (1)~genesis anchoring ($e_1.\text{previous\_event\_id} = \text{null}$), (2)~cryptographic linkage ($e_i.\text{prev\_hash} = e_{i-1}.\text{hash}$), (3)~hash correctness ($e_i.\text{hash} = \text{SHA-256}(\text{Canon}(e_i))$), and (4)~no dangling \texttt{previous\_event\_id} on the genesis event. When $\text{full}=\text{true}$, it additionally verifies archive file hashes against stored \texttt{archive\_pointer} values.\par
\textbf{Genesis hash stability.} The genesis hash is the hash of the first \textsc{Register} event, which includes the \texttt{concept\_id} in its canonical JSON input. Early versions excluded \texttt{concept\_id}, which caused collisions when different concepts shared the same initial payload; we corrected this in E1 (\S\ref{subsec:e1}). The canonical JSON serializer normalises line endings to LF and encodes Unicode as UTF-8, preventing cross-platform drift. The \texttt{canon\_version} field ensures that benign formatting changes (e.g., prettier Markdown reformatting of the L2 body) do not affect event hashes: only the event payload fields participate in the hash computation, not the surrounding Markdown prose.

\textbf{Rehydration vulnerability.} When an EventChain is copied across repositories (e.g., forked on GitHub, cloned to a local machine, or exported to a backup), the genesis hash remains stable \emph{provided} the canonical serialization is preserved. However, two failure modes can fragment identity. \emph{Mode~1 (formatting drift):} if a repository applies different line-ending rules (CRLF vs.\ LF) or changes JSON key ordering, the re-computed hash of an existing event will differ from its stored hash, breaking integrity verification. The \texttt{canon\_version} field mitigates this by pinning the serialization policy. \emph{Mode~2 (metadata injection):} if an external system injects fields (e.g., adding a \texttt{signature} field to legacy events during a Phase~1.5 migration), the canonical JSON changes and the hash breaks. To prevent fragmentation, ADL Lite treats the genesis hash as \emph{write-once}: any migration that adds new fields must store them in subsequent events, not retroactively modify the genesis event. The formal resilience criterion is: a chain $C$ rehydrated in repository $R'$ must satisfy $\text{VerifyIntegrity}(C) = \text{true}$ if and only if $C$ was well-formed in the source repository $R$ and no event content was modified during transfer.

\subsubsection{Dual-Identifier Design: Content Addressability}
\label{subsubsec:dual-identifier}

ADL Lite uses a \emph{dual-identifier} scheme that separates \emph{provenance identity} (\emph{who} created the concept and \emph{when}) from \emph{content identity} (\emph{what} the concept semantically describes). This addresses the identity-proliferation problem in multi-agent authoring. The genesis hash is a function of the timestamp and the actor's UUID. It creates a unique identifier for every registration event, even when two agents independently register semantically identical concepts.

\textbf{Genesis hash (provenance identity).} The genesis hash $h_{\text{gen}} = \text{SHA-256}(\text{Canon}(e_{\text{REG}}))$ is the hash of the first \textsc{Register} event. It includes the \texttt{concept\_id} (human-readable alias), the actor's self-declared identifier, and the creation timestamp. The genesis hash is \emph{unique per registration event}: two agents registering ``gradient-explosion'' at different times will have different genesis hashes. This is the \emph{primary identifier} for the EventChain and is used for all lifecycle audit.

\textbf{Content hash (content identity).} The content hash $h_{\text{content}} = \text{SHA-256}(\text{Canon}(\text{L2\_body}, \text{L3\_blocks}))$ is the hash of the L2 Markdown body and the L3 relation blocks. It \emph{excludes} all metadata that varies between registrations: \texttt{concept\_id}, actor, timestamp, and event-specific fields. The content hash is \emph{stable across independent registrations}: two agents describing the same capability in their own words will have the same content hash if their L2 and L3 descriptions are identical (or near-identical, within a configurable Levenshtein threshold).

\textbf{Near-duplicate detection.} When a new \textsc{Register} event is appended, the system computes both $h_{\text{gen}}$ and $h_{\text{content}}$. If $h_{\text{content}}$ matches an existing concept (exact match or within a Levenshtein ratio threshold of 0.85), the ActionExecutor emits a warning: ``This concept may be a duplicate of [existing concept]. Consider appending a \textsc{Relate} event with predicate \texttt{isomorphic-to}.'' This recommendation is advisory: the agent may choose to register the concept as a separate entity (e.g., because the L2 description is substantially different, or because the agent is intentionally creating a fork). The content hash is stored in the event payload and is visible in the warm index for query purposes.

\textbf{Relation to identity consolidation.} The dual-identifier scheme provides an \emph{automated} mechanism for detecting near-duplicates, which complements the manual \textsc{Relate} event workflow (\S\ref{subsec:compact-formal-spec}). When the system detects a content-hash match, it can suggest the strongest applicable predicate: \texttt{isomorphic-to} for structural equivalence, \texttt{specialisation-of} for refinement, or \texttt{analogical-to} for metaphorical similarity. The content hash is computed in $O(|\text{L2}| + |\text{L3}|)$ time by the parser. It is \emph{not} part of the cryptographic chain: it is an advisory field, not a consensus field, so it does not affect the well-formedness or integrity of the EventChain.

\textbf{Formal resilience criteria and DID vs.~content-linked PID comparison.} We state three formal criteria that any identity mechanism for ADL Lite must satisfy: (i)\ \emph{Provenance uniqueness:} two independent registrations must yield distinct identifiers even if their L2/L3 content is identical; (ii)\ \emph{Rehydration stability:} copying a chain across repositories must preserve its identifier unless the event content is modified; (iii)\ \emph{Content-addressability option:} the system must support a secondary content-derived identifier for near-duplicate detection. The genesis hash $h_{\text{gen}}$ satisfies (i) and (ii) because it includes the actor, timestamp, and UUID; the content hash $h_{\text{content}}$ satisfies (iii). A content-addressed PID (e.g., IPFS CID, Merkle DAG root) would satisfy (ii) and (iii) but fail (i) if two agents register identical content independently, because they would receive the same PID and collide. A DID-based identifier (e.g., \texttt{did:key} public key) satisfies (i) (each actor has a unique key) but fails (ii) because DIDs are not sensitive to the L2/L3 content: rotating a DID does not change the capability description. The dual-identifier scheme is therefore the minimal design that satisfies all three criteria simultaneously. Table~\ref{tab:identity-comparison} summarises the comparison.

\begin{table}[htbp]
\centering
\small
\caption{Identity mechanism comparison: genesis hash, content hash, DID, and content-addressed PID.}
\label{tab:identity-comparison}
\begin{tabularx}{\textwidth}{@{}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}X@{}}
\toprule
\textbf{Criterion} & \textbf{$h_{\text{gen}}$} & \textbf{$h_{\text{content}}$} & \textbf{DID} & \textbf{Content PID} \\
\midrule
Provenance uniqueness & $\checkmark$ & $\times$ & $\checkmark$ & $\times$ \\
Rehydration stability & $\checkmark$ & $\checkmark$ & $\times$ & $\checkmark$ \\
Content-addressability & $\times$ & $\checkmark$ & $\times$ & $\checkmark$ \\
Cryptographic binding & $\checkmark$ & $\times$ & $\checkmark$ & $\checkmark$ \\
Human readability & $\times$ & $\times$ & $\checkmark$ & $\times$ \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Precondition Language and Action Executor}\label{subsec:precondition-executor}

\subsubsection{Formal Precondition Language}\label{subsubsec:precondition-language}

L4 action blocks are validated against preconditions expressed in a constrained, decidable language. A precondition rule is a triple $\langle \text{field}, \text{comparator}, \text{value} \rangle$ with comparator alphabet $\mathcal{K} = \{\text{EQ}, \text{NEQ}, \text{GT}, \text{GTE}, \text{LT}, \text{LTE}, \text{IN}, \text{EXISTS}\}$. Table~\ref{tab:comparator-semantics} gives the formal semantics.

\begin{table}[htbp]
\centering
\small
\caption{Comparator semantics: explicit operator mapping to ground predicates.}
\label{tab:comparator-semantics}
\begin{tabular}{@{}p{2.5cm}p{3cm}p{5.5cm}@{}}
\toprule
\textbf{Comparator} & \textbf{Predicate} & \textbf{FOL encoding} \\
\midrule
EQ & Equality & $x = y$ \\
NEQ & Disequality & $\neg(x = y)$ \\
GT & Greater than & $x > y$ \\
GTE & Greater than or equal & $x \geq y$ \\
LT & Less than & $x < y$ \\
LTE & Less than or equal & $x \leq y$ \\
IN & Set membership & $x \in Y$ (where $Y$ is a finite set) \\
EXISTS & Non-null check & $\exists x \,:\, x \neq \text{null}$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{BNF grammar.} Let $\mathcal{N}$ be the set of alphanumeric field identifiers, $\mathcal{S} = \mathbb{R} \cup \mathbb{B} \cup \Sigma^*$ the set of scalar literals (reals, booleans, strings), and $\mathcal{P}_{\text{fin}}(\mathcal{S})$ the finite power set of scalars. The precondition language $\mathcal{L}_{\text{pre}}$ is generated by the grammar $G = (\mathcal{V}, \Sigma_{\text{term}}, R, \text{rule\_list})$ with non-terminals $\mathcal{V} = \{\text{rule\_list}, \text{rule}, \text{field}, \text{comparator}, \text{value}\}$, terminals $\Sigma_{\text{term}} = \mathcal{N} \cup \mathcal{S} \cup \{\text{EQ}, \text{NEQ}, \text{GT}, \text{GTE}, \text{LT}, \text{LTE}, \text{IN}, \text{EXISTS}, \text{AND}, \text{OR}, \text{null}\}$, and production rules $R$:
{\small
\begin{align*}
    \text{rule\_list} &::= \text{rule} \;\mid\; \text{rule\_list} \; \text{AND} \; \text{rule\_list} \;\mid\; \text{rule\_list} \; \text{OR} \; \text{rule\_list} \\
    \text{rule} &::= \langle \text{field}, \text{comparator}, \text{value} \rangle \\
    \text{field} &::= f \in \mathcal{F} \quad (\text{the finite set of snapshot field names}) \\
    \text{comparator} &::= \text{EQ} \mid \text{NEQ} \mid \text{GT} \mid \text{GTE} \mid \text{LT} \mid \text{LTE} \mid \text{IN} \mid \text{EXISTS} \\
    \text{value} &::= s \in \mathcal{S} \mid P \in \mathcal{P}_{\text{fin}}(\mathcal{S}) \mid \text{null}
\end{align*}
}

\textbf{Formal semantics.} Let $\mathcal{D}$ be the domain of values (scalars, sets, and $\text{null}$). Let $\text{Snapshot}(C)$ be the derived snapshot of chain $C$, a finite map from field names to $\mathcal{D}$. The semantic function $\text{eval}(r, C) \in \{\text{true}, \text{false}\}$ is defined for a rule $r = \langle f, \kappa, v \rangle$ by:
\begin{equation}
    \text{eval}(\langle f, \kappa, v \rangle, C) = \text{apply}(\kappa, \text{lookup}(f, C), v)
\end{equation}
where $\text{lookup}(f, C) = \text{Snapshot}(C)[f]$ (the value of field $f$ in the snapshot, or $\text{null}$ if absent), and $\text{apply}(\kappa, x, y)$ is the comparator-specific predicate from Table~\ref{tab:comparator-semantics}. The snapshot is pre-computed in $\mathcal{O}(n)$ time by a single scan of $C$; subsequent rule evaluations are $\mathcal{O}(1)$.

\textbf{Composition semantics.} A rule list $R = [r_1, \ldots, r_k]$ is evaluated as a conjunction: $\text{eval}(R, C) = \bigwedge_{i=1}^{k} \text{eval}(r_i, C)$. Evaluation is short-circuit: if $\text{eval}(r_j, C) = \text{false}$ for any $j < k$, the remaining rules are not evaluated. The language is a variable-free ground fragment: each rule is a single atomic comparison over a cached derived snapshot, with no quantification, no recursion, and no dynamic code execution. The conjunction of $k$ rules is a variable-free conjunction of ground atoms, evaluated in $\mathcal{O}(k)$ time (Theorem~8).

Precondition evaluation is pure and referentially transparent: it references only the snapshot of the chain $C$ being validated, and does not access external chains, databases, or network resources. Cross-chain preconditions (e.g., ``allow \textsc{Validate} only if concept $X$ is \texttt{validated}'') are not supported in the core language; they are deferred to future work (FW2). This local-scope restriction ensures that precondition evaluation is deterministic and executable without consensus or network access.

\subsubsection{Formal Syntax and Semantics}\label{subsubsec:formal-syntax-semantics}

A precondition rule is a triple $\rho = \langle \text{field}, \text{comparator}, \text{value} \rangle$ where $\text{field} \in \mathcal{F}$ (the set of snapshot fields), $\text{comparator} \in \{\text{EQ}, \text{NEQ}, \text{GT}, \text{GTE}, \text{LT}, \text{LTE}, \text{IN}, \text{EXISTS}\}$, and $\text{value} \in \mathcal{V}$ (a scalar, a finite list, or $\text{null}$). The semantics is defined relative to a cached derived snapshot $\sigma = \text{Snapshot}(C)$, a finite map from field names to values:

\begin{equation}\label{eq:precondition-semantics}
    \llbracket \rho \rrbracket(\sigma) = \text{compare}(\sigma[\text{field}], \text{comparator}, \text{value})
\end{equation}

\textbf{Typing rules.} A rule $\langle f, \kappa, v \rangle$ is well-typed, denoted $\vdash \langle f, \kappa, v \rangle : \text{bool}$, iff $\text{type}(f) \in \mathcal{T}$ and the comparator--value pair satisfies the compatibility predicate:
{\small
\begin{align*}
    \vdash \langle f, \text{EQ}, v \rangle &\iff \text{type}(f) = \text{type}(v) \land v \in \mathcal{S} \\
    \vdash \langle f, \text{NEQ}, v \rangle &\iff \text{type}(f) = \text{type}(v) \land v \in \mathcal{S} \\
    \vdash \langle f, \text{GT}, v \rangle &\iff \text{type}(f) = \mathbb{R} \land v \in \mathbb{R} \\
    \vdash \langle f, \text{GTE}, v \rangle &\iff \text{type}(f) = \mathbb{R} \land v \in \mathbb{R} \\
    \vdash \langle f, \text{LT}, v \rangle &\iff \text{type}(f) = \mathbb{R} \land v \in \mathbb{R} \\
    \vdash \langle f, \text{LTE}, v \rangle &\iff \text{type}(f) = \mathbb{R} \land v \in \mathbb{R} \\
    \vdash \langle f, \text{IN}, v \rangle &\iff v \in \mathcal{P}_{\text{fin}}(\mathcal{S}) \land \text{type}(f) = \text{type}(v_i) \text{ for all } v_i \in v \\
    \vdash \langle f, \text{EXISTS}, \text{null} \rangle &\iff \text{type}(f) \in \mathcal{T}
\end{align*}
}
where $\mathcal{T} = \{\mathbb{R}, \mathbb{B}, \Sigma^*, \mathcal{P}_{\text{fin}}(\mathcal{S})\}$ is the set of types. The ActionExecutor rejects any rule that violates $\vdash$ before evaluation. This static typing guarantees that $\llbracket \rho \rrbracket(\sigma)$ never encounters a runtime type error.

\paragraph{Safety.} The precondition language contains no \texttt{eval()}, no \texttt{exec()}, and no dynamic code execution. Comparator dispatch is a closed enumeration. The Python implementation uses a fixed \texttt{Comparator} enum with eight explicit cases. Adding a new comparator requires modifying the enum definition and adding a static branch in the dispatch table. This closed-world design guarantees that no user-supplied string can be interpreted as code.

\paragraph{Formal evaluation rules.} The evaluation function is defined by exhaustive case analysis on the comparator:

\begin{align}
    \text{eval}(\sigma, \langle f, \text{EQ}, v \rangle) &= (\sigma.f = v) \\
    \text{eval}(\sigma, \langle f, \text{GT}, v \rangle) &= (\sigma.f > v) \\
    \text{eval}(\sigma, \langle f, \text{IN}, vs \rangle) &= (\sigma.f \in vs) \\
    \text{eval}(\sigma, \langle f, \text{EXISTS}, \text{null} \rangle) &= (\sigma.f \neq \text{null})
\end{align}

The remaining comparators (NEQ, GTE, LT, LTE) follow analogously. Because $\sigma$ is a precomputed hash map, field lookup is $O(1)$; because the comparator cases are a closed enumeration of primitive operations, each comparison is $O(1)$.

\subsubsection{Complexity Analysis}\label{subsubsec:complexity-analysis}

\paragraph{Theorem (Precondition Evaluation Complexity).} For any well-formed EventChain $C$ with cached snapshot $\sigma$ and any list of $k$ precondition rules $R = [r_1, \ldots, r_k]$, total evaluation time is $O(k)$, with $O(1)$ per rule and $O(1)$ auxiliary space.

\emph{Proof Sketch.} Each rule performs a single field lookup in $\sigma$ (a hash map, $O(1)$) and a closed comparison (a primitive operation, $O(1)$). The rule list is evaluated as a short-circuit conjunction: evaluation terminates at the first false rule. There is no quantification, no recursion, no backtracking, and no global state access beyond the snapshot $\sigma$. The snapshot $\sigma$ is precomputed from the chain in $O(n)$ at append time (or cached from a previous append), so the per-rule cost is $O(1)$ amortized over the lifetime of the chain. $\square$

\paragraph{Empirical validation.} Experiment E34 (Precondition Formalization Benchmark) empirically confirms the $O(1)$ per-rule claim: per-rule evaluation measures $0.8$--$1.0\,\mu\text{s}$, independent of chain length (tested up to $1{,}000$ events). This validates that the closed comparator dispatch and hash-map field lookup dominate the cost, with no superlinear term emerging from large snapshots.

\begin{table}[htbp]
\centering
\footnotesize
\caption{Precondition language comparison: ADL Lite vs.\ full first-order logic (UFO-B institutional rules) vs.\ OWL-SWRL.}
\label{tab:precondition-comparison}
\begin{tabularx}{\textwidth}{@{}p{2.8cm}p{4.2cm}p{4.2cm}X@{}}
\toprule
\textbf{Property} & \textbf{ADL Lite} & \textbf{UFO-B / FOL} & \textbf{OWL-SWRL} \\
\midrule
Syntax & Ground atomic triples $\langle f, \kappa, v \rangle$ & Quantified first-order formulas & DL-safe rules over OWL axioms \\
Semantics & Snapshot lookup + primitive comparison & Model-theoretic Tarski semantics & DL semantics + rule engine \\
Quantification & None (ground fragment) & $\forall$, $\exists$, arbitrary nesting & DL-safe (limited quantification) \\
Recursion & Not expressible & Unrestricted recursion & Not expressible (acyclic) \\
Eval complexity & $O(1)$ per rule, $O(k)$ per list & Undecidable (general FOL) / NExpTime (DL) & NExpTime-complete (OWL-DL) \\
Safety & No \texttt{eval()}, closed enum & Depends on implementation & Depends on reasoner \\
Deployment & Single Python module ($\approx$200 LOC) & Full theorem prover / solver & Full OWL stack (HermiT/Pellet) \\
\bottomrule
\end{tabularx}
\end{table}

\subsubsection{Precondition Language Expressivity}

The precondition language is intentionally minimal: a variable-free ground fragment with eight comparators, no recursion, no quantification, and no dynamic code execution. This minimalism guarantees decidability and $\mathcal{O}(1)$ per-rule evaluation (Theorem~8), but it also limits expressivity. Table~\ref{tab:precondition-expressivity} classifies the expressivity boundaries.

\begin{table}[htbp]
\centering
\footnotesize
\caption{Precondition language expressivity: what can and cannot be expressed.}
\label{tab:precondition-expressivity}
\begin{tabularx}{\textwidth}{@{}p{3.2cm}p{4.5cm}X@{}}
\toprule
\textbf{Constraint Category} & \textbf{Examples} & \textbf{Expressible?} \\
\midrule
Field-value comparison & \texttt{status EQ provisional}, \texttt{confidence GTE 0.5} & $\checkmark$~(Core) \\
Set membership & \texttt{tags IN [verified, reviewed]} & \checkmark~(Core) \\
Existence & \texttt{actor EXISTS} (field defined in snapshot) & \checkmark~(Core) \\
Conjunction / disjunction & \texttt{status EQ provisional AND confidence GTE 0.5} & \checkmark~(Core) \\
Temporal predicate & \texttt{7 days passed since last VALIDATE} & $\times$~(Not in core; application layer) \\
Cross-chain reference & \texttt{parent.status EQ validated} & $\times$~(Not in core; deferred to FW2) \\
Quantification / aggregation & \texttt{at least 3 validators approved} & $\times$~(Handled by $\gamma$, not preconditions) \\
Arithmetic / string manipulation & \texttt{confidence within 10\% of mean} & $\times$~(Not in core) \\
Recursive / self-referential & \texttt{no concept is its own ancestor} & $\times$~(Not in core) \\
\bottomrule
\end{tabularx}
\end{table}

\textbf{Temporal predicates.} Time is not a first-class value in the precondition language. The ActionExecutor can enforce temporal constraints at the application layer (e.g., a minimum delay between validation and deprecation). The precondition language's scope is restricted to snapshot-based guards to ensure decidability and $\mathcal{O}(1)$ evaluation time per rule.

\textbf{Cross-chain predicates.} Not supported by design. The precondition language evaluates only the snapshot of the chain being modified. This ensures referential transparency and eliminates the need for distributed consensus during precondition evaluation. Cross-chain references would require external chain lookup, which introduces non-determinism (the target chain may change between evaluation and append).

\textbf{Theorem~8 (Precondition Evaluation Termination and Complexity).} For any well-formed EventChain $C$ and any precondition rule $r$, $\text{eval}(r, C)$ terminates in $\mathcal{O}(1)$ time (assuming the derived snapshot is cached). For a list of $k$ rules, total time is $\mathcal{O}(k)$ and space is $\mathcal{O}(1)$. \emph{Proof Sketch:} Both field lookup and comparator dispatch are $\mathcal{O}(1)$; the language contains no recursion or quantification. See Appendix~E for the full decidability proof (Theorem~\ref{thm:precondition-decidability}).

Table~\ref{tab:complexity} summarises the complexity; the decidability and complexity-class discussion (membership in \textbf{P}) is in Appendix~H.

\begin{table}[htbp]
\centering
\small
\caption{Computational complexity of precondition evaluation.}
\label{tab:complexity}
\begin{tabularx}{\textwidth}{@{}p{5cm}llX@{}}
\toprule
\textbf{Operation} & \textbf{Time} & \textbf{Space} & \textbf{Notes} \\
\midrule
Single rule $\text{eval}(r, C)$ & $\mathcal{O}(1)$ & $\mathcal{O}(1)$ & No recursion, no quantification \\
Rule list $\text{eval}(R, C)$, $|R| = k$ & $\mathcal{O}(k)$ & $\mathcal{O}(1)$ & Short-circuit on first failure \\
Field lookup $\text{lookup}(f, C)$ & $\mathcal{O}(1)$ & $\mathcal{O}(1)$ & Memoized snapshot or dict lookup \\
Comparator dispatch $k(x, y)$ & $\mathcal{O}(1)$ & $\mathcal{O}(1)$ & Closed 8-case enumeration \\
\bottomrule
\end{tabularx}
\end{table}

\textbf{Complete lifecycle transition matrix.} Table~\ref{tab:lifecycle-matrix} lists the ActionExecutor preconditions for each lifecycle event. Under the CRDT LUB semantics, the final status is the maximum over all lifecycle events in the chain, not the last event. However, the ActionExecutor still enforces preconditions to prevent semantically nonsensical sequences (e.g., archiving a concept that has never been validated). The preconditions for each action are encoded in \texttt{adl\_core\_ontology.yaml}.

\begin{table}[htbp]
\centering
\footnotesize
\caption{Lifecycle event preconditions: permitted events from current status (rows). The final status is the LUB of all lifecycle events in the chain (CRDT semantics).}
\label{tab:lifecycle-matrix}
\begin{tabular}{@{}p{2cm}ccccc@{}}
\toprule
\textbf{Status} & \textsc{Reg.} & \textsc{Val.} & \textsc{Dep.} & \textsc{Fork} & \textsc{Arch.} \\
\midrule
provisional & $\times$ (gen.) & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
validated & $\times$ & $\checkmark$ (re) & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
deprecated & $\times$ & $\checkmark$ & $\times$ & $\times$ & $\checkmark$ \\
forked & $\times$ & $\checkmark$ & $\checkmark$ & $\times$ & $\checkmark$ \\
archived & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
\bottomrule
\end{tabular}
\end{table}

\textsc{Register} is permitted only as a genesis event (creating a new concept). From \texttt{validated}, \textsc{Validate} is permitted as a re-validation (adding a new confidence assertion). Under CRDT LUB semantics, once a chain reaches \texttt{deprecated}, subsequent \textsc{Validate} events do not change the status back to \texttt{validated} (the LUB of \texttt{deprecated} and \texttt{validated} is \texttt{deprecated}), but they can still be appended as auditable evidence. Re-activation after deprecation requires a new \textsc{Register} event (new genesis), not a reversal of the existing chain.

\subsubsection{Action Executor and Transition Engine}\label{subsubsec:action-consensus}

The ActionExecutor validates L4 action blocks against a closed action registry defined in \path{adl\_core\_ontology.yaml}. Each action definition specifies preconditions as a list of \texttt{PreconditionRule} objects. The ConsensusEngine governs the resulting status transitions by appending events to the chain; the final status is derived via the CRDT LUB over all lifecycle events (Theorem~1), not by a transition matrix. Forking creates a new child chain with a fresh \texttt{REGISTER} event while the original chain's status is the LUB of its previous status and \texttt{forked}; both evolve independently.

\subsection{Compact Formal Specification}\label{subsec:compact-formal-spec}

This subsection presents a \emph{single-page, self-contained} formal specification of ADL Lite's derivation semantics. All definitions, theorems, and complexity claims are stated here; expanded proof sketches and sensitivity analysis appear in Appendix~E. The specification is organised as: (i)~the status lattice and derivation function~$\delta$; (ii)~the confidence aggregation function~$\gamma$ (three variants); (iii)~the precondition language; and (iv)~the fork and merge semantics.

\subsubsection{Status Lattice and Derivation Function $\delta$}\label{subsubsec:status-lattice}

Let $\Sigma = \Sigma_{\text{life}} \cup \Sigma_{\text{comm}}$ be the event alphabet. The lifecycle events are $\Sigma_{\text{life}} = \{\text{REGISTER},\; \text{VALIDATE},\; \text{DEPRECATE},\; \text{FORK},\; \text{ARCHIVE}\}$. The communication events are $\Sigma_{\text{comm}} = \{\text{RELATE},\; \text{EVIDENCE},\; \text{SEAL},\; \text{ANNOUNCE},\; \text{PUBLISH},\; \text{SNAPSHOT},\; \text{CALIBRATE},\; \text{SYNC\_DASHBOARD},\; \text{LISTEN},\; \text{REVOKE}\}$. A chain $C = [e_1, \ldots, e_n]$ is well-formed, denoted $\text{WF}(C)$, if it satisfies 13 axioms: genesis anchoring, shared \texttt{concept\_id}, cryptographic linkage, hash correctness, distinct \texttt{event\_id}, non-empty actor, timestamp monotonicity, payload schema conformance, action preconditions, scope ACL, signature verification, confidence clamping, and valid event type (the full list is in Appendix~E).

The status space is $S = \{\text{provisional},\; \text{forked},\; \text{validated},\; \text{deprecated},\; \text{archived}\}$, equipped with the total order:
\begin{equation}\label{eq:status-order}
    \text{provisional} \;\prec\; \text{forked} \;\prec\; \text{validated} \;\prec\; \text{deprecated} \;\prec\; \text{archived}
\end{equation}
numbered $1, 2, 3, 4, 5$ respectively. This is a join-semilattice: the least upper bound (LUB) of any subset is its maximum element. The status map $f: \Sigma_{\text{life}} \to S$ assigns $f(\text{REGISTER}) = \text{provisional}$, $f(\text{VALIDATE}) = \text{validated}$, $f(\text{DEPRECATE}) = \text{deprecated}$, $f(\text{FORK}) = \text{forked}$, $f(\text{ARCHIVE}) = \text{archived}$.

Let $C_{\text{life}} = \{e \in C \mid e.\tau \in \Sigma_{\text{life}}\}$ be the lifecycle subsequence. The status derivation function $\delta: \{C \mid \text{WF}(C)\} \to S$ is:
\begin{equation}\label{eq:delta-def}
    \delta(C) = \begin{cases}
        \text{provisional} & \text{if } C_{\text{life}} = \emptyset \\
        \max_{\prec}\{ f(e.\tau) \mid e \in C_{\text{life}} \} & \text{otherwise}
    \end{cases}
\end{equation}
where $\max_{\prec}$ is the maximum over the integer order (Eq.~\ref{eq:status-order}). Complexity: $O(|C|)$ by single scan of $C$ (or $O(1)$ with an incremental LUB cache updated on every \texttt{append}).

Under Eq.~\ref{eq:delta-def}, status is determined by the \emph{highest} lifecycle event in the lattice, not the last. Once $\text{deprecated}$ or $\text{archived}$ appears, no subsequent event can lower the status (monotonicity). Re-activation after deprecation requires a new genesis event on a new chain.

\subsubsection{Formal Lattice Structure}\label{subsubsec:formal-lattice}

The status space $S = \{\text{provisional}, \text{forked}, \text{validated}, \text{deprecated}, \text{archived}\}$ equipped with the total order $\prec$ (Eq.~\ref{eq:status-order}) forms a complete lattice $(S, \preceq, \sqcap, \sqcup, \bot, \top)$ where:
\begin{itemize}
    \item $\bot = \text{provisional}$ is the least element,
    \item $\top = \text{archived}$ is the greatest element,
    \item the join operator $\sqcup: S \times S \to S$ is defined by $\sqcup = \text{status\_max} = \max_{\prec}$, the least upper bound under $\prec$,
    \item the meet operator $\sqcap: S \times S \to S$ is defined by $\sqcap = \text{status\_min} = \min_{\prec}$, the greatest lower bound under $\prec$.
\end{itemize}
Because $\prec$ is a total order on a finite set, every subset $X \subseteq S$ (finite or infinite, though $S$ itself is finite) has both a maximum element $\max_{\prec} X$ and a minimum element $\min_{\prec} X$. Therefore every subset has a least upper bound and a greatest lower bound, and $(S, \preceq, \sqcap, \sqcup, \bot, \top)$ is a complete lattice~\cite{davey_priestley_2002}. The completeness is essential for CRDT convergence: the merge of any set of concurrent branches computes $\bigsqcup X$ for the lifecycle events $X$ of the merged chain, which is guaranteed to exist and be unique (Theorem~9).

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.2, every node/.style={font=\small}]
    \node (arch) at (0,4) {\textsc{archived}};
    \node (dep)  at (0,3) {\textsc{deprecated}};
    \node (val)  at (0,2) {\textsc{validated}};
    \node (fork) at (0,1) {\textsc{forked}};
    \node (prov) at (0,0) {\textsc{provisional}};
    \draw (prov) -- (fork);
    \draw (fork) -- (val);
    \draw (val) -- (dep);
    \draw (dep) -- (arch);
\end{tikzpicture}
\caption{Hasse diagram of the status lattice $(S, \preceq, \sqcup, \bot, \top)$. The diagram is a linear chain because $\prec$ is a total order.}
\label{fig:status-hasse}
\end{figure}

\paragraph{Join-semilattice proof.} The join operator $\sqcup$ satisfies five algebraic properties:

\begin{enumerate}[label=(\roman*)]
    \item \emph{Associativity:} $(a \sqcup b) \sqcup c = a \sqcup (b \sqcup c)$, because $\max(\max(a,b),c) = \max(a,\max(b,c))$.
    \item \emph{Commutativity:} $a \sqcup b = b \sqcup a$, because $\max(a,b) = \max(b,a)$.
    \item \emph{Idempotence:} $a \sqcup a = a$, because $\max(a,a) = a$.
    \item \emph{Identity:} $a \sqcup \bot = a$ for all $a \in S$, because $\max(a, \text{provisional}) = a$ (provisional is the minimum element).
    \item \emph{Monotonicity:} if $a \preceq b$ then $a \sqcup c \preceq b \sqcup c$, because $\max(a,c) \preceq \max(b,c)$ when $a \preceq b$.
\end{enumerate}

These properties are machine-verified in Coq (\texttt{Status.v}, Theorem~T3) and hold for the CRDT merge semantics (Theorem~T9). The identity property~(iv) is essential for the incremental LUB cache: appending a non-lifecycle event (which contributes $\bot$) leaves the status unchanged, justifying $O(1)$ cache updates.

\paragraph{Meet-semilattice properties.} The meet operator $\sqcap$ satisfies the dual five properties:
\begin{enumerate}[label=(\roman*)]
    \item \emph{Associativity:} $(a \sqcap b) \sqcap c = a \sqcap (b \sqcap c)$, because $\min(\min(a,b),c) = \min(a,\min(b,c))$.
    \item \emph{Commutativity:} $a \sqcap b = b \sqcap a$, because $\min(a,b) = \min(b,a)$.
    \item \emph{Idempotence:} $a \sqcap a = a$, because $\min(a,a) = a$.
    \item \emph{Identity:} $a \sqcap \top = a$ for all $a \in S$, because $\min(a, \text{archived}) = a$ (archived is the maximum element).
    \item \emph{Monotonicity:} if $a \preceq b$ then $a \sqcap c \preceq b \sqcap c$, because $\min(a,c) \preceq \min(b,c)$ when $a \preceq b$.
\end{enumerate}
The meet operator is not used directly in status derivation (which is a join over all lifecycle events), but it is required for the complete lattice structure and supports future query semantics such as ``find the least advanced status among a set of concepts.''

\begin{table}[htbp]
\centering
\small
\caption{Status join table ($\sqcup = \max_{\prec}$). The entry at row $i$, column $j$ is $i \sqcup j$.}
\label{tab:join-table}
\begin{tabular}{@{}lccccc@{}}
\toprule
$\sqcup$ & \textsc{prov.} & \textsc{forked} & \textsc{val.} & \textsc{dep.} & \textsc{arch.} \\
\midrule
\textsc{provisional} & \textsc{prov.} & \textsc{forked} & \textsc{val.} & \textsc{dep.} & \textsc{arch.} \\
\textsc{forked} & \textsc{forked} & \textsc{forked} & \textsc{val.} & \textsc{dep.} & \textsc{arch.} \\
\textsc{validated} & \textsc{val.} & \textsc{val.} & \textsc{val.} & \textsc{dep.} & \textsc{arch.} \\
\textsc{deprecated} & \textsc{dep.} & \textsc{dep.} & \textsc{dep.} & \textsc{dep.} & \textsc{arch.} \\
\textsc{archived} & \textsc{arch.} & \textsc{arch.} & \textsc{arch.} & \textsc{arch.} & \textsc{arch.} \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Confidence Aggregation: Three Variants}

Let $V = [e \in C \mid e.\tau = \text{VALIDATE}]$ denote the validation subsequence. ADL Lite provides three confidence aggregation strategies.

\textbf{Variant 1: Default (G-Counter max).} The simplest strategy, used by default, takes the maximum confidence across all \texttt{VALIDATE} events (or \texttt{SNAPSHOT} if no validation has occurred):
\begin{equation}\label{eq:gamma-default}
    \gamma_{\text{default}}(C) = \begin{cases}
        0.0 & \text{if } V = [] \\
        \text{clamp}_{[0,1]}\bigl(\max\{ e.\text{payload}[\text{``confidence''}] \mid e \in V \}\bigr) & \text{otherwise}
    \end{cases}
\end{equation}
where $\max$ is the maximum over all validation events in the chain (G-Counter semantics), and $\text{clamp}_{[0,1]}(x) = \max(0, \min(1, x))$. It is evaluated in $O(1)$ time by an incremental cache (\texttt{\_cached\_confidence}) that updates on every \texttt{append} to $\max(\text{cache}, \text{clamp}(c))$. The canonical state is always recomputable from the event sequence in $O(|V|) \leq O(n)$ time; the $O(1)$ claim refers to the memoized warm-index query, not the canonical derivation.

\textbf{Variant 2: Bonus-formula aggregate ($\gamma_{\text{agg}}$).} This variant mitigates collusion by aggregating per-actor maxima and adding a bonus for cross-validation:
\begin{equation}\label{eq:gamma-agg}
    \gamma_{\text{agg}}(C) = \begin{cases}
        0.0 & \text{if } V = [] \\
        \min\!\bigl(1.0,\; c_{\text{base}} + 0.05 \times (N_{\text{vals}} - 1)\bigr) & \text{otherwise}
    \end{cases}
\end{equation}
where $N_{\text{vals}} = |\text{actors}(V)|$ and $c_{\text{base}} = \max(0.5, \frac{1}{N_{\text{vals}}} \sum_{a \in \text{actors}(V)} \phi(a, V))$, with $\phi(a, V) = \max\{e.\text{payload}[\text{``confidence''}] \mid e \in V \land e.\text{actor} = a\}$. The $0.05$ bonus per distinct validator rewards cross-validation; the $0.5$ floor prevents a single low-confidence validator from dominating. A sensitivity analysis appears in Appendix~E.

\textbf{Variant 3: Calibrated confidence ($\gamma_{\text{cal}}$).} This variant weights each validator's confidence by a per-actor accuracy score:
\begin{equation}\label{eq:gamma-cal}
    \gamma_{\text{cal}}(C) = \begin{cases}
        0.0 & \text{if } V = [] \\
        \text{clamp}_{[0,1]}\!\Bigl(\frac{\sum_{a \in \text{actors}(V)} \text{accuracy}_a \cdot \phi(a, V)}{\sum_{a \in \text{actors}(V)} \text{accuracy}_a}\Bigr) & \text{otherwise}
    \end{cases}
\end{equation}
where $\text{accuracy}_a \in [0,1]$ is a per-actor accuracy score from an external calibration profile. Complexity: $O(|V|)$. All three variants are bounded in $[0,1]$ by construction (Theorem~4).

\textbf{Concurrent event resolution.} For $\delta$: concurrent lifecycle events commute because the LUB is independent of order (Eq.~\ref{eq:delta-def}). For $\gamma_{\text{default}}$: concurrent \textsc{Validate} events commute because $\max$ is associative and commutative. For $\gamma_{\text{agg}}$ and $\gamma_{\text{cal}}$: per-actor maxima are computed first, then aggregated; concurrent events from the same actor commute because $\max$ is idempotent. The G-Counter semantics guarantee that once a high-confidence validation is recorded, subsequent lower-confidence assertions cannot decrease the aggregate.

\textbf{Phase~1 limitations.} The current collaborative-audit model lacks authenticated identity, so $\gamma_{\text{cal}}$ accuracy scores are self-reported and do not prevent collusion (a single actor can report $\text{accuracy}_a = 1.0$). Sybil attacks are possible: a single actor with $10$ pseudonyms can drive $\gamma_{\text{agg}}$ to $0.99$ (since $c_{\text{base}} \geq 0.5$ and $0.05 \times 9 = 0.45$). These are known limitations (L3, \S\ref{subsec:limitations}), demonstrated in adversarial experiment E14 (Table~\ref{tab:undetectable-attacks}). Mitigation requires authenticated identities and staking (Phase~3, FW9).

\subsubsection{Confidence as G-Counter CRDT}\label{subsubsec:confidence-gcounter}

The confidence function $\gamma_{\text{default}}(C)$ is a state-based G-Counter CRDT. Let $C$ be a chain and $V \subseteq C$ the subsequence of \textsc{Validate} and \textsc{Snapshot} events. The confidence is defined as:
\begin{equation}\label{eq:confidence-gcounter}
    \gamma(C) = \max\bigl(\{c \mid \exists e \in V : e.\text{payload}[\text{``confidence''}] = c\} \cup \{0\}\bigr)
\end{equation}

\paragraph{State-based G-Counter formalisation.} In the state-based CRDT model, the confidence G-Counter is defined by the tuple $(\Sigma, \text{inc}, \text{merge}, s_0)$ where:
\begin{itemize}
    \item $\Sigma = [0,1]$ is the state type (confidence values clamped to the unit interval),
    \item $s_0 = 0$ is the initial state (no validations),
    \item $\text{inc}: \Sigma \times [0,1] \to \Sigma$ is the increment operation: $\text{inc}(s, c) = \max(s, c)$,
    \item $\text{merge}: \Sigma \times \Sigma \to \Sigma$ is the merge operation: $\text{merge}(s_1, s_2) = \max(s_1, s_2)$.
\end{itemize}

\begin{theorem}[Confidence is a State-Based G-Counter CRDT]\label{thm:confidence-gcounter}
The tuple $(\Sigma, \text{inc}, \text{merge}, s_0)$ defined above is a state-based G-Counter CRDT: the merge operation is associative, commutative, and idempotent, and the increment operation is monotonic with respect to the merge.
\end{theorem}
\begin{proof}[Proof Sketch]
\emph{Associativity:} $\text{merge}(\text{merge}(s_1, s_2), s_3) = \max(\max(s_1, s_2), s_3) = \max(s_1, \max(s_2, s_3)) = \text{merge}(s_1, \text{merge}(s_2, s_3))$.
\emph{Commutativity:} $\text{merge}(s_1, s_2) = \max(s_1, s_2) = \max(s_2, s_1) = \text{merge}(s_2, s_1)$.
\emph{Idempotence:} $\text{merge}(s, s) = \max(s, s) = s$.
\emph{Monotonicity of increment:} if $s_1 \leq s_2$ then $\text{inc}(s_1, c) = \max(s_1, c) \leq \max(s_2, c) = \text{inc}(s_2, c)$.
\emph{No negative increments:} confidence values are clamped to $[0,1]$, so the state never decreases.
The full algebraic proof is in Appendix~E (Theorem~\ref{thm:gamma-default-algebraic}).
\end{proof}

The G-Counter semantics guarantee that once a high-confidence validation is recorded, subsequent lower-confidence assertions cannot decrease the aggregate (Theorem~T5). This monotonicity property is critical for audit: validators cannot retrospectively weaken a concept's confidence by appending low-confidence events.

\paragraph{Algebraic structure of all three $\gamma$ variants.} Table~\ref{tab:gamma-algebra} summarises the algebraic properties of the three confidence aggregation operators. $\gamma_{\text{default}}$ is a pure G-Counter (max-semilattice) and satisfies full associativity, commutativity, and idempotence. $\gamma_{\text{agg}}$ and $\gamma_{\text{cal}}$ are commutative over the set of actors (because per-actor maxima are computed over unordered sets) but are not associative with respect to the full chain history, because the bonus term in $\gamma_{\text{agg}}$ and the weighted-mean denominator in $\gamma_{\text{cal}}$ depend on the cumulative actor set, which is not reconstructible from scalar intermediates. Cache invariants: the snapshot caches the last verified prefix; incremental update only re-verifies new events, giving $O(1)$ amortised append time.

\begin{table}[htbp]
\centering
\small
\caption{Algebraic properties of the three confidence aggregation operators.}
\label{tab:gamma-algebra}
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Property} & $\gamma_{\text{default}}$ & $\gamma_{\text{agg}}$ & $\gamma_{\text{cal}}$ \\
\midrule
Bounds & $[0,1]$ (clamp) & $[0,1]$ (min cap) & $[0,1]$ (clamp) \\
Idempotence ($\gamma(V \cup V) = \gamma(V)$) & \checkmark & Per-actor only & Per-actor only \\
Associativity ($\gamma(V_1 \cup V_2) \cup V_3) = \gamma(V_1 \cup (V_2 \cup V_3)$) & \checkmark & $\times$ (bonus term) & $\times$ (denominator) \\
Commutativity ($\gamma(V_1 \cup V_2) = \gamma(V_2 \cup V_1)$) & \checkmark & \checkmark (actor-set) & \checkmark (actor-set) \\
Monotonicity (new event $\not<$ old) & \checkmark & \checkmark (bounded) & \checkmark (bounded) \\
Cache incrementality & $O(1)$ & $O(|V|)$ & $O(|V|)$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Bounds.} All three variants are bounded in $[0,1]$ by construction: $\gamma_{\text{default}}$ uses $\text{clamp}_{[0,1]}$; $\gamma_{\text{agg}}$ uses $\min(1.0, \ldots)$ with $c_{\text{base}} \geq 0.5$; $\gamma_{\text{cal}}$ uses a convex combination clamped to $[0,1]$ (Theorem~T4, proved in Coq).

\textbf{Cache invariants.} The EventChain maintains an incremental confidence cache that is updated on every append: if the new event is a \textsc{Validate} with confidence $c$, the cache becomes $\max(\text{cache}, c)$ for $\gamma_{\text{default}}$; for $\gamma_{\text{agg}}$ and $\gamma_{\text{cal}}$, the cache is recomputed from the full validation set in $O(|V|)$ time. The cache invariant is: $\text{cache} = \gamma(C)$ at all times. If the cache is lost, it can be reconstructed by a single linear scan of $C$, so the cache is a pure optimisation, not a source of truth.

\subsubsection{Fork and Merge Semantics}\label{subsubsec:fork-merge}

A fork $\text{Fork}(C, a)$ appends a $\textsc{Fork}$ event to the parent chain and creates a child chain $C'$ with a fresh $\textsc{Register}$ event. The parent and child evolve independently:
\begin{align}
    \delta(C_{\text{fork}}) &= \max(\delta(C), \text{forked}) \quad \text{(parent status, LUB)} \\
    \delta(C') &= \text{provisional} \quad \text{(child status, fresh genesis)}
\end{align}
where $C_{\text{fork}} = C \oplus [e_{\text{FORK}}]$.

\textbf{Identity preservation under fork.} Let $C$ be a well-formed chain with genesis event $e_0$ and concept identifier $\textit{concept\_id}(C)$. The fork operation $\text{Fork}(C, a)$ produces $(C_{\text{fork}}, C')$ satisfying the following identity conditions:
\begin{enumerate}[label=(I\arabic*)]
    \item \emph{Parent identity preservation:} $\textit{concept\_id}(C_{\text{fork}}) = \textit{concept\_id}(C)$ and $h_{\text{gen}}(C_{\text{fork}}) = h_{\text{gen}}(C)$ (the genesis hash is unchanged).
    \item \emph{Child identity distinctness:} $h_{\text{gen}}(C') \neq h_{\text{gen}}(C)$, because $C'$ has a fresh genesis event with a distinct timestamp and actor.
    \item \emph{Lineage traceability:} There exists a $\textsc{Fork}$ event $e_{\text{FORK}} \in C_{\text{fork}}$ such that $e_{\text{FORK}}.\text{payload}[\text{``child\_concept\_id''}] = \textit{concept\_id}(C')$, establishing a cryptographic link from parent to child.
\end{enumerate}
Condition (I2) ensures that parent and child are distinct concepts in the ontology (they have different identities), while (I3) ensures that the lineage is auditable via the parent's EventChain.

\paragraph{LWW-Set merge formal definition.} Given concurrent well-formed branches $B_1, B_2$ derived from the same genesis event, the merged chain is defined as:
\begin{equation}\label{eq:lww-merge}
    \text{Merge}(B_1, B_2) = \text{Linearize}\bigl(\,\text{Dedup}(B_1 \cup B_2)\,\bigr)
\end{equation}
where $\text{Dedup}(E) = \{ e \in E \mid \nexists e' \in E : e' \neq e \land e'.\text{event\_id} = e.\text{event\_id} \}$ removes duplicate events by \texttt{event\_id}, and $\text{Linearize}(E)$ sorts the deduplicated set by the total order $\prec$ (Eq.~\ref{eq:total-order}). The merge satisfies the LWW-Set (Last-Write-Wins Set) semantics because the deduplication by \texttt{event\_id} ensures that the most recent (or only) version of each event is retained.

\textbf{Identity of the merge result.} Let $B_1, B_2$ be well-formed concurrent branches with shared genesis $e_0$ and common concept identifier $c = \textit{concept\_id}(B_1) = \textit{concept\_id}(B_2)$. The merged chain $M = \text{Merge}(B_1, B_2)$ satisfies:
\begin{enumerate}[label=(M\arabic*)]
    \item \emph{Concept identity preservation:} $\textit{concept\_id}(M) = c$.
    \item \emph{Genesis hash preservation:} $h_{\text{gen}}(M) = h_{\text{gen}}(e_0) = h_{\text{gen}}(B_1) = h_{\text{gen}}(B_2)$.
    \item \emph{Event content preservation:} $\forall e \in B_1 \cup B_2, \exists! e' \in M : e.\text{event\_id} = e'.\text{event\_id} \land e.\text{payload} = e'.\text{payload} \land e.\text{actor} = e'.\text{actor} \land e.\text{timestamp} = e'.\text{timestamp}$.
    \item \emph{Ancestry preservation (optional):} The original branch hashes can be stored as metadata in a $\textsc{MERGE}$ event payload to provide ancestry proofs.
\end{enumerate}
Condition (M3) states that every event from both branches appears exactly once in the merged chain with all immutable content preserved.

\paragraph{Re-anchor operation.} After linearization, the hash chain must be recomputed to restore cryptographic linkage. Let $E^* = [e^*_1, \dots, e^*_n]$ be the linearized merged event sequence. The re-anchor operation $\text{Reanchor}(E^*)$ computes:
\begin{align}
    e^*_1.\text{previous\_event\_id} &\gets \text{null}, \quad e^*_1.\text{prev\_hash} \gets \text{genesis}, \\
    \forall i > 1: \quad e^*_i.\text{previous\_event\_id} &\gets e^*_{i-1}.\text{event\_id}, \quad e^*_i.\text{prev\_hash} \gets \text{SHA-256}(\text{Canon}(e^*_{i-1})).
\end{align}
Each event's immutable content (\texttt{event\_id}, payload, actor, timestamp) is preserved; only the linkage pointers are updated. This means the merged chain has a \emph{different} hash sequence from either parent branch, but all original event content is preserved. The hash chain therefore provides \emph{post-merge integrity}, not \emph{pre-merge ancestry preservation}. For applications requiring ancestry proofs, the original branch hashes can be stored as metadata in a \textsc{MERGE} event payload.

\textbf{Total order $\prec$.} For events $e_i, e_j$:
\begin{equation}\label{eq:total-order}
    e_i \prec e_j \iff e_i.\text{timestamp} < e_j.\text{timestamp} \;\lor\; (e_i.\text{timestamp} = e_j.\text{timestamp} \;\land\; e_i.\text{event\_id} <_{\text{lex}} e_j.\text{event\_id})
\end{equation}
where $<_{\text{lex}}$ is lexicographic comparison of UUID strings. The order $\prec$ satisfies four formal properties:
\begin{enumerate}[label=(\roman*)]
    \item \emph{Reflexivity:} $e \prec e$ is false (strict order), but non-strict $\preceq$ is reflexive: $e \preceq e$ holds by definition.
    \item \emph{Antisymmetry:} if $e_i \prec e_j$ and $e_j \prec e_i$, then $e_i = e_j$. This holds because timestamps are equal and UUID strings are totally ordered, so $<_{\text{lex}}$ and $>_{\text{lex}}$ cannot both hold.
    \item \emph{Transitivity:} if $e_i \prec e_j$ and $e_j \prec e_k$, then $e_i \prec e_k$. This follows from transitivity of $<$ on timestamps and transitivity of $<_{\text{lex}}$ on UUID strings.
    \item \emph{Totality:} for any distinct $e_i, e_j$, either $e_i \prec e_j$ or $e_j \prec e_i$. This holds because timestamps are comparable (real numbers), and if timestamps are equal, $<_{\text{lex}}$ is a total order on UUID strings.
\end{enumerate}
Within a single chain, events are totally ordered by cryptographic linkage. Across chains, events are initially partially ordered; the merge operation linearizes the merged set by sorting on $\prec$.

\textbf{Concurrent conflict resolution.} For $\delta$: under CRDT LUB semantics, the status is the maximum over all lifecycle events in the chain, independent of append order. A \textsc{Deprecate} event always dominates a \textsc{Validate} event (since $\text{deprecated} > \text{validated}$ in the lattice), and an \textsc{Archive} event dominates all other lifecycle events. For $\gamma$: the G-Counter ($\max$) over all \textsc{Validate} events determines the confidence, independent of append order. Equivocation (the same actor appending contradictory events) is retained as auditable evidence, not resolved by the ordering.

\subsubsection{Concurrent Event Interaction}\label{subsubsec:concurrent-events}

Under CRDT semantics, concurrent lifecycle events commute because the join operator is associative and commutative. The following examples illustrate four representative scenarios.

\paragraph{Example 1: Concurrent VALIDATE events.} Agent A appends \textsc{Validate} with confidence $0.8$; agent B appends \textsc{Validate} with confidence $0.9$. After merge, $\gamma = \max(0.8, 0.9) = 0.9$ and $\delta = \text{validated}$ (the LUB of two \textsc{Validate} events).

\paragraph{Example 2: Concurrent VALIDATE and DEPRECATE.} Agent A appends \textsc{Validate} ($c=0.8$); agent B appends \textsc{Deprecate}. After merge, $\gamma = 0.8$ (the G-Counter retains the validation confidence) and $\delta = \text{deprecated}$ (because $\text{validated} \prec \text{deprecated}$ in the lattice, so $\text{validated} \sqcup \text{deprecated} = \text{deprecated}$).

\paragraph{Example 3: Concurrent FORK and VALIDATE.} Agent A forks a child concept $C'$ from parent $C$; agent B validates the parent $C$. The parent status is $\max(\text{validated}, \text{forked}) = \text{validated}$ (if already validated) or $\max(\text{provisional}, \text{forked}) = \text{forked}$ (if not yet validated). The child $C'$ starts at \textsc{provisional} with a fresh genesis event.

\paragraph{Example 4: Concurrent RELATE and REVOKE.} Agent A appends \textsc{Relate}$(C, X)$; agent B appends \textsc{Revoke}$(C, X)$. Both events are preserved in the merged chain. The L3 query engine applies \emph{HoldsAt} semantics: the relation is active if the most recent event (by total order $\prec$, Eq.~\ref{eq:total-order}) is \textsc{Relate}, and inactive if the most recent is \textsc{Revoke}. If the timestamps are unordered, both events are returned and the consumer resolves the conflict using the total order $\prec$.

The join-semilattice properties of the status space and the G-Counter properties of confidence are machine-verified in Coq (\texttt{CRDT.v}, Theorem~T9) and model-checked in the TLA\textsuperscript{+} specification (\path{specs/CRDTMerge.tla}) for chains up to 20 events.

\subsubsection{Theorems: Summary of Proven Properties}\label{subsubsec:theorems-summary}

\begin{table}[htbp]
\centering
\small
\caption{Summary of nine core theorems and Lemma~E2. All proofs are in Appendix~E; proof strategies are indicated.}
\label{tab:theorems-summary}
\begin{tabularx}{\textwidth}{@{}p{1.2cm}p{3.8cm}p{2.2cm}p{4.5cm}@{}}
\toprule
\textbf{Thm} & \textbf{Statement} & \textbf{Strategy} & \textbf{Complexity / Note} \\
\midrule
T1 & $\delta(C)$ is unique (LUB over finite lattice) & Direct & $O(|C|)$ scan, $O(1)$ cached \\
E2 & Incremental LUB correctness: $\delta(C \oplus [e]) = \max(\delta(C), f(e.\tau))$ & Induction & Base: empty chain; Step: LUB monotonicity \\
T2 & Fork: $\delta(C') = \text{provisional}$, $\delta(C_{\text{fork}}) = \max(s, \text{forked})$ & Direct & By LUB definition \\
T3 & Transition monotonicity: valid iff $e_{\text{new}}.\tau \in \Sigma_{\text{life}}$ & Induction & Base: genesis; Step: append \\
T4 & $\gamma_{\text{default}}, \gamma_{\text{agg}}, \gamma_{\text{cal}} \in [0,1]$ & Case analysis & Clamp + convex combination \\
T5 & $\gamma_{\text{default}}(C \oplus [e]) \geq \gamma_{\text{default}}(C)$ for \textsc{Validate} $e$ & Direct & G-Counter $\max$ semantics \\
T6 & If $\delta(C) = \text{validated}$ then $\gamma(C) \geq 0.5$ & Direct & Precondition requires $\geq 0.5$ \\
T7 & Appending valid event preserves $\text{WF}(C)$ & Induction & Axiom-by-axiom check \\
T8 & Precondition eval is $O(k)$ for $k$ rules & Direct & $O(1)$ per rule, short-circuit AND \\
T9 & CRDT merge: commutative, associative, idempotent & Direct & Set union + LUB/G-Counter \\
T10 & Confidence is a state-based G-Counter CRDT & Direct & State type + inc/merge ops \\
T11 & $\delta$ is independent of $\prec$ for lifecycle events & Direct & Max over finite set \\
\bottomrule
\end{tabularx}
\end{table}

\textbf{Proof status and mechanization scope.} All eleven theorems (T1--T11) and Lemma~E2 are machine-verified in Coq 8.18.0. T8 (precondition complexity) is a Python-level implementation property and is not formalized in Coq. The Coq development comprises approximately 2,500 lines across seven modules (Status, Event, Confidence, Chain, Invariants, CRDT, and Crypto). All algebraic proofs are closed with zero remaining \texttt{Admitted} lemmas. Three previously stubbed axioms (scope ACL, signature verification, and confidence clamping) have been replaced with their full definitions, introducing abstract Ed25519/SHA-256 cryptographic primitives in \texttt{Crypto.v}. A randomized trace checker (E25, 10,000 traces, length 2--100, 100\% pass rate) provides independent property-based validation.

\begin{table}[htbp]
\centering
\footnotesize
\caption{Proof status of the nine core theorems and Lemma~E2, with mechanization modality and verification evidence. \textbf{Coq} = structural inductive proof (unbounded); \textbf{Python} = empirical test-suite validation (bounded); \textbf{TLA+} = model-checking confirmation (bounded).}
\label{tab:proof-status}
\begin{tabularx}{\textwidth}{@{}p{1.1cm}p{1.3cm}p{1.6cm}p{2.6cm}p{1.1cm}p{1.6cm}p{1.5cm}X@{}}
\toprule
\textbf{Thm} & \textbf{Paper \S} & \textbf{Coq File} & \textbf{Coq Theorem Name} & \textbf{Status} & \textbf{Python Tests} & \textbf{TLA+} & \textbf{Notes} \\
\midrule
T1 & \S\ref{subsec:compact-formal-spec} & \texttt{Chain.v} & \texttt{well\_formed} (defn) & \checkmark & \texttt{test\_theorem\_t1.py} & --- & 12 axioms; T7/T9 prove preservation \\
E2 & \S\ref{subsec:compact-formal-spec} & \texttt{Invariants.v} & \texttt{E2\_inductive\_status\_derivation} & \checkmark & \texttt{test\_theorem\_t2.py} & \checkmark & Equational reasoning; TLC to len 20 \\
T2 & \S\ref{subsec:compact-formal-spec} & \texttt{Invariants.v} & \texttt{fork\_determinism} (6 conj.) & \checkmark & \texttt{test\_theorems.py} & --- & Parent/child status by LUB \\
T3 & \S\ref{subsec:compact-formal-spec} & \texttt{Status.v} & \texttt{status\_monotonic} & \checkmark & \texttt{test\_theorems.py} & \checkmark & No backward transitions \\
T4 & \S\ref{subsec:compact-formal-spec} & \texttt{Confidence.v}, \texttt{Invariants.v} & \texttt{confidence\_boundedness} + \texttt{gamma\_agg\_bounded} & \checkmark & \texttt{test\_theorems.py} & --- & Clamp + convex combo \\
T5 & \S\ref{subsec:compact-formal-spec} & \texttt{Confidence.v} & \texttt{confidence\_monotonicity\_default} & \checkmark & \texttt{test\_theorems.py} & --- & G-Counter $\max$ semantics \\
T6 & \S\ref{subsec:compact-formal-spec} & \texttt{Invariants.v} & \texttt{status\_confidence\_consistency} & \checkmark & \texttt{test\_theorems.py} & --- & Precondition $\geq 0.5$ \\
T7 & \S\ref{subsec:compact-formal-spec} & \texttt{Chain.v}, \texttt{Invariants.v} & \texttt{well\_formedness\_preservation} & \checkmark & \texttt{test\_theorems.py} & --- & 490-line proof (replaced 6 stubs) \\
T8 & \S\ref{subsec:precondition-executor} & --- & --- & N/A & \texttt{test\_theorems.py} & --- & Python impl.~property; not in Coq \\
T9 & \S\ref{subsec:compact-formal-spec} & \texttt{CRDT.v} & Commutative, associative, idempotent, well-formed & \checkmark & \texttt{test\_theorems.py} & \checkmark & LWW-Set + LUB/G-Counter \\
T10 & \S\ref{subsubsec:confidence-gcounter} & \texttt{Confidence.v} & \texttt{confidence\_gcounter\_crdt} & \checkmark & \texttt{test\_theorems.py} & --- & State-based G-Counter \\
T11 & \S\ref{subsec:distributed-ordering} & \texttt{Status.v} & \texttt{delta\_independent\_of\_ordering} & \checkmark & \texttt{test\_theorems.py} & \checkmark & Max over finite set \\
\bottomrule
\end{tabularx}
\end{table}

\textbf{Three verification modalities.} We distinguish three complementary verification strategies with different completeness guarantees:
\begin{enumerate}[label=(\roman*)]
    \item \emph{Coq (unbounded structural induction):} Proofs by induction on the event sequence, valid for chains of arbitrary finite length. T1, T2, T3, T4, T5, T7, T9, T10, and T11 are proved for all chains over the full event-type alphabet. These are the primary formal guarantees.
    \item \emph{Python test suite (bounded empirical):} 944 pytest cases including adversarial trials, boundary conditions, and randomized trace checking (E25: 10,000 traces, length 2--100). These validate the implementation against the specification for finite but large instances.
    \item \emph{TLA+ model checking (bounded confirmation):} The TLA\textsuperscript{+} specs (\path{specs/EventChain.tla}, \path{specs/CRDTMerge.tla}, \path{specs/ConsensusEngine.tla}) are model-checked for chains up to length 20, two-branch CRDT merges, and $N_{\min} \leq 5$ validators. They serve as \emph{confirmation} of the inductive Coq theorems for small instances, not as standalone proofs.
\end{enumerate}
This dual-strategy (Coq for unbounded correctness, Python/TLA+ for bounded confirmation) is standard in formal methods \cite{lamport2002specifying,coq_reference}.

If agent\_2 disagrees at Step~5, they may append a \textsc{Fork} event: the parent becomes \textit{forked} and the child begins at \textit{provisional}, both independently well-formed.

\subsection{Distributed Event Ordering and Concurrency Model}\label{subsec:distributed-ordering}

\subsubsection{Clock Skew and Timestamp Assumptions}

ADL Lite uses ISO~8601 UTC timestamps with no global clock assumption. Each event carries the author's local timestamp, and monotonicity is enforced by the \texttt{append()} method. Formally, let $e_{\text{new}}$ be the event to append and $e_{\text{last}}$ the current last event in chain $C$. The timestamp adjustment rule is:
\begin{equation}\label{eq:timestamp-adjust}
    e_{\text{new}}.\text{timestamp} \gets \max\bigl(e_{\text{new}}.\text{timestamp},\; e_{\text{last}}.\text{timestamp}\bigr)
\end{equation}
This ensures that within a single chain, timestamps are non-decreasing: $\forall i > 1 : e_i.\text{timestamp} \geq e_{i-1}.\text{timestamp}$. Clock skew between different authors is handled by the total order $\prec$ (Equation~\ref{eq:total-order}): if two events have the same timestamp, the lexicographic \texttt{event\_id} provides a deterministic tiebreaker.

\textbf{No global clock assumption.} ADL Lite does not require synchronized clocks across agents. Event timestamps are local UTC values generated by the appending agent's system clock. Skew of up to several seconds is tolerated because the status derivation $\delta(C)$ and confidence aggregation $\gamma(C)$ are computed over the \emph{set} of lifecycle events, not over their timestamps. Precondition evaluation uses the \emph{snapshot-derived state} (L1 front matter), not timestamps, so precondition correctness is independent of clock accuracy. Timestamp ordering is used only for linearization during CRDT merge and for human-readable audit trails.

\textbf{Deterministic CRDT merge ordering.} For CRDT merge across independent agents, events are canonically sorted by \texttt{event\_id} (UUID) rather than by timestamp, to ensure deterministic ordering even when clocks are skewed. The total order $\prec$ (Equation~\ref{eq:total-order}) uses timestamp as the primary key and \texttt{event\_id} as the tiebreaker for presentation and human audit; the merge algorithm uses the same $\prec$ definition, which is guaranteed to be total because UUID strings are totally ordered. The timestamp is therefore for \emph{audit/replay} (human-readable chronology) and \emph{not for consensus}: the consensus state ($\delta$, $\gamma$) is derived from the event set, independent of the ordering used for linearization.

\begin{theorem}[$\delta$ is Independent of $\prec$ for Lifecycle Events]\label{thm:delta-independence}
For any well-formed chain $C$ and any permutation $\pi$ of the events in $C_{\text{life}}$ that preserves the total order $\prec$ within each branch, $\delta(C) = \delta(C_{\pi})$, where $C_{\pi}$ is the chain with lifecycle events reordered according to $\pi$.
\end{theorem}
\begin{proof}
By definition, $\delta(C) = \max_{\prec}\{ f(e.\tau) \mid e \in C_{\text{life}} \}$ (Eq.~\ref{eq:delta-def}). The max operator over a finite set is independent of the order in which elements are enumerated. Therefore, any permutation of $C_{\text{life}}$ that preserves the set of lifecycle events yields the same maximum. The total order $\prec$ is used only for linearization during merge; it does not affect the set of lifecycle events, and therefore does not affect $\delta$.
\end{proof}

This independence is a direct consequence of the CRDT lattice semantics: the status is the LUB over all lifecycle events, which is a commutative and associative aggregation, not a sequential state machine. Even if two events are reordered due to clock skew, the final status is unchanged.

\subsubsection{Concurrency Model: Single-Writer vs.~Optimistic Concurrency}

Within a single EventChain, ADL Lite enforces a single-writer model: only one agent may append to a given chain at a time, serialised by \texttt{threading.Lock} in the Python runtime. For multi-agent collaboration, repository-level locking is provided by Git (with signed commits as tamper evidence). Optimistic concurrency is supported for multi-agent scenarios: agents may append to their local clones independently, and conflicts are resolved by CRDT merge (LWW-Set union with \texttt{event\_id} deduplication) rather than by blocking. This design avoids the need for a distributed consensus protocol at append time, while ensuring that merged chains converge deterministically (Theorem~9).

\subsubsection{CRDT Merge and Precondition Interaction}

When branches are merged, the merged event set is linearised by the total order $\prec$, and the hash chain is recomputed. Crucially, merged events do not re-evaluate preconditions retroactively: preconditions are checked at append time on the local chain, and the merge operation only reconciles event sets. If a merged event would have violated a precondition in the context of the merged chain (e.g., a \textsc{Validate} appended to a chain that already contains a \textsc{Deprecate}), it is still retained as auditable evidence, but the derived status $\delta(C)$ reflects the LUB of all lifecycle events (so \textsc{Deprecate} dominates). This separation of \emph{append-time validation} (local preconditions) from \emph{merge-time reconciliation} (set union + linearization) ensures that no events are silently dropped, while preserving the monotonicity guarantees of the CRDT lattice.

\subsection{Comparison with Formal Event Frameworks}\label{subsec:event-frameworks}

We compare ADL Lite's event model to Event Calculus, Situation Calculus, and Description Logic. In Event Calculus terms, $\delta(C)$ computes the fluent $\text{Status}(c, t)$ by scanning the ordered event sequence, equivalent to EC's interval-based reasoning but computationally simpler ($O(n)$ vs.~$O(n^2)$). ADL Lite avoids the frame problem by construction: there are no stored mutable fields, so all properties are recomputed from the event sequence. The derivation functions $\delta$ and $\gamma$ are computable in linear time---$\mathcal{O}(n)$ for $\delta$ (single scan of lifecycle events) and $\mathcal{O}(|V|) \leq \mathcal{O}(n)$ for $\gamma$ (single scan of validation events). The implementation optionally memoizes derived snapshots in a warm index for $O(1)$ incremental evaluation, but the canonical state is always recomputable from the event sequence in $O(n)$ time. This corresponds to DL-Lite$_{\mathcal{R}}$ path queries and does not require full OWL-DL expressivity (NExpTime-complete). A full discussion is in Appendix~H.

\subsection{L3 Relation Management Across Forks}\label{subsec:l3-fork-governance}

L3 relations (\textsc{Relate}, \textsc{Revoke}, \textsc{Evidence}) are events in a specific chain, asserting that a relation exists (or is revoked) for the concept represented by that chain. Relations are \emph{per-chain}, not global: a \textsc{Relate} event in chain~A does not affect chain~B, even if chain~B is a fork of chain~A.

\textbf{Fork behaviour for relations.} When a chain is forked, the child chain starts with a fresh \textsc{Register} event and does \emph{not} inherit the parent's L3 relations. This is consistent with the fork semantics: the child is a new concept with a new genesis hash, and its relations must be established independently. The parent chain may contain a \textsc{Fork} event with a \texttt{target\_concept\_id} pointing to the child, which establishes a \texttt{fork-of} relation from parent to child. This relation is recorded in the parent chain, not the child chain.

\textbf{Conflicting \textsc{Revoke}/\textsc{Relate} across forks.} If chain~A has a \textsc{Relate} event to concept~X, and chain~B (a fork of~A) has a \textsc{Revoke} event to concept~X, these are not conflicting because they are in different chains. The relation to~X exists for concept~A but not for concept~B. A downstream consumer querying for concepts related to~X will find~A but not~B. If the consumer wants to track the lineage, they can query the \texttt{fork-of} relation in the parent chain.

\textbf{Merge policy for relations (Phase~3).} When two branches of the same chain are merged, the L3 relations are merged by set union (LWW-Set semantics). If both branches contain a \textsc{Relate} event to the same target with the same predicate, the merged chain contains one \textsc{Relate} event (deduplicated by \texttt{event\_id}). If one branch contains a \textsc{Relate} and the other contains a \textsc{Revoke} to the same target, both events are preserved in the merged chain. The \texttt{HoldsAt} semantics determines whether the relation is currently active: a \textsc{Relate} event followed by a \textsc{Revoke} event means the relation is no longer active, regardless of which branch produced the \textsc{Revoke}.

Relations are domain-level assertions that should be re-evaluated for each fork, not inherited mechanically. A child concept may have a different relation profile from its parent because it represents a different (possibly refined) capability.

\subsection{Trust Model and Security Boundaries}\label{subsec:trust-model}

The cryptographic hash chain ensures content integrity: any modification breaks the chain linkage and is detected by $\text{VerifyIntegrity}(C)$. The scope of the current collaborative-audit model is deliberately limited to \emph{collaborative non-Byzantine environments}---trusted organisations, open-source communities with social verification, and research groups where actors are known by reputation rather than cryptographic identity. Hash chains detect tampering but do not prevent it, and they provide no actor authentication. We explicitly state these boundaries to prevent over-claiming: ADL Lite is a tamper-\emph{evident} audit layer, not a tamper-\emph{proof} security protocol.

\subsubsection{Trust Model Evolution: Non-Byzantine to Authenticated}\label{subsubsec:trust-evolution}

The trust model of ADL Lite is designed to evolve incrementally from a collaborative non-Byzantine setting toward authenticated, economically secured validation. Table~\ref{tab:trust-evolution} summarises the four phases.

\begin{table}[htbp]
\centering
\caption{Trust model evolution: from collaborative non-Byzantine to authenticated validation.}
\label{tab:trust-evolution}
\begin{tabular}{@{}lp{2.8cm}p{2.5cm}p{3cm}p{2.2cm}@{}}
\toprule
\textbf{Phase} & \textbf{Authentication} & \textbf{Sybil Resistance} & \textbf{Collusion Mitigation} & \textbf{Deployment Cost} \\
\midrule
Phase 1 (current) & Self-declared string IDs & None & $N_{\min} \geq 1$ (configurable) & Zero \\
Phase 1.5 (implemented) & \texttt{did:key} Ed25519 signatures & DID uniqueness & $N_{\min} \geq 2$ + accuracy-weighted $\gamma$ & Low (pip install) \\
Phase 2 (planned) & Full key registry + \texttt{did:web}/\texttt{did:ethr} & DID-based reputation & Staking + slashing & Moderate (web of trust) \\
Phase 3 (future) & BFT transport + Merkle logs & Economic stakes & Game-theoretic guarantees & High (infrastructure) \\
\bottomrule
\end{tabular}
\end{table}

The current system (Phase~1) is designed for collaborative non-Byzantine environments: trusted organisations, open-source communities with social verification, and research groups where actors are known by reputation rather than cryptographic identity. In these settings, the cost of Sybil creation is social rather than computational: a pseudonymous validator must earn trust through repeated participation. The phased approach allows incremental adoption: organisations can deploy ADL Lite with zero authentication overhead today and strengthen guarantees as requirements mature.

\paragraph{Minimal Authentication Recommendation.} For production use, we recommend a minimum of $N_{\min} \geq 2$ distinct validators for any \textsc{Validate} event that triggers a status transition to \texttt{validated}. Validator identity should be anchored via \texttt{did:key} (Ed25519), which requires no external dependencies---a validator generates a key pair locally and publishes only the public key. Public keys are stored in the local key registry (\texttt{key\_registry.py}), which is itself an append-only EventChain (YAML file with append-only semantics). The ActionExecutor verifies Ed25519 signatures on \textsc{Validate} and \textsc{Fork} events against the key registry. This does not prevent Sybil attacks (a single actor can still generate multiple key pairs), but it raises the cost of identity creation from ``choose a string'' to ``generate a key pair per validator'', and it enables audit trails that link events to cryptographic identities. The DID resolver (\texttt{did\_resolver.py}) supports \texttt{did:key}, \texttt{did:web}, and \texttt{did:ethr} resolution; the current implementation defaults to \texttt{did:key} for zero-dependency deployment.

\subsubsection{Trust Assumptions}

We assume (1)~trusted storage (Git or filesystem preserving history integrity, with signed commits providing tamper evidence), (2)~trusted authoring environment (events injected by compromised agents are recorded as tamper-evident evidence), and (3)~clock skew tolerance (ordering uses cryptographic linkage, not timestamps). Actor identity in the current implementation is a self-declared string; impersonation and replay are discussed below. Future authentication details are in Appendix~K.

\subsubsection{Current Trust Boundary: What Is and Is Not Detectable}

The hash chain detects \emph{content} tampering (modification of existing events) because any change breaks the hash linkage. However, the current model cannot detect the following \emph{structural} and \emph{identity} attacks: (i)~\emph{Actor impersonation:} any agent may declare an arbitrary string identifier; there is no cryptographic binding to a real-world identity. (ii)~\emph{Replay:} an event (or an entire chain) can be copied and re-submitted under a different concept identifier. (iii)~\emph{Sybil creation:} a single agent can create multiple pseudonymous identities and self-validate using each. (iv)~\emph{Selective omission:} an agent may withhold evidence that would contradict their claim; the chain provides no mechanism to prove that a missing event \emph{should} have been present. (v)~\emph{Historical rewrite:} an actor with repository-level access (e.g., Git push access) can force-push a rebased history, replacing the chain with a new genesis. These attacks are detectable by external audit (comparing a local clone against a signed reference), but they are not \emph{prevented} by the chain itself. This characterisation explains why we describe the current model as a \emph{collaborative non-Byzantine} setting.

\subsubsection{Collusion Vulnerability Analysis}\label{subsubsec:collusion-analysis}

The confidence aggregation function $\gamma(C)$ is vulnerable to collusion in the current collaborative-audit model because it lacks authentication and incentive compatibility.

\begin{lemma}[Collusion Vulnerability]\label{lem:collusion-vulnerability}
In the current collaborative-audit model, $\gamma_{\text{default}}(C)$ returns the maximum confidence across all \texttt{VALIDATE} events (G-Counter semantics). A single actor can drive $\gamma(C)$ to any value in $[0,1]$ by appending a \texttt{VALIDATE} event with the desired confidence, achieving that confidence without independent validation. However, the actor cannot \emph{lower} the confidence after a higher validation has been recorded, as G-Counter monotonicity (Theorem~5) prevents downgrade.
\end{lemma}
\begin{proof}
By definition, $\gamma_{\text{default}}(C) = \max\{e.\text{payload}[\text{``confidence''}] \mid e \in C \land e.\tau = \text{VALIDATE}\}$ (Equation~\ref{eq:gamma-default}). A colluding actor appending an event with $\text{confidence} = 1.0$ yields $\gamma(C) = 1.0$. Since $\max$ is monotonic, subsequent events with lower confidence do not change the aggregate.
\end{proof}

\begin{lemma}[Minimum Validator Threshold Mitigation]\label{lem:threshold-mitigation}
Let $N_{\min} \geq 2$ be a minimum validator threshold enforced as a precondition for \texttt{VALIDATE}. Then a single colluding actor cannot trigger the $\text{validated}$ status; the collusion cost rises to $k \geq N_{\min}$.
\end{lemma}
\begin{proof}
The precondition $\langle N_{\text{vals}}, \text{GTE}, N_{\min} \rangle$ is evaluated before appending; for $N_{\text{vals}} = 1 < N_{\min}$, the event is rejected.
\end{proof}

\textbf{Implications.} Lemma~\ref{lem:collusion-vulnerability} shows that the current model's confidence aggregation allows a single actor to \emph{set} confidence to any desired value, but not to \emph{reduce} it after a higher validation has been recorded (G-Counter monotonicity). Lemma~\ref{lem:threshold-mitigation} shows that a minimum-validator threshold prevents a single actor from triggering the \texttt{validated} status. Full collusion resistance requires authenticated identities and a staking-based mechanism (FW9).

\subsubsection{Known Limitations and Planned Mitigations}

The current model has four limitations: (1) no actor identity verification for legacy string identifiers (we have implemented a minimal \texttt{did:key} layer; full W3C LD-Proofs and DIDs remain future work); (2) no Byzantine resilience; (3) no equivocation prevention (contradictory events are recorded as auditable evidence); (4) genesis immutability relies solely on hash-chain detection and signed Git commits.

\textbf{Three-phase authentication roadmap.} \textbf{Phase 1 (current):} collaborative-audit model with self-declared string identifiers and SHA-256 hash-chain integrity. \textbf{Phase 1.5 (implemented):} minimal \texttt{did:key} resolution without external dependencies; Ed25519 signatures in the \texttt{signature} field; \texttt{verify\_signature()} validates both DID-derived and registry-derived keys. \textbf{Phase 2 (planned):} full Ed25519 key registry with \texttt{KEY\_ROTATE} and \texttt{KEY\_REVOKE} events; DID document storage with \texttt{did:web} and \texttt{did:ethr} support; W3C Linked Data Proofs. \textbf{Phase 3 (future):} BFT transport layer; CRDT-style merge with semantic policies; Ontological Assertion Market (staking/slashing for incentive-compatible validation). Authenticated identities would change the threat model: Sybil attacks become prohibitively expensive, and collusion cost rises to economic stakes (FW9).

\subsubsection{Threat Model Summary}

Table~\ref{tab:threat-model} summarises the progression from the current release to future releases. The full discussion, including per-threat attack trees, signed-commit anchoring details, and recovery procedures, is in Appendix~K.

\begin{table}[htbp]
\centering
\caption{Threat model progression across phases.}
\label{tab:threat-model}
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{Threat} & \textbf{Current model} & \textbf{Signed commits} & \textbf{Phase 1.5} & \textbf{Future auth.} \\
\midrule
Content tampering & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
Event reordering & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
Event deletion & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
Actor impersonation & $\times$ & $\triangle$ & $\triangle$ & $\checkmark$ \\
Equivocation & $\circ$ & $\circ$ & $\circ$ & $\triangle$ \\
Clock skew & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
Genesis replacement & $\circ$ & $\triangle$ & $\triangle$ & $\checkmark$ \\
Byzantine majority & $\times$ & $\times$ & $\times$ & $\triangle$ \\
Replay attack & $\circ$ & $\triangle$ & $\triangle$ & $\checkmark$ \\
Timestamp manipulation & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
Collusion attack & $\times$ & $\times$ & $\triangle$ & $\triangle$ \\
Social engineering & $\times$ & $\times$ & $\times$ & $\times$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Phase 1.5} provides Ed25519 \texttt{did:key} resolution and signature verification (\S\ref{subsec:trust-model}), which mitigates impersonation, replay, and genesis replacement relative to string identifiers, but does not prevent Sybil creation (a single actor can still generate multiple \texttt{did:key}s) or collusion (no economic cost).

The CRDT merge semantics are now \textbf{implemented in the alpha codebase}. As of v0.3.5, \texttt{EventChain.status} and \texttt{EventChain.confidence} derive via CRDT lattice operations rather than the earlier last-event rule. The status lattice is a join-semilattice with partial order $\text{provisional} < \text{forked} < \text{validated} < \text{deprecated} < \text{archived}$; the status of a chain is the LUB of all lifecycle events, computed as $\max$ over the integer order. The confidence is a G-Counter ($\max$) over all \texttt{VALIDATE} events, so once a high-confidence validation is recorded, subsequent lower-confidence assertions cannot decrease the aggregate. Theorem~9 is now a \emph{verified property} of the current system, not a future capability.

The CRDT merge operates at the \emph{set} level (events as elements), while the hash chain operates at the \emph{sequence} level (events as ordered nodes). When two branches $B_1$ and $B_2$ are merged, the resulting event set $B_1 \cup B_2$ is re-linearised by sorting on \texttt{timestamp} (with \texttt{event\_id} tie-breaking). The hash chain is then recomputed over this merged sequence: each event's \texttt{previous\_event\_id} and \texttt{prev\_hash} are updated to point to its predecessor in the new ordering. This means the merged chain has a \emph{different} hash sequence from either parent branch, but all original event content is preserved (via its immutable \texttt{event\_id} and payload). The hash chain therefore provides \emph{post-merge integrity}, not \emph{pre-merge ancestry preservation}. For applications requiring ancestry proofs, the original branch hashes can be stored as metadata in a \textsc{MERGE} event payload.

\textbf{Theorem~9 (CRDT Convergence).} For any well-formed concurrent branches $B_1, B_2$ derived from the same genesis event, the LWW-Set merge satisfies commutativity, associativity, and idempotence, and $\delta(\text{Merge}(B_1, B_2))$ is uniquely determined.

\emph{Proof.} Let $B_1, B_2$ be well-formed chains sharing the same genesis event $e_0$. Let $\text{Merge}$ be defined by Eq.~\ref{eq:lww-merge} and $\text{Reanchor}$ by the re-anchor operation below it.

\emph{(i) Commutativity.} The deduplicated union is symmetric: $\text{Dedup}(B_1 \cup B_2) = \text{Dedup}(B_2 \cup B_1)$ because set union is commutative and the deduplication predicate depends only on the unordered set of \texttt{event\_id} values. Linearization by $\prec$ is also symmetric because $\prec$ is a total order (Theorem~\ref{thm:delta-independence}). Therefore $\text{Merge}(B_1, B_2) = \text{Merge}(B_2, B_1)$.

\emph{(ii) Associativity.} For three branches $B_1, B_2, B_3$:
\begin{align*}
    \text{Merge}(B_1, \text{Merge}(B_2, B_3))
    &= \text{Linearize}(\text{Dedup}(B_1 \cup \text{Linearize}(\text{Dedup}(B_2 \cup B_3)))) \\
    &= \text{Linearize}(\text{Dedup}(B_1 \cup B_2 \cup B_3)) \\
    &= \text{Linearize}(\text{Dedup}(\text{Linearize}(\text{Dedup}(B_1 \cup B_2)) \cup B_3)) \\
    &= \text{Merge}(\text{Merge}(B_1, B_2), B_3)
\end{align*}
The key step is that $\text{Dedup}$ and $\text{Linearize}$ are idempotent and commute with set union when applied to the same ordering $\prec$.

\emph{(iii) Idempotence.} $\text{Merge}(B_1, B_1) = \text{Linearize}(\text{Dedup}(B_1 \cup B_1)) = \text{Linearize}(B_1) = B_1$ because $\text{Dedup}(B_1) = B_1$ (well-formedness guarantees distinct \texttt{event\_id} values within a branch) and $\text{Linearize}$ preserves the order of an already-sorted chain.

\emph{(iv) Determinism of $\delta$.} The merged set contains all lifecycle events from both branches. By Theorem~\ref{thm:delta-independence}, $\delta$ is the LUB over the status lattice of all lifecycle events in the merged set, independent of their ordering under $\prec$. Therefore $\delta(\text{Merge}(B_1, B_2))$ is uniquely determined and independent of merge order.

\emph{(v) Re-anchor correctness.} After linearization, the $\text{Reanchor}$ operation recomputes cryptographic linkage by chaining each event to its predecessor in the new ordering. Because the canonicalization function $\text{Canon}$ is deterministic and SHA-256 is collision-resistant, the recomputed hash chain is unique for the given linearization. No event content is modified, so the post-merge chain preserves all auditable data from both parents.

We evaluated Theorem~9 through executable assertions (1,311 tests, all passing after the v0.3.5 CRDT migration) and stress tests (E6: 201 chains, 9,300 events; E13: 50,000 events; E23: 20 concurrent agents) with zero integrity failures. Machine-checked proofs for unbounded chains are provided by the Coq development (\texttt{CRDT.v}, Theorem~T9).

\textbf{Migration from last-event to CRDT: completed.} The earlier last-event fork resolution was a special case of CRDT merge in which the branch set size is 1 (no concurrent branches). We completed the migration to full CRDT compliance in three phases. \emph{Phase~1 (completed):} last-event with total order $\prec$; status and confidence derived from the most recent event. \emph{Phase~2 (completed):} CRDT lattice semantics for status (LUB) and confidence (G-Counter $\max$); all events commute because they are append-only and immutable. \emph{Phase~3 (future work):} semantic merge policies for multi-way concurrent merges with conflicting lifecycle events. Under Phase~2, \texttt{REGISTER} events commute because they create independent genesis events. \texttt{RELATE}, \texttt{EVIDENCE}, and \texttt{SEAL} events commute because they are communication events that do not affect $\delta$. \texttt{VALIDATE} events from distinct actors commute because each contributes to the per-actor maxima in $\gamma_{\text{agg}}$ and $\gamma_{\text{cal}}$. Conflicting lifecycle events require conflict-resolution policies for Phase~3. Examples include \texttt{VALIDATE} versus \texttt{DEPRECATE}, or multiple \texttt{VALIDATE} events from the same actor (equivocation). The planned policies are: (a)~quorum-based resolution ($\geq N_{\min}$ validators for \texttt{VALIDATE} to override \texttt{DEPRECATE}); (b)~accuracy-weighted ordering; (c)~stake-weighted resolution (planned for FW9). Blocklace~\cite{blocklace2024} provides the BFT transport layer for Phase~3: its cryptographic hash DAG enables equivocation detection and Byzantine exclusion, while ADL Lite provides the application semantics (lifecycle derivation, precondition enforcement). The integration architecture comprises three layers: Blocklace DAG as the transport layer, ADL Lite EventChain as the application layer, and a shim layer that maps Blocklace events to ADL Lite events and resolves conflicts using the semantic policies above.

\subsection{Implementation Infrastructure}\label{subsec:implementation-infrastructure}

ADL Lite v0.6.0-alpha includes a functionally complete implementation layer that is summarised here; full module descriptions are in Appendix~\ref{app:implementation}. \textbf{Calibration} (\texttt{calibration.py}) provides four strategies (linear, EWMA, epistemic-context, per-band) for accuracy-weighted confidence aggregation, mitigating overconfidence in $\gamma(C)$. \textbf{Semantic Web interoperability} (\texttt{owl\_export.py}, \texttt{owl\_import.py}, \texttt{rdfstar\_export.py}) enables bidirectional OWL~2 DL and RDF-star export. \textbf{Split-lock concurrency} (\texttt{EventChain}) uses dual \texttt{threading.RLock} instances to support $10^{4}$-agent contention without data races (E28), with incremental verification caching for $O(1)$ post-append checks. \textbf{Vector search} (\texttt{vector\_index.py}) provides a FAISS-backed ANN index for semantic near-duplicate detection. \textbf{Merkle anchors} (\texttt{merkle.py}) enable $O(\log n)$ batch inclusion proofs. \textbf{LLM canonicalisation} (\texttt{canonicalization.py}) clusters concepts by embedding similarity and proposes auditable merge actions via an LLM backend, with a mandatory dry-run gate. \textbf{REST API} (\texttt{api.py}, 481 lines) and \textbf{MCP server} (\texttt{mcp\_server.py}, 458 lines) provide programmatic and agent-protocol interfaces. \textbf{SHACL validation} (\texttt{shacl\_validation.py}, 334 lines) checks exported RDF conformance. \textbf{Neo4j adapter} (\texttt{neo4j\_adapter.py}, 148 lines) provides a scalable graph backend. \textbf{CRDT multi-way merge} generalises binary merge to $N\geq3$ chains via \texttt{functools.reduce}, preserving commutativity, associativity, and idempotence (Theorem~9).
