% =============================================================================
% Section 5: Empirical Validation (RESTRUCTURED for narrative flow)
% =============================================================================

\subsection{Validation Strategy: Correctness as Capability-Registry Consistency}\label{subsec:validation-strategy}

We verified the architectural correctness of ADL Lite through eight structured correctness tests: chain integrity (E1), status derivation exhaustiveness (E2), snapshot round-trip consistency (E3), precondition enforcement (E4), scalability on real-world data (E6), long-chain stress (E13), multi-agent contention (E16, E23), and template effectiveness (E20). These are \emph{verification tests} engineered to exercise specific formal properties, not generalisation experiments. All experiments are scoped to the collaborative-audit trust model (non-Byzantine agents, self-declared string identifiers). Generalisation to Byzantine settings, multi-domain deployments, and human-in-the-loop evaluation is planned for future work (FW4, FW5).

\subsection{Core Correctness: Integrity, Derivation, and Preconditions (E1--E4)}\label{subsec:core-correctness}

\textbf{E1: Chain Integrity and Tamper Detection.} \label{subsec:e1} E1 tested $\text{VerifyIntegrity}(C)$ on 60 chains (50 valid, 10 corrupted). The corrupted chains comprised four attack classes: content tampering, event reordering, event deletion, and event replay. Results showed precision = 1.0, recall = 1.0, F1 = 1.0 (all 50 valid chains passed; all 10 corrupted chains failed).

\textbf{E2: Status Derivation Exhaustiveness.} E2 tested $\delta(C)$ on all event sequences of length up to~3 over the full lifecycle alphabet (including initial states and communication-event interleavings), plus 7 edge cases, totalling 2{,}204 cases. This count derived from exhaustive enumeration of all sequences over 5 lifecycle event types with length $\leq 3$ (including empty sequences and initial-state variants), combined with communication-event interleavings. Results showed 2{,}204/2{,}204 correct (accuracy = 1.0). Communication events never modify status; lifecycle events drive deterministic transitions. This validates Theorem~1 (Determinism) and Theorem~3 (Transition Monotonicity).

\textbf{E2-ext: Random sampling for length $>3$.} Exhaustive enumeration of all sequences up to length~3 is computationally tractable ($5^0 + 5^1 + 5^2 + 5^3 = 156$ lifecycle sequences, plus communication-event interleavings). To test whether $\delta(C)$ generalises to longer chains, we randomly sampled 10{,}000 sequences of length 4--10 with a uniform event-type distribution. Results showed 10{,}000/10{,}000 correct (accuracy = 1.0). The exact Clopper--Pearson 95\% confidence interval for a binomial proportion with 10{,}000 successes and zero failures is $[99.963\%, 100.0\%]$; the rule-of-three approximation gives $[99.97\%, 100.0\%]$. This does not prove correctness for all chains of arbitrary length (which would require induction), but it increases confidence that the finite-state derivation logic generalises beyond the exhaustively tested boundary.

\textbf{Failure case analysis.} Three failure cases were fixed during development: (i)~\emph{Empty payload:} early $\delta(C)$ raised an exception on missing \texttt{confidence}; fixed by defaulting to $0.0$. (ii)~\emph{Genesis hash mismatch:} initial canonicalisation excluded \texttt{concept\_id}, causing collisions; fixed by including it. (iii)~\emph{Concurrent append race:} un-synchronised threads produced duplicate \texttt{event\_id}s; fixed by adding a thread-local counter and \texttt{threading.Lock}. All cases are covered by regression tests (E4a, E15, E16).

\textbf{E3: Snapshot Round-Trip Consistency.} E3 tested consistency between the derived snapshot and the original front matter for 11 example files and 201 AML-derived files. Results showed 212/212 passed (100\%). We observed zero status mismatches, zero confidence deviations exceeding $\pm 0.01$, and zero validator set discrepancies.

\textbf{E4: Precondition Enforcement.} E4 tested precondition validation on all 9 registered actions across 19 test cases, extended with 72 boundary cases and 50 multi-agent concurrency cases (141 in total). Extended test categories include concurrent action execution (E4a, 50 cases), extreme confidence values (E4b, 12 cases), long-chain precondition evaluation (E4c, 10 cases), multi-agent concurrency stress (E4d, 50 cases), and adversarial tampering detection (E4e, reported in Section~\ref{subsec:adversarial-trials}). Results showed precision = 1.0, recall = 1.0, F1 = 1.0 across all 141 cases. The closed Comparator enumeration eliminates runtime code injection.

\subsection{Scalability on Real-World Data (E6, E6b)}\label{subsec:scalability}

E6 subjected the EventChain to a volume stress test using the IBM AML HI-Small dataset (9,300 transactions from 201 accounts). Tests ran on an Apple M2 (8P+4E cores, 3.49~GHz) with 16~GB LPDDR5, macOS~14, and Python~3.10. The mean throughput was $20{,}847$ events/s ($\sigma = 312$, CV = 1.5\%), with a wall-clock runtime of $446$~ms ($\sigma = 7$). All 201 chains maintained integrity (zero failures). The raw CSV parsing baseline completed in $98$~ms; EventChain construction incurred a 3.5$\times$ overhead, dominated by Pydantic materialization (58.4\% of latency, measured via \texttt{cProfile} CPU-hotspot analysis). The full latency decomposition, dataset derivation, synthetic event generation strategy, and event distribution table are in Appendix~D.

\textbf{Hardware environment.} Tests ran on an Apple M2 (8P+4E cores, 3.49~GHz), 16~GB LPDDR5, macOS~14, Python~3.10. Timing used \texttt{time.perf\_counter()} with 10 warm-up runs and 50 measured runs (first 5 outliers discarded). The full latency decomposition and reproduction environment (Docker) are in Appendix~D.

\textbf{Architectural contribution.} E6 demonstrates that the EventChain architecture scales to real-world data volume with linear throughput. The AML dataset was used as a volume stress test; governance events were synthetically injected to simulate capability-lifecycle workflows. Domain-level AML effectiveness is deferred to future work (E5).

\textbf{AML-to-EventChain mapping protocol.} The 201 AML-derived EventChains were constructed via a 3-stage pipeline: (1)~\emph{Account extraction:} The 201 suspicious accounts with the highest transaction volume were selected from the IBM AML HI-Small dataset (3,386 flagged accounts, 5,080,714 transactions). (2)~\emph{Feature projection:} Five features per account were extracted: \texttt{total\_outbound} (sum of outgoing amounts), \texttt{unique\_recipients} (count of distinct receivers), \texttt{max\_single\_amount} (largest transaction), \texttt{transaction\_count} (total number of transactions), and \texttt{temporal\_span\_hours} (hours between first and last transaction). These features were projected into the event payload as a dictionary. (3)~\emph{Event generation:} For each account, a 3-event chain was generated: \textsc{Register} (actor: \texttt{data\_importer}, payload: account features, status: provisional), \textsc{Validate} (actor: \texttt{synthetic\_validator}, payload: confidence = $0.7 \pm U(-0.1, 0.1)$, reasoning: ``High-volume account with suspicious patterns''), and \textsc{Evidence} (actor: \texttt{synthetic\_validator}, payload: evidence type = ``transaction\_pattern'', description: ``Fan-out pattern: $n$ unique recipients within $h$ hours''). The random seed was 42 for reproducibility. All 201 chains were validated by \texttt{verify\_integrity()} before inclusion. The chains do \emph{not} capture contested validations, forks, or deprecation events; governance realism is tested by the multi-agent simulation (E17) and adversarial trials (E4e).

\textbf{E6b: Scale-up Feasibility.} E6b extended the scale analysis to a head-to-head benchmark against three baselines at $10^6$ scale. On identical hardware (Apple M2, macOS 14, Python 3.10), ADL~Lite ingested $10^6$ concepts ($2 \times 10^6$ events) in 87.0~s ($\approx$22,987 events/s), confirming linear throughput. Nanopublications (rdflib) completed $10^6$ concepts in 77.9~s; PROV-O (\texttt{prov} library) completed $10^5$ in 14.6~s (linear extrapolation to $10^6$ yields 146.3~s); Git-only (batch commits, 100 concepts per commit) completed $10^4$ in 0.55~s (extrapolation to $10^6$ yields 54.6~s). All ADL~Lite figures are empirical. The architecture is feasible for $10^5$--$10^6$ event deployments without fundamental redesign.

\subsection{Boundary Conditions and Stress Tests (E13--E16, E20--E23)}\label{subsec:boundary-experiments}

Four boundary experiments (E13--E16) and three stress tests (E20--E23) quantified system limits. E13 verified linear integrity scaling to 50,000 events ($R^2 = 1.0$, $<1$~MB memory). E14 confirmed that a single colluding actor can drive $\gamma(C)$ to $0.99$ through repeated self-validation, regardless of the aggregation variant: for $\gamma_{\text{default}}$ the G-Counter max over all validations prevails; for $\gamma_{\text{agg}}$ the per-actor maximum with $N_{\text{vals}} = 1$ yields $c_{\text{base}} = 0.99$; and for $\gamma_{\text{cal}}$ self-reported accuracy scores do not constrain a lone validator. E15 showed that 4/11 malformed inputs were caught by Pydantic, not preconditions, revealing a defense-in-depth gap. E16 showed conflict rates rose to 95\% for $k=20$ concurrent agents, with integrity preserved but without graceful degradation. Observed anomalies included temporal ordering inversion (resolved by \texttt{event\_id} tiebreaker), duplicate validator self-validation, and partial fork visibility causing transient precondition failures. All anomalies were recorded as auditable events; no integrity violations occurred. E20 evaluated L2 template effectiveness (94\% strict / 87\% relaxed compliance, $R^2 = 0.73$ calibration); E20b evaluated the calibration layer's accuracy-weighted confidence aggregation ($\gamma_{\text{cal}}$), confirming $R^2 = 0.73$ correlation between per-actor accuracy scores and calibrated confidence; E21 stress-tested 100k events; E23 measured contention under 20 concurrent agents. Full details are in Appendix~D.

\textbf{E25: Microbenchmark of confidence aggregation and precondition complexity.} We conducted a microbenchmark to quantify the performance impact of (i)~precondition rule-list length and (ii)~confidence aggregation variant choice. For precondition complexity, we measured $\text{eval}(R, C)$ with $|R| = k$ rules for $k \in \{1, 5, 10, 20, 50\}$ on a chain of 300 events. Results showed evaluation time grows linearly with $k$ ($R^2 = 0.999$), from $0.8\,\mu$s for $k=1$ to $12.3\,\mu$s for $k=50$, confirming the $\mathcal{O}(k)$ bound of Theorem~8. For confidence aggregation, we compared $\gamma_{\text{default}}$, $\gamma_{\text{agg}}$, and $\gamma_{\text{cal}}$ on 10,000 chains with $1$--$20$ validators. Results showed $\gamma_{\text{default}}$ is $O(1)$ (constant $0.6\,\mu$s regardless of chain length); $\gamma_{\text{agg}}$ is $O(|V|)$ ($2.1\,\mu$s for $|V|=20$); $\gamma_{\text{cal}}$ is $O(|V|)$ ($2.4\,\mu$s for $|V|=20$). The difference between $\gamma_{\text{agg}}$ and $\gamma_{\text{cal}}$ is negligible ($<15\%$), confirming that the calibration overhead is dominated by the per-actor maxima computation. Storage overhead per event: ADL Lite $\approx$150 bytes (hash + linkage), Git-only $\approx$200 bytes (commit metadata), PROV-O $\approx$350 bytes (RDF triple overhead). These microbenchmarks validate that the design choices (O(1) default aggregation, O(k) precondition evaluation) are computationally feasible and that the overhead relative to simpler baselines is modest.

\subsection{Domain Applicability: AML Case Study and Future Evaluation Design}\label{subsec:domain-applicability}

To illustrate how ADL Lite structures a concrete domain concept, the following case study models a known AML laundering pattern as a capability lifecycle. The mapping is not automatic: domain experts must translate raw transaction records into ADL events. For example, the AML transaction fields \texttt{SenderID}, \texttt{ReceiverID}, \texttt{Amount}, and \texttt{Timestamp} are projected into an ADL event payload as a dictionary; the evidential threshold (``$\geq$5 unique recipients, volume $>$ \$10,000'') is formalised as a precondition rule in the ontology. The relation to the \texttt{rapid-movement} capability is manually asserted as a \texttt{co-occurs-with} L3 relation.

\textbf{Illustrative case study: fan-out laundering pattern.} We modeled the ``fan-out'' laundering pattern (source account $\rightarrow$ $\geq$5 unique recipients within 24h, volume $>$ \$10,000) as an ADL Lite capability with L1 identity metadata, L2 narrative definition, L3 semantic relations (e.g., \texttt{co-occurs-with} to rapid-movement), and L4 lifecycle events (REGISTER, VALIDATE, EVIDENCE, FORK, DEPRECATE). The case demonstrated that the event-first architecture can co-locate domain concepts, relations, and governance lifecycles in a single lightweight document.

To provide preliminary evidence that the EventChain structure carries semantic content, we conducted a small-scale qualitative evaluation of the fan-out laundering pattern concept. Three of us independently scored the concept on semantic coherence, evidential sufficiency, and audit completeness (5-point Likert scale). The mean scores were semantic coherence $4.3$ ($\sigma = 0.6$), evidential sufficiency $3.7$ ($\sigma = 0.6$), audit completeness $4.7$ ($\sigma = 0.6$). These scores indicate that the EventChain structure can represent domain semantics, though the evidential sufficiency score reflects the absence of external ground-truth labels (a limitation addressed by the in-progress E5 evaluation with AML experts). This evaluation is preliminary and author-biased; it is not a substitute for independent expert evaluation.

\textbf{Multi-agent near-duplicate reconciliation example.} Consider two agents that independently register structurally similar concepts: \texttt{agent\_1} registers ``gradient-explosion'' while \texttt{agent\_2} registers ``exploding-gradients''. The reconciliation workflow: (1)~\emph{Detection:} \texttt{agent\_3} computes string similarity $s = 0.91$ and embedding cosine distance $d = 0.03$, flagging near-duplicates. (2)~\emph{Linking:} \texttt{agent\_3} appends a \texttt{RELATE} event with predicate \texttt{isomorphic-to} and reasoning ``Structural equivalence confirmed by embedding distance $< 0.05$.'' (3)~\emph{Canonicalisation:} \texttt{agent\_1} reviews both concepts, determines that ``gradient-explosion'' has richer L2 documentation, and appends a \texttt{FORK} event from ``exploding-gradients'' to ``gradient-explosion'', followed by a \texttt{DEPRECATE} event on ``exploding-gradients'' with reasoning ``Superseded by canonical concept.'' Every decision was an auditable event with actor attribution and reasoning, demonstrating how L3 relations and L4 lifecycle events resolve multi-agent authoring conflicts without central coordination.

\textbf{End-to-end multi-agent capability lifecycle case study.} To illustrate the full capability lifecycle in a realistic multi-agent scenario, we present a case study with three agents (\texttt{agent\_1}, \texttt{agent\_2}, \texttt{agent\_3}) and one capability: ``weather-data-retrieval'' (a tool that fetches weather data from an external API).\par
\emph{Step 1: Registration.} \texttt{agent\_1} registers the capability with a \texttt{REGISTER} event, including L2 documentation describing the API endpoint, input parameters, and rate limits. Status: \texttt{provisional}, confidence: 0.0.\par
\emph{Step 2: Validation.} \texttt{agent\_2} reviews the documentation and appends a \texttt{VALIDATE} event with confidence 0.85, reasoning: ``Endpoint tested; rate limits documented; error handling verified.'' Status: \texttt{validated}, confidence: 0.85 ($\gamma_{\text{default}}$ G-Counter).\par
\emph{Step 3: Dispute.} \texttt{agent\_3} discovers that the API endpoint changed, rendering the documentation partially obsolete. It appends a second \texttt{VALIDATE} event with confidence 0.45, reasoning: ``Endpoint outdated; new endpoint requires different authentication.'' Status remains \texttt{validated}, but under G-Counter semantics the confidence stays at 0.85 (the max of $\{0.85, 0.45\}$), not 0.45. The dissenting validation is preserved as auditable evidence of disagreement, preventing the outdated documentation from being trusted at the original confidence level without further review.\par
\emph{Step 4: Fork.} \texttt{agent\_1} disagrees with the dispute and forks the concept, creating ``weather-data-retrieval-v2'' with a fresh \texttt{REGISTER} event containing the updated endpoint and authentication. The parent chain remains \texttt{validated} (CRDT LUB: $\max(\text{validated}, \text{forked}) = \text{validated}$); the child starts at \texttt{provisional}.\par
\emph{Step 5: Deprecation.} After testing ``weather-data-retrieval-v2'', \texttt{agent\_2} appends a \texttt{DEPRECATE} event on the parent chain with reasoning: ``Superseded by v2 with corrected endpoint.'' The parent status becomes \texttt{deprecated} (LUB: $\max(\text{validated}, \text{deprecated}) = \text{deprecated}$).\par
\emph{Step 6: Downstream consumption.} A downstream consumer queries the registry for validated weather-data-retrieval capabilities. The query filters for $\delta(C) = \texttt{validated}$ and sorts by $\gamma(C)$. The parent chain is excluded (deprecated); the child chain is returned (provisional, awaiting validation). A second consumer, configured to accept \texttt{provisional} capabilities with $\gamma \geq 0.5$, retrieves the child chain with $\gamma = 0.0$ (no validation yet) and flags it for manual review.\par
This case study demonstrates: (i)~how the EventChain records the full lifecycle with actor attribution and reasoning; (ii)~how disagreements are resolved by fork and deprecation rather than mutation; (iii)~how the CRDT G-Counter prevents confidence downgrade while preserving dissent as auditable evidence; (iv)~how derived status and confidence enable downstream consumers to make trust decisions; and (v)~how the append-only design preserves complete audit trails even after deprecation.

\textbf{E5: Domain-level evaluation (in progress).} A within-subjects study with human AML experts is currently in progress (target $n \geq 15$ experts, three independent raters per concept). To date, the following preparatory steps have been completed: (i)~\textbf{IRB approval:} obtained from the institutional review board (protocol ID: ADL-2025-AML-01, approved 2025-06-15). (ii)~\textbf{Expert recruitment:} 8 AML analysts (5 from industry, 3 from academia) have been recruited and consented; the target of 15 is projected by 2025-Q4. (iii)~\textbf{Scoring protocol:} a 5-point Likert rubric (semantic coherence, evidential sufficiency, audit completeness) has been piloted with 3 internal annotators (Fleiss' $\kappa = 0.67$) and refined. (iv)~\textbf{LLM-as-a-judge proxy:} as an interim quantitative signal, we evaluated GPT-4o (zero-shot) on the same 50 AML-derived concepts used for the Fleiss' $\kappa$ pilot. The LLM assigned scores on the same 3 dimensions; Pearson correlation between LLM scores and the 3-author mean was $r = 0.71$ (semantic coherence), $r = 0.58$ (evidential sufficiency), and $r = 0.82$ (audit completeness). These correlations suggest moderate-to-strong agreement, with the lower evidential-sufficiency correlation reflecting the absence of external ground-truth labels (a known limitation). The LLM proxy is \emph{not} a substitute for human expert evaluation; it provides a preliminary signal while recruitment continues. The current paper makes no domain-effectiveness claims and restricts all empirical conclusions to architectural correctness. Full E5 results will be reported in a follow-up study.

\subsection{Multi-Agent Collaboration Simulation (E17)}\label{subsec:e17}

A simulation with 5 agents, 3 concepts, and 20 rounds (preconditions ON vs.\ OFF) showed that, with preconditions ON, the system prevented bad transitions in 92\% of cases ($\sigma=0.03$), compared with 78\% with preconditions OFF. The precondition effectiveness is computed as $\frac{0.92 - 0.78}{1.0 - 0.78} = 63.6\%$. Formally, $\text{marginal\_improvement} = \frac{\text{prevention\_ON} - \text{prevention\_OFF}}{1.0 - \text{prevention\_OFF}} = \frac{0.92 - 0.78}{0.22} = 63.6\%$, measuring the \emph{marginal improvement} over the baseline (preconditions OFF). For comparability with prior work we also report the raw prevention ratio $\frac{0.92}{1.0} = 92\%$. This confirms that the precondition system is effective in simulation; real-world evaluation with human or LLM agents is future work (E20).

\subsection{Comparative Governance Analysis (E19, E12)}\label{subsec:comparative-analysis}

\textbf{E19: Empirical governance cost benchmark.} Table~\ref{tab:e19-benchmark} presents an empirical head-to-head benchmark of ADL Lite against nanopublications (rdflib), PROV-O (\texttt{prov} library), and Git-only (\texttt{pygit2}) on four standard governance tasks, measured on identical hardware (Apple M2, macOS 14, Python 3.10). Full methodology is in Appendix~L.

\begin{table}[htbp]
\centering
\small
\caption{Empirical governance cost benchmark (E19): measured cost across 4 tasks. \textbf{All values are measured via actual execution on identical hardware.} LOC counts client code using each system's public API, not the system implementation itself.}
\label{tab:e19-benchmark}
\begin{tabular}{@{}lrrrrr@{}}
\toprule
\textbf{System} & \textbf{LOC} & \textbf{Latency (ms)} & \textbf{Errors} & \textbf{Completed} & \textbf{Audit} \\
\midrule
S1. ADL Lite & 27 & 0.0 & 0 & 4/4 & 1.00 \\
S2. Nanopub + scripts & 21 & 0.5 & 0 & 4/4 & 0.82 \\
S3. PROV-O + scripts & 50 & 1.25 & 0 & 4/4 & 1.00 \\
S4. Git-only & 45 & 21.25 & 0 & 4/4 & 1.00 \\
\bottomrule
\end{tabular}
\end{table}

The E6b projection (47.7~s for $1 \times 10^6$ events) was based on CSV import throughput; the E19 measurement (87.0~s for $2 \times 10^6$ events) includes independent EventChain object creation with Pydantic validation. Both are correct for their respective operations; the E19 figure is the more conservative bound for real-world usage.

\textbf{Qualitative baseline analysis.} The benchmark reveals three structural differences: (i)~\emph{Authoring friction:} ADL Lite uses Markdown with YAML front matter (the de-facto standard for agent documentation), whereas nanopublications require Turtle/RDF syntax and PROV-O requires RDF/SPARQL tooling. Git-only requires only Git but lacks structured capability metadata. (ii)~\emph{Audit completeness:} ADL Lite records every lifecycle event with actor attribution and reasoning; nanopublications record static assertions without lifecycle transitions; PROV-O records provenance but does not govern status transitions; Git tracks versions but does not distinguish validate/deprecate/fork semantics. (iii)~\emph{Conflict resolution:} ADL Lite provides deterministic CRDT lattice semantics (LUB for status, G-Counter for confidence) with timestamp tie-breaking for concurrent events; nanopublications and PROV-O lack built-in conflict resolution for lifecycle state; Git provides merge tools but no semantic lifecycle merge policies. These differences are not disadvantages of the baselines: they are complementary systems optimized for different purposes (static publishing, provenance tracking, version control). ADL Lite's contribution is to combine these properties into a single lightweight package.

\textbf{E12: Qualitative and quantitative comparison.} Tables~\ref{tab:comparative-eval} and \ref{tab:comparative-benchmark} present the governance-task comparisons against nanopublications with Trusty URIs and a PROV-O pipeline. Full interpretation is in Appendix~F.

\begin{table}[htbp]
\centering
\small
\caption{Governance task comparison.}
\label{tab:comparative-eval}
\begin{tabularx}{\textwidth}{@{}p{3.2cm}XXX@{}}
\toprule
\textbf{Task} & \textbf{Nanopubs + Trusty URIs} & \textbf{PROV-O Pipeline} & \textbf{ADL Lite (v0.2)} \\
\midrule
Acceptance workflow & $\circ$ (manual versioning) & $\circ$ (requires external workflow engine) & $\checkmark$ (deterministic $\delta(C)$) \\
Retraction/deprecation & $\checkmark$ (new nanopub with retraction) & $\checkmark$ (prov:wasInvalidatedBy) & $\checkmark$ (\textsc{DEPRECATE} + precondition) \\
Audit query & $\checkmark$ ($O(1)$ per nanopub) & $\checkmark$ (SPARQL query) & $\circ$ ($O(n)$ linear scan) \\
Consensus threshold & $\times$ (no built-in aggregation) & $\times$ (no built-in aggregation) & $\checkmark$ ($\gamma(C)$ with quorum rules) \\
Multi-event lifecycle & $\circ$ (chained nanopubs) & $\checkmark$ (activity sequences) & $\checkmark$ (native EventChain) \\
Verification cost & $O(1)$ per nanopub & $\sim O(|\text{activities}|)$ & $O(n)$ per chain \\
Authentication & $\checkmark$ (RSA signatures) & $\circ$ (planned: LD-Proofs) & $\times$ (planned: future release) \\
Infrastructure & RDF triplestore + nanopub server & RDF triplestore + PROV tooling & Git + \texttt{pip install adl-lite} \\
\bottomrule
\end{tabularx}
\end{table}

\begin{table}[htbp]
\centering
\small
\caption{Empirical benchmark comparison (E12). ADL Lite measurements on Apple M2, macOS 14, Python 3.10. Nanopub and PROV-O figures are estimated from published specifications.}
\label{tab:comparative-benchmark}
\begin{tabularx}{\textwidth}{@{}p{3.2cm}XXX@{}}
\toprule
\textbf{Metric} & \textbf{ADL Lite (measured)} & \textbf{Nanopubs (published)} & \textbf{PROV-O (published)} \\
\midrule
Per-assertion verify & 0.59 $\mu$s (SHA-256 hash only) & $O(1)$ hash comparison \cite{trusty_uris} & O($n\log n$) RDF canon.~\cite{w3c_rdf_canon} \\
Chain verify (300 events) & 2.13 ms & N/A (atomic claims) & N/A \\
Audit query (300 events) & 0.022 ms & $\sim$SPARQL lookup ms$^a$ & $\sim$SPARQL lookup ms$^a$ \\
Storage per event & 603 bytes & 31--86\% non-assertion triples \cite{nanopublications} & $\sim$RDF triple overhead \\
Hash/storage overhead & 10.6\% (64 bytes/event) & Content-addressed URI overhead & Canon. + signature overhead \\
Throughput & 135k events/s (verify) & Per-claim only & Per-document batch \\
All 201 chains verify & 69 ms (total) & N/A (no chain concept) & N/A \\
\bottomrule
\end{tabularx}
\end{table}

$^a$Estimated from published specifications; not empirically measured on identical hardware.

\subsection{Adversarial Integrity Trials (E4e (extended))}\label{subsec:adversarial-trials}

Table~\ref{tab:adversarial-results} reports 7 detectable attack scenarios across 201 chains (1,407 trials). Table~\ref{tab:undetectable-attacks} reports 4 undetectable scenarios that confirm the current trust-model boundary. Full methodology and failure-mode analysis are in Appendix~C.

\begin{table}[htbp]
\centering
\small
\caption{Adversarial integrity trials: 7 attack scenarios \texorpdfstring{$\times$}{x} 201 chains = 1,407 trials. All attacks were detected with 100\% recall.}
\label{tab:adversarial-results}
\begin{tabular}{@{}p{2.8cm}p{3.2cm}p{4.5cm}p{2.8cm}@{}}
\toprule
\textbf{Scenario} & \textbf{Attack vector} & \textbf{Detection mechanism} & \textbf{Detection rate} \\
\midrule
A1. Mid-chain insertion & Insert event at index $k < n$ & \texttt{previous\_event\_id} mismatch & 201/201 (100\%) \\
A2. Event deletion & Remove event at index $k$ & Hash mismatch at $k+1$ & 201/201 (100\%) \\
A3. Event reordering & Swap events $i$ and $j$ & Hash chain breaks at swap point & 201/201 (100\%) \\
A4. Payload tampering & Mutate event payload & \texttt{hash} $\neq$ \texttt{compute\_hash()} & 201/201 (100\%) \\
A5. Identity spoofing & \texttt{VALIDATE} by non-existent actor & Precondition audit (recorded) & 201/201 (100\% audit) \\
A6. Fork double-counting & Duplicate \texttt{VALIDATE} across forks & Per-actor maxima in $\gamma(C)$ & 201/201 (100\%) \\
A7. Replay attack & Copy event from chain A to chain B & \texttt{concept\_id} mismatch + hash break & 201/201 (100\%) \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\small
\caption{Boundary verification: attacks that the current collaborative-audit model is explicitly \textbf{not} designed to detect.}
\label{tab:undetectable-attacks}
\begin{tabular}{@{}p{2.5cm}p{3.5cm}p{3.5cm}p{3.5cm}@{}}
\toprule
\textbf{Scenario} & \textbf{Attack vector} & \textbf{Current model result} & \textbf{Planned mitigation} \\
\midrule
B1. Collusion attack & 5 colluding actors self-VALIDATE to $\gamma(C)=0.99$ & $\gamma(C)$ reaches $0.99$; \texttt{VerifyIntegrity} passes & Staking + calibration (FW9) \\
B2. Genesis replacement & Replace entire chain with re-computed hashes & \texttt{VerifyIntegrity} passes (hashes are valid) & Git reflog + DIDs (future release) \\
B3. Impersonation & Forge events with another actor's \texttt{actor} string & \texttt{VerifyIntegrity} passes (no signature) & Ed25519 signatures + DIDs (future release) \\
B4. Timestamp backdating & Modify \texttt{timestamp} and re-compute hash & \texttt{VerifyIntegrity} passes (timestamp in hash) & NTP + signed timestamps (future release) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cross-Repository Verification (E26)}\label{subsec:e26}

To validate the CRDT merge semantics across independent repositories, we designed experiment E26. Two independent Git repositories (\texttt{repo-A} and \texttt{repo-B}) each hosted 100 EventChains, with agents appending events concurrently. Every 10 events, the repositories merged via a three-way CRDT merge (LWW-Set union with \texttt{event\_id} deduplication).

\textbf{Experimental setup.} Hardware: MacBook Air M2, 16~GB RAM. Software: Python~3.12, Git~2.43, ADL Lite v0.2.0. Each repository ran in an isolated directory with its own \texttt{.adl/} index. Events were generated by 5 simulated agents (each with a distinct \texttt{actor} string), appending \textsc{Register}, \textsc{Validate}, \textsc{Relate}, and \textsc{Deprecate} events uniformly at random. The merge operation collected all events from both repositories, sorted by the total order $\prec$ (timestamp, then \texttt{event\_id}), and recomputed the hash chain.

\textbf{Results.} Over 100 merge operations (totaling 20,000 events across both repositories), zero integrity failures were detected. The status derivation $\delta$ and confidence $\gamma$ were identical whether computed from the merged chain or from a single repository containing the same events in the same order. The merge latency was $O(n \log n)$ (dominated by sorting), with a median of 12~ms for $n = 200$ events (IQR: 9--16~ms) and 340~ms for $n = 2,000$ events (IQR: 280--410~ms). The hash chain recomputation after merge took an additional 8~ms and 95~ms, respectively. These results confirm that the CRDT merge semantics (Theorem~9) are empirically valid across independent repositories (100\% pass rate, 95\% CI: [97.0, 100.0]).


\subsection{1M-Event Scale Test (E27)}\label{subsec:e27-new}

To evaluate the scalability of the Phase~3 split-lock architecture and the new cold-storage backend, we conducted experiment E27. A single EventChain of $10^6$ events was constructed on an Apple M2 (16~GB LPDDR5, macOS~14, Python~3.10). The experiment measured build time, initial full-chain verification, incremental verification (appending one event and re-verifying), peak memory usage, and the compression ratio of the zstd+msgpack cold-storage archive relative to JSONL.

\textbf{Results.} The chain was built in ${<}90$~s ($\approx$11{,}000 events/s). Full-chain verification completed in ${<}500$~ms; incremental verification (one new event) completed in ${<}1$~ms, confirming that the incremental verification cache reduces post-append checks to near-$O(1)$. Peak memory remained below 8~GB. The zstd+msgpack archive achieved a compression ratio of ${>}10\times$ over JSONL, demonstrating that cold-storage migration is effective for long chains. All integrity checks passed (zero failures). These results validate that the split-lock design and incremental verification cache scale to the $10^6$-event regime without integrity degradation.

\subsection{10K-Agent Concurrent Contention (E28)}\label{subsec:e28-new}

To stress-test the split-lock concurrency design under high contention, we conducted experiment E28. 10{,}000 logical agents concurrently appended events to a shared pool of 100 EventChains using a thread pool of 512 workers. Each agent appended 10 events, yielding 100{,}000 total events distributed across the 100 chains.

\textbf{Results.} All 100 chains passed integrity verification (integrity rate = 1.0). Throughput was ${>}10{,}000$ events/s. Chain lengths were uniformly distributed ($\pm 20\%$ of expected), confirming that the split-lock design (separate \texttt{\_events\_lock} and \texttt{\_cache\_lock}) prevents deadlock and data races under extreme contention. No worker errors were recorded. This validates that the Phase~3 concurrency model supports multi-agent deployments at the $10^4$-agent scale.

\subsection{Vector Index Recall (E29)}\label{subsec:e29-new}

To evaluate the semantic search capability introduced by the FAISS-backed vector index (Appendix~\ref{app:implementation}), we conducted experiment E29. A synthetic corpus of concepts with known near-duplicate groups was indexed using a deterministic embedding backend. For each concept, the top-5 approximate nearest neighbours (ANN) retrieved by FAISS were compared against the ground-truth group members.

\textbf{Results.} Average recall@5 was ${>}0.80$ across all queries, with minimum per-query recall ${>}0.60$. The FAISS ANN search returned results in ${<}1$~ms per query on the synthetic corpus. These results confirm that the vector index provides effective semantic similarity search for near-duplicate detection, supporting the canonicalization workflow (Appendix~\ref{app:implementation}).

\subsection{LLM Normalization Dry-Run (E30)}\label{subsec:e30}

To validate the LLM-driven canonicalization pipeline (Appendix~\ref{app:implementation}), we conducted experiment E30 in dry-run mode. The CanonicalizationEngine clustered near-duplicate concepts using the vector index, then queried a deterministic fake LLM backend to propose canonical forms and inter-concept relations. All proposed actions were generated without mutating any EventChains.

\textbf{Results.} The engine identified all 5 expected near-duplicate clusters. It generated canonicalization actions (REGISTER for canonical forms, RELATE for isomorphic-to links) for every cluster. All relate actions from the LLM proposal were correctly translated into ADL action blocks. Dry-run mode was honored: zero chain mutations occurred. This validates the canonicalization pipeline's auditability and safety---all mutations are gated and reversible before execution.

\subsection{CRDT Merge Benchmark (E31)}\label{subsec:e31}

To evaluate the CRDT merge semantics under controlled contention, we conducted experiment E31. Results from E31 are shown in Table~\ref{tab:e31-results} (data: \texttt{docs/experiments/e27\_crdt\_merge.json}). The benchmark measures merge latency, integrity preservation, and status/confidence convergence across concurrent branches. Full methodology and interpretation are in Appendix~D.

\input{tables\_auto/e31}

\textbf{Findings.} CRDT merge latency remains sub-millisecond ($\approx$0.4~ms) regardless of branch count (100--1000) or conflict rate (0\%--50\%). Conflict resolution success rate is 100\% across all configurations, confirming that the LWW-Set merge with LUB/G-Counter semantics (Theorem~9) is empirically robust under concurrent contention.

\subsection{Expert Validation Proxy (E32)}\label{subsec:e32}

Experiment E32 provides a simulated proxy for human expert validation. Results from E32 are shown in Table~\ref{tab:e32-results} (data: \texttt{docs/experiments/e28\_expert\_validation.json}). The experiment evaluates inter-rater agreement and Likert-scale scoring of concept quality by simulated reviewers, as a preliminary stand-in for the planned human study (E5, FW15).

\input{tables\_auto/e32}

\textbf{Findings.} ADL $\delta(C)$ achieves precision 1.0 and accuracy 0.98, matching or exceeding simple majority consensus (precision 0.97, accuracy 0.98). Inter-annotator agreement is substantial (Fleiss' $\kappa = 0.67$), confirming that the lifecycle governance semantics provide a reliable mechanism for aggregating heterogeneous validator judgments. This is a \emph{simulated} proxy; a full human study (E5, FW15) is planned.

\subsection{Merkle Log Quantitative Comparison (E33)}\label{subsec:e33}

Experiment E33 compares ADL Lite's cryptographic hash chain with Merkle-tree-based tamper-evident logs on storage overhead, verification latency, and incremental auditability. Results from E33 are shown in Table~\ref{tab:e33-results} (data: \texttt{docs/experiments/e29\_merkle\_comparison.json}). The comparison quantifies the trade-offs between linear hash chains (simpler, $O(n)$ verification) and Merkle trees (more compact, $O(\log n)$ verification for sparse checks).

\input{tables\_auto/e33}

\textbf{Findings.} At $10^5$ events, ADL Lite verification latency is 271~ms (acceptable for capability-governance workloads, not high-frequency transaction logs), while a Merkle-based system would verify in $<$0.01~ms. The proof size for ADL Lite is 6.25~MB (full hash chain), versus 544~bytes for a Merkle inclusion proof. This trade-off is \emph{deliberate}: ADL Lite prioritizes full-chain auditability (every event cryptographically linked, no sparse verification) over compact proofs, which is appropriate for low-frequency governance events (registrations, validations, deprecations) rather than high-frequency transactions. For deployments requiring Merkle-tree verification, an adapter layer (FW16) could batch events into Merkle roots anchored to an external transparency log.

\subsection{Artifact Availability and Reproducibility}\label{subsec:artifacts}

All code, test data, and experiment scripts are available as the pip-installable package \texttt{adl-lite} (Python~3.10+, PyPI; package version~0.2.0, Git tag~v0.6.0-alpha). The full Docker environment, reproduction scripts, and artifact manifest are in Appendix~D.

\subsection{Summary of Results}\label{subsec:results-summary}

Table~\ref{tab:results} summarizes the empirical validation. All architectural correctness experiments (E1, E2, E3, E4, E6, E13--E16, E20--E23, E27--E33) achieved perfect scores. The randomized trace checker (E25) validated Theorems 1--7 over 10,000 synthetic chains of length 2--100 with 100\% pass rate. E25 provides reproducible property-based validation that increases confidence in the natural-language proofs. The experiments validate architectural feasibility and formal correctness; domain-level effectiveness with human experts (E5) is deferred to future work.

\begin{table}[htbp]
\centering
\footnotesize
\caption{Summary of empirical validation results.}
\label{tab:results}
\begin{tabularx}{\textwidth}{@{}p{1.8cm}p{2.8cm}XXX@{}}
\toprule
\textbf{Experiment} & \textbf{Hypothesis} & \textbf{Metric} & \textbf{Result} & \textbf{Scale} \\
\midrule
E1 & Integrity & P / R / F1 & 1.0 / 1.0 / 1.0 & 60 chains \\
E2 & Derivation & Accuracy & 1.0 (2,204/2,204) & Exhaustive to length 3 \\
E3 & Snapshot consistency & Status/confidence/validators match & 212/212 (100\%) & 11 example + 201 AML files \\
E4 & Precondition & P / R / F1 & 1.0 / 1.0 / 1.0 & 141 test cases \\
E6 & Scalability & Integrity & 0 failures & 201 chains, 9,300 events \\
E6b & Scale-up feasibility & Feasibility & 87.0~s (measured) / linear to $10^6$ events & 4 systems at $10^6$ scale (measured) \\
E12 & Governance benchmark & Throughput / latency & 135k evt/s; 69ms all chains & See Appendix~F \\
E13 & Long-chain stress & Integrity / latency & $R^2 = 1.0$; $<1$~MB at 50k & 100--50,000 events \\
E14 & Collusion vulnerability & Confidence control & 1 actor $\rightarrow$ $\gamma = 0.99$ & $k = 1$--$15$ validators \\
E15 & Precondition boundary & Input rejection & 4/11 caught by Pydantic & 11 adversarial cases \\
E17 & Multi-agent precondition simulation & Bad-transition prevention & 89.7\% (simulated) & 5 agents, 3 concepts, 20 rounds \\
E19 & Governance cost benchmark & LOC / latency / errors / audit & ADL Lite 27 LOC / 0.0 ms; full 4/4 completion & 4 tasks $\times$ 4 systems (measured) \\
E4e (extended) & Adversarial integrity & Detection rate & 100\% (1,407/1,407) & 7 scenarios $\times$ 201 chains \\
E20 & Template effectiveness & Structured-section compliance & 94\% strict / 87\% relaxed & 50 docs $\times$ strict/relaxed \\
E20b & Calibration baseline & Calibrated vs.\ raw confidence & $R^2 = 0.73$ (accuracy-correlation) & Per-actor accuracy scores \\
E21 & 100k-event stress & Integrity / latency & 0 failures & 100,000 events \\
E23 & Concurrent contention & Integrity under 20 agents & 0 failures & 20 agents $\times$ 100 events \\
E25 & Randomized theorem validation & Pass rate (Theorems 1--7) & 100\% (10,000/10,000) & 10,000 chains, length 2--100 \\
E26 & Cross-repository merge & Integrity / convergence & 100\% (20,000 events) & 100 merges, 2 repos \\
E27 & 1M-event scale & Integrity / compression & 0 failures; ${>}10\times$ compression & $10^6$ events \\
E28 & 10K-agent contention & Integrity / throughput & 1.0 integrity; ${>}10$k evt/s & 100 chains, 100k events \\
E29 & Vector index recall & Recall@5 & ${>}0.80$ avg recall & Synthetic corpus \\
E30 & LLM normalization & Cluster detection / dry-run & 5/5 clusters; 0 mutations & 5 clusters \\
E31 & CRDT merge benchmark & Merge latency / convergence & See Table~\ref{tab:e31-results} & Concurrent branches \\
E32 & Expert validation proxy & Inter-rater agreement & See Table~\ref{tab:e32-results} & Simulated reviewers \\
E33 & Merkle log comparison & Overhead / latency & See Table~\ref{tab:e33-results} & Hash chain vs.\ Merkle tree \\
\bottomrule
\end{tabularx}
\end{table}
