% =============================================================================
% Section 6: Discussion
% =============================================================================

\subsection{Capability-First Audit: Ontological Implications}\label{subsec:ontological-implications}

The empirical results support the event-first design; the deeper contribution lies in four ontological questions.

\textbf{Q1: What is the ontological status of a capability?} In ADL Lite, a capability is a generically dependent continuant that exists only from its genesis event (\texttt{REGISTER}) onward; prior to that, there is only potential. This parallels BFO's view that a process exists only from its beginning.

\textbf{Q2: Are status and confidence real or derived?} Status is computed by $\delta(C)$ and has no existence independent of the chain, aligning with DOLCE's view that some properties are constructed by cognitive agents rather than discovered~\cite{dolce}.

\textbf{Q3: What happens to a capability when it is deprecated?} Deprecation appends a \texttt{DEPRECATE} event; the capability retains its \texttt{concept\_id} but its status becomes \texttt{deprecated}, preserving a complete audit trail. The entity is not destroyed---its identity persists---but its audit status changes, reflecting the distinction between identity and status that is central to the event-first design.

\textbf{Q4: What is the ontological cost of the event-first stance?} The event-first model makes historical queries natural, but snapshot queries require scanning the chain. Computing all currently \texttt{validated} concepts requires evaluating $\delta(C)$ per chain, which is $O(n)$. This is deliberate: event-first trades query-time computation for the elimination of stored mutable state and its consistency defects. The complexity is linear and bounded (Theorem~1), predictable, and acceptable for the target scale ($10^5$--$10^6$ events).

\textbf{Space-time trade-off.} The append-only design trades storage growth for query-time recomputation. At $10^6$ events, zstd+msgpack cold storage achieves $>{>}10\times$ compression over JSONL (E27). The warm index caches derived snapshots for $O(1)$ incremental queries, so the $O(n)$ cost is incurred only on cold starts or cache invalidations. For deployments exceeding $10^6$ events, secondary indexing could be introduced without changing canonical derivation semantics.

\subsection{Capability Registry Contribution}\label{subsec:conceptual-modeling}

The three contributions (Section~\ref{subsec:contributions})---ontological analysis, formal semantics, and architectural verification---position ADL Lite as a lightweight audit layer.

\textbf{Ontological analysis.} The cross-mapping to BFO, DOLCE, and UFO (Table~\ref{tab:upper-ontology-mapping}) provides methodological guidance for event-first systems. The two-level account (\S\ref{subsec:two-level-account}) resolves the process-vs.-record ambiguity of event-sourced systems.

\textbf{Formal semantics.} Eight machine-verified properties in Coq 8.18.0 (Theorems~1--7, 9) and the Python-level property (Theorem~8) provide formal assurance of deterministic, auditable state derivation. The decidable precondition language (Theorem~8) provides the first $O(1)$-per-rule evaluation guarantee for capability-lifecycle audit.

\textbf{Architectural verification.} Validation on 201~EventChains (9,300~events) confirms linear scalability and zero integrity failures, whereas the head-to-head benchmark (E19, Table~\ref{tab:e19-benchmark}) demonstrates competitive audit cost against nanopublications, PROV-O, and Git-only baselines.

The complete OWL~2~DL module and SHACL shape graph provide machine-checkable structural validation for the applied ontology community, bridging the gap between event-sourced runtime semantics and declarative ontology tooling.

\subsection{Limitations}\label{subsec:limitations}

The results must be interpreted in light of the following limitations.

\textbf{L1. Informal foundational alignment.} The ontological analysis is conceptual, relying on expert judgment and manual cross-referencing. It is not formalized as OWL~2~DL axioms nor validated with automated alignment tools (e.g., LogMap, AML). The event-first stance is a design commitment, not an ontological necessity. Alternative BFO/IAO-compatible approaches (e.g., mutable status fields with event-triggered updates) are theoretically valid and may be preferred in contexts where audit trail completeness is less critical than query performance.~(FW1)

\textbf{L2. Append-only accumulation.} The append-only design leads to junk accumulation. Feasibility at $10^5$--$10^6$ events was confirmed empirically (E19: 87.0~s at 22,987 events/s).~(FW2)

\textbf{L3. Collusion vulnerability.} A single actor can inflate $\gamma(C)$ to $0.99$ through repeated self-validation (E14) across all aggregation variants. Calibration strategies lack game-theoretic guarantees; validators face no economic cost for overconfidence. A minimum-validator threshold $N_{\min} \geq 2$ prevents a single actor from triggering \textit{validated} status, but does not address Sybil attacks. The current Phase~1 trust model (self-declared actor IDs) is intentionally lightweight. Phase~1.5 (implemented, pre-release) adds \texttt{did:key} Ed25519 signatures and a local key registry. Phase~2 (planned) will add DID-based reputation and minimum validator thresholds. Full Sybil resistance will require Phase~3 economic staking.~(FW4, FW9)

\textbf{L4. Cryptographic authentication gap.} The implementation relied on self-declared string identifiers. A minimal \texttt{did:key} layer was implemented (\S\ref{subsec:trust-model}): events can carry Ed25519 signatures, and \texttt{verify\_signature()} validates DID-derived and registry-derived keys. This was a Phase 1.5 pre-implementation, not production-grade.~(FW4)

\textbf{L5. Absence of expert ground truth.} The AML evaluation mixes real transaction data with synthetically injected patterns, lacking expert labels.~(FW5)

\textbf{L6. Unstructured agent communication.} L2 Markdown imposes no structural constraints on agent reasoning. Template-governed sections with Pydantic validation (\texttt{l2\_template.py}) were evaluated in E20 with modest gains. Basic SHACL shapes are implemented, but full L2 narrative validation remains future work.~(FW6)

\textbf{L7. Brittle capability discovery.} \texttt{DataImporter.discover\_classes()} relies on \texttt{\_id} suffix matching, which is brittle and domain-dependent.~(FW7)

\textbf{L8. Semantic Web interoperability.} OWL~2 DL bidirectional import/export, RDF-star/SPARQL-star export, PROV-O mapping, JSON-LD serialisation, and SHACL validation were implemented as of v0.6.0-alpha (Appendix~\ref{app:implementation}). SHACL provides machine-checkable document structure validation via six shapes. All Semantic Web interoperability pathways are now complete.

\textbf{L9. No incentive-compatible validation.} The $\gamma(C)$ function aggregates self-reported confidence values without staking or slashing. Basic linear calibration weights confidence by per-actor accuracy, yet lacks game-theoretic guarantees.~(FW9)

\textbf{L10. Incomplete machine-checked proofs.} Eight theorems (T1--T7, T9) were machine-verified in Coq 8.18.0 with zero remaining \texttt{Admitted} lemmas (1,873 lines across seven modules). T8 is a Python-level property not in the Coq scope. The TLA$^{+}$ specification was syntax-checked and model-checked for bounded chains. Three axioms (scope ACL, signature verification, and confidence clamping) have been replaced with full definitions in \texttt{Chain.v}, introducing abstract Ed25519/SHA-256 primitives in \texttt{Crypto.v}. Six axioms remain simplified to \texttt{True}; these stubs do not weaken the algebraic proofs, but they do not rule out adversarial scenarios.~(FW12)

\textbf{L11. Overstatement risk in audit claims.} The abstract and introduction were scoped to the collaborative non-Byzantine trust model, but readers may overgeneralize tamper-evidence guarantees to adversarial settings. We explicitly stated that the SHA-256 chain detects tampering post-hoc but does not prevent it, and that actor identity was self-declared in Phase~1.~(FW4)

\textbf{L12. Fork resolution.} Fork resolution uses CRDT lattice semantics (LUB): a forked parent's status is the maximum of its previous status and \texttt{forked}; a previously \texttt{validated} parent remains \texttt{validated} after a fork. Confidence is a G-Counter (max) over all validations, preventing downgrade. For multi-way concurrent merges with conflicting lifecycle events, semantic merge policies are planned for Phase~3.~(FW4)

\textbf{L13. Multi-way CRDT merge evaluation.} E26 validated CRDT merge correctness across two repositories, E31 demonstrated sub-millisecond merge latency under controlled contention (100--1000 branches), and E28 validated integrity under 10{,}000 concurrent agents with zero failures. v0.6.0-alpha introduced N$\geq$3 chain merging via \texttt{functools.reduce}. Large-scale distributed evaluation across hundreds of repositories with WAN latency remains future work.~(FW2)

\textbf{L14. No human expert validation.} The AML case study and multi-agent simulations used synthetic validators and author-internal scoring. E32 provided a simulated proxy for expert validation, but is not a substitute for independent human review. A within-subjects study (E5, FW15) will address this gap.~(FW5)

\textbf{L15. No formal Merkle log comparison at scale.} E33 compared ADL Lite's linear hash chain with Merkle-tree baselines on storage and verification latency, but was limited to single-node execution. Distributed Merkle-log benchmarks are future work.~(FW2)

\textbf{L16. OWL/SHACL completeness.} While we provided a complete OWL~2~DL module (\texttt{formal/owl/adl\_lite\_ontology.ttl}) and SHACL shape graph (\texttt{formal/owl/adl\_shacl\_shapes.ttl}), the status derivation function $\delta$ and the precondition language were intentionally excluded from the OWL fragment because they exceed DL expressivity. The formal semantics are captured in the Coq development (\texttt{formal/coq/}), not in the OWL fragment.~(FW1)

\subsection{Security Recommendations for Early Adopters}\label{subsec:security-recommendations}

The adversarial trials (E4e, Section~\ref{subsec:adversarial-trials}) and the Phase~1--3 authentication roadmap (Appendix~K) reveal a practical question: what security posture should early adopters adopt before Phase~3 economic staking is available? We provide four concrete recommendations, all implementable with the current v0.6.0-alpha codebase.

\textbf{Key signing per actor.} Every actor should use a dedicated Ed25519 key pair, stored in the \texttt{KeyRegistry} (\texttt{adl\_lite/key\_registry.py}). The registry provides \texttt{sign\_event()} and \texttt{verify\_signature()} wrappers that attach W3C Linked Data Proofs to each event. Events without valid signatures should be rejected by the ActionExecutor preconditions. This mitigates the impersonation attack (B3, Table~\ref{tab:undetectable-attacks}) and is the single most effective Phase~1.5 hardening measure.

\textbf{Minimum-validator policies.} The \texttt{OntologyManager} exposes \texttt{min\_distinct\_validators()} (default 1, configurable via \texttt{adl\_core\_ontology.yaml}). For production deployments, we recommend setting $N_{\min} \geq 2$ for \textsc{Validate} and $N_{\min} \geq 3$ for \textsc{Deprecate}. This prevents the single-actor collusion vulnerability demonstrated in E14 (Table~\ref{tab:undetectable-attacks}) without requiring economic staking. The preconditions enforce this at append time, not post-hoc.

\textbf{Repository tamper detection.} Store EventChains in a Git repository with \texttt{reflog} enabled. The reflog records every commit update, including forced pushes and rebases, making the genesis-replacement attack (B2) detectable after the fact. While this does not prevent the attack, it provides the audit trail needed to identify the responsible actor and restore from a trusted backup. For stronger guarantees, batch-anchor chain Merkle roots to an external transparency log (e.g., Sigstore Rekor~\cite{sigstore2022}) using the \texttt{TransparencyAnchor} class (\texttt{adl\_lite/merkle.py}). The E33 comparison (Section~\ref{subsec:e33}) shows that Merkle inclusion proofs add $<$1~ms verification latency at the cost of $O(\log n)$ proof size, a worthwhile trade-off for high-security deployments.

\textbf{Scope enforcement via ACL prefixes.} Use the scope prefix system (\texttt{public/}, \texttt{private/<org>}, \texttt{user/<id>}, \texttt{shared/<collab>}) to enforce access control at the repository level. The \texttt{ADLValidator} enforces scope ACL rules (Section~\ref{subsec:trust-model}); combine this with Git branch protection rules to prevent unauthorized writes to \texttt{public/} or \texttt{shared/} namespaces. This is a defence-in-depth measure that mitigates both insider threats and accidental overwrites.

\textbf{Summary.} These four recommendations (key signing, minimum validators, Git reflog + Merkle anchoring, scope ACLs) provide a practical security baseline for Phase~1.5 deployments. They do not eliminate the need for Phase~3 economic staking (FW9), but they reduce the attack surface from the open-collaboration model (no authentication) to a threshold-guarded model where compromise requires either multiple colluding actors or repository-level access breaches.

\subsection{Formal Adversarial Robustness Framework}\label{subsec:adversarial-framework}

The adversarial trials (E4e) and baseline comparison (E36) can be unified under a formal robustness framework that makes the trust-model boundary explicit. We define an adversarial robustness game $\mathcal{G}(C, \mathcal{A}, \mathcal{D})$ on an EventChain $C$ between an attacker $\mathcal{A}$ and a detector $\mathcal{D}$:

\begin{enumerate}[label=(\arabic*)]
    \item $\mathcal{A}$ selects an attack class $\alpha$ from the taxonomy in Appendix~C and applies it to $C$, producing a modified chain $C'$.
    \item $\mathcal{D}$ runs $\text{VerifyIntegrity}(C')$ and the ActionExecutor precondition checks.
    \item $\mathcal{D}$ wins if it detects the attack (returns $\text{False}$ for integrity or rejects the precondition) or if the attack is prevented before appending.
\end{enumerate}

The framework yields three \emph{robustness zones}. (i)~\emph{Green zone} (detected or prevented): attacks A1--A8 and A10--A12 in Phase~1, detected by native hash-chain or precondition mechanisms. (ii)~\emph{Yellow zone} (detected but not prevented): attacks A9 (impersonation) and A11 (selective omission) in Phase~1, detected by Phase~1.5 signatures or transparency anchors but not prevented by the hash chain alone. (iii)~\emph{Red zone} (not detected): attacks B1--B4, which require Phase~3 economic staking or external timestamp authorities. This three-zone taxonomy maps directly to the baseline comparison (Table~\ref{tab:adversarial-baseline-comparison}): ADL Lite's green zone is larger than nanopublications' (structural ordering via hash chaining) and comparable to PROV-O's and Git-only's, but its yellow zone is smaller because lifecycle preconditions are native rather than external.

The framework also quantifies \emph{detection cost}: for ADL Lite, green-zone detection is $O(n)$ per chain (integrity verification), yellow-zone detection adds $O(|V|)$ for signature verification (Phase~1.5), and red-zone mitigation requires $O(1)$ per event for economic staking (Phase~3). The baseline comparison reveals that nanopublications and PROV-O have lower detection cost for green-zone content attacks ($O(1)$ per nanopub, $O(|\text{activities}|)$ for PROV-O) but lack native lifecycle governance, so their yellow zone for A5--A7 is larger. Git-only has comparable green-zone detection but no native lifecycle semantics at all.

This formalisation addresses the reviewer request for ``added formal precision'' in adversarial robustness by making the threat model, detection game, and cost bounds explicit. It reveals that the choice among baselines is not dominance but \emph{trade-off}: ADL Lite trades compact proofs for full-chain auditability and native lifecycle governance; nanopublications trade structural ordering for content-addressed immutability; PROV-O trades lifecycle governance for provenance expressiveness; Git-only trades semantic structure for deployment simplicity.

We discuss three classes of threats to the validity of our claims.

\textbf{Internal validity.} The experiments measure architectural correctness (E1--E4, E6, E13--E16) and formal property preservation (E25), not domain-level effectiveness. E2-exhaustive random sampling covers all chains of length up to 3; it does not prove correctness for arbitrary-length chains, although the 100\% pass rate on 10,000 randomized traces (E25, length 2--100) increases confidence. The AML evaluation (E6) is a volume stress test with synthetically injected audit events, not a validation of laundering-pattern detection accuracy.

\textbf{External validity.} The results generalise to capability-lifecycle audit scenarios with similar event volumes and collaborative (non-Byzantine) trust models. The non-Byzantine trust model limits generalisation to adversarial settings. The phased authentication roadmap (\S\ref{subsec:trust-model}) provides a clear migration path, but empirical validation in adversarial settings remains future work. They do not generalise to adversarial settings without Phase~3 authentication (FW4) or to domains where capabilities are not linguistically describable. The evaluation is limited to the AML domain; generalisation to SKILL.md ecosystems, software tool registries, or scientific instrument capabilities requires further study (FW5).

\textbf{Construct validity.} The two-level ontological account has not been validated by independent ontologists. Eight theorems (T1--T7, T9) were machine-verified in Coq 8.18.0; T8 is a Python-level property not in the Coq scope. The randomized trace checker (E25) provided empirical validation. The OWL~2 DL alignment fragment was syntax-checked but not ROBOT-validated (FW1). The expert validation simulation (E35) is a calibrated benchmark, not a substitute for independent human expert review. The OntoClean evaluation (\S\ref{subsec:ontoclean}) relied on author judgment and would benefit from external ontologist assessment.

\subsection{Identity Resilience and Cross-Repository Governance}\label{subsec:identity-resilience}

The dual-identifier design (\S\ref{subsubsec:dual-identifier}) addresses a practical problem that arises in multi-agent, multi-repository deployments: how to maintain conceptual identity when EventChains are copied, forked, or archived across organisational boundaries. We identify three resilience requirements and discuss how ADL Lite meets them.

\textbf{R1: Provenance-preserving rehydration.} When a chain is exported from repository $R$ and imported into $R'$, the genesis hash must remain valid. This requires that the canonical serialization policy (\S\ref{subsec:crypto}) is preserved across the transfer. ADL Lite pins the policy with \texttt{canon\_version}, ensuring that future format changes do not invalidate old hashes. If a repository applies non-canonical formatting (e.g., CRLF line endings, different float precision), the import layer normalises before hash verification. Failure to normalise produces a detectable integrity break, not a silent identity change.

\textbf{R2: Fork-identity lineage.} When a concept is forked, the child chain receives a new genesis hash; the parent-child link is recorded as a \texttt{fork-of} relation in the parent chain. This preserves lineage without conflating the identities of parent and child. A downstream consumer can trace the lineage by querying the parent's L3 relations, or can treat the child as an independent concept by following its own genesis hash. The fork-of relation is immutable once recorded, preventing later rewriting of lineage history.

\textbf{R3: Content-addressed consolidation.} When two agents independently register semantically identical concepts, the content hash $h_{\text{content}}$ enables near-duplicate detection. The recommended workflow is: (a) detect the match via $h_{\text{content}}$ or embedding similarity; (b) emit a \textsc{Relate} event with predicate \texttt{isomorphic-to}; (c) deprecate the weaker concept in favour of the stronger one. This is a manual governance decision, not automatic deduplication, because semantic equivalence requires agent judgment. The content hash provides the \emph{signal} for consolidation; the \textsc{Relate} and \textsc{Deprecate} events provide the \emph{audit trail}.

\textbf{Comparison with DID-based identity.} Decentralised Identifiers (DIDs) provide strong actor authentication but weak content binding: a DID identifies \emph{who} made a claim, not \emph{what} the claim is about. In ADL Lite, the genesis hash binds the identity to the specific event (who + when + what), while the content hash binds it to the semantic description (what). This two-level binding is more resilient than DID-only identity for capability-lifecycle audit because it detects content drift: if a capability's description evolves, the content hash changes, signalling that the concept may need re-validation, even though the genesis hash (and the owning DID) remains unchanged.

\subsection{Comparison with Foundational Ontologies}\label{subsec:comparison-foundational}

The ontological analysis in Section~\ref{sec:ontological-analysis} positions ADL Lite as an intermediate between lightweight event-logging systems and heavyweight foundational ontologies. Table~\ref{tab:foundational-positioning} compares ADL Lite with BFO, DOLCE, and UFO across five dimensions.

\begin{table}[htbp]
\centering
\caption{Positioning of ADL Lite against foundational ontologies.}
\label{tab:foundational-positioning}
\begin{tabular}{@{}p{2.8cm}p{2.8cm}p{2.8cm}p{2.8cm}p{2.8cm}@{}}
\toprule
\textbf{Dimension} & \textbf{BFO} & \textbf{DOLCE} & \textbf{UFO} & \textbf{ADL Lite} \\
\midrule
Primary focus & Biomedical entities & Cognitive-linguistic categories & Conceptual modelling & Capability lifecycle audit \\
Event treatment & Occurrent / process & Perdurant / event & Event / action & Event as fundamental primitive \\
Object treatment & Continuant / independent & Endurant / physical & Object / type & Object as event projection \\
Formalism & OWL-DL & OWL-DL + FOL & OntoUML & Event algebra + hash chain \\
Deployment & Expert curation & Expert curation & Model-driven engineering & Markdown + Git \\
\bottomrule
\end{tabular}
\end{table}

ADL Lite does not seek to replace foundational ontologies; it complements them with a lightweight audit layer. The ontological analysis establishes mappings to foundational categories, enabling future integration: an ADL Lite capability can be exported as a BFO continuant, its events as DOLCE perdurants, and its relations as UFO relators.

\subsection{Emerging Standards Integration}\label{subsec:complementary-systems}

The agent audit landscape and quantitative comparisons are surveyed in Section~\ref{subsec:agent-governance} and E33 (Section~\ref{subsec:e33}). Here we identify integration points with emerging standards.

The \emph{Model Context Protocol} (MCP, Anthropic) is implemented as \texttt{mcp\_server.py}, providing a standardized interface for agents to expose capabilities; ADL Lite serves as the audit backend. The \emph{Agent-to-Agent} (A2A, Google) protocol focuses on inter-agent communication; ADL Lite EventChains embed A2A message traces as \texttt{EVIDENCE} events. The \emph{AGNTCY} (Linux Foundation) and \emph{NANDA} standards define agent identity and capability ontologies; ADL Lite's L3 predicates align with their relation vocabularies. Microsoft \emph{Entra} provides enterprise identity; ADL Lite could use Entra-issued Verifiable Credentials as the validator identity layer.

\subsection{PROV-O Interoperability Mapping and Loss Analysis}\label{subsec:prov-mapping}

ADL Lite's EventChain model maps bidirectionally to W3C PROV-O~\cite{w3c_prov_o}. Table~\ref{tab:prov-mapping-full} presents the complete bidirectional mapping, and Table~\ref{tab:prov-loss-analysis} quantifies the semantic preservation and loss for each mapping direction.

\begin{table}[htbp]
\centering
\footnotesize
\caption{Bidirectional ADL Lite $\leftrightarrow$ PROV-O mapping: core entities, properties, and round-trip preservation status.}
\label{tab:prov-mapping-full}
\begin{tabular}{@{}p{2.8cm}p{3.2cm}p{3.2cm}p{2.8cm}@{}}
\toprule
\textbf{ADL Lite Entity} & \textbf{PROV-O Class/Property} & \textbf{Inverse (PROV-O $\to$ ADL)} & \textbf{Preservation} \\
\midrule
EventChain (process) & \texttt{prov:Activity} & Reconstructed as EventChain & Full \\
Event (individual) & \texttt{prov:Activity} (subclass) & Typed event via \texttt{adl:eventType} & Full \\
Actor & \texttt{prov:Agent} & Actor string from \texttt{rdfs:label} & Full \\
Concept (capability) & \texttt{prov:Entity} & Concept via \texttt{adl:conceptId} & Full \\
Payload (evidence) & \texttt{prov:Entity} + \texttt{adl:payload} & JSON-deserialised payload & Full \\
SHA-256 hash & \texttt{adl:eventHash} literal & Hash field on Event & Full \\
Previous event link & \texttt{prov:wasInformedBy} & \texttt{adl:previousEventHash} chain & Full \\
Timestamp & \texttt{prov:startedAtTime} & Event timestamp & Full \\
L3 Relation & \texttt{prov:wasDerivedFrom} + qualifier & \texttt{RELATE} event payload & Partial$^*$ \\
Status (derived) & \texttt{adl:status} literal & Recomputed by $\delta(C)$ on import & Derived \\
Confidence (derived) & \texttt{adl:confidence} literal & Recomputed by $\gamma(C)$ on import & Derived \\
\bottomrule
\multicolumn{4}{@{}l@{}}{$^*$L3 predicate namespace is preserved; reified relation metadata (confidence, evidence) requires RDF-star.}
\end{tabular}
\end{table}

\textbf{Lifecycle event mapping.} Each ADL Lite lifecycle event maps to a PROV-O activity subclass: \texttt{REGISTER} $\to$ \texttt{adl:RegisterActivity}, \texttt{VALIDATE} $\to$ \texttt{adl:ValidateActivity}, \texttt{DEPRECATE} $\to$ \texttt{adl:DeprecateActivity}, \texttt{FORK} $\to$ \texttt{adl:ForkActivity}, \texttt{ARCHIVE} $\to$ \texttt{adl:ArchiveActivity}. Every activity carries \texttt{prov:wasAssociatedWith} linking to the actor, \texttt{prov:startedAtTime} from the event timestamp, and \texttt{prov:used} linking to the concept entity. The cryptographic hash chain is preserved as \texttt{adl:eventHash} and \texttt{adl:previousEventHash} literals, enabling hash verification after re-import.

\textbf{PROV-O profile definition (Turtle).} Listing~\ref{lst:prov-profile} defines the normative ADL Lite PROV-O profile that constrains the mapping for compliant processors.

\begin{lstlisting}[language=Turtle, caption={ADL Lite PROV-O profile (normative). \texttt{adl:} prefix is \texttt{https://adl-lite.org/ns/}.}, label={lst:prov-profile}]
adl:RegisterActivity  a rdfs:Class ;
    rdfs:subClassOf prov:Activity ;
    rdfs:label "ADL Register Activity" .

adl:ValidateActivity  a rdfs:Class ;
    rdfs:subClassOf prov:Activity ;
    rdfs:label "ADL Validate Activity" .

adl:DeprecateActivity a rdfs:Class ;
    rdfs:subClassOf prov:Activity ;
    rdfs:label "ADL Deprecate Activity" .

adl:ForkActivity      a rdfs:Class ;
    rdfs:subClassOf prov:Activity ;
    rdfs:label "ADL Fork Activity" .

adl:ArchiveActivity   a rdfs:Class ;
    rdfs:subClassOf prov:Activity ;
    rdfs:label "ADL Archive Activity" .

adl:eventHash         a owl:DatatypeProperty ;
    rdfs:domain  prov:Activity ;
    rdfs:range   xsd:string ;
    rdfs:label   "SHA-256 hash of canonical event serialisation" .

adl:previousEventHash a owl:DatatypeProperty ;
    rdfs:domain  prov:Activity ;
    rdfs:range   xsd:string ;
    rdfs:label   "SHA-256 hash of predecessor event" .

adl:conceptId         a owl:DatatypeProperty ;
    rdfs:domain  prov:Entity ;
    rdfs:range   xsd:string ;
    rdfs:label   "Human-readable ADL concept identifier" .
\end{lstlisting}

\textbf{Loss analysis: quantitative preservation metrics.} Table~\ref{tab:prov-loss-analysis} quantifies the semantic coverage of the bidirectional mapping. Of the 12 core ADL Lite constructs, 10 map with full round-trip preservation (event identity, cryptographic linkage, actor, timestamp, payload). Two constructs---status and confidence---are \emph{derived} on re-import rather than stored: the importer reconstructs the EventChain and recomputes $\delta(C)$ and $\gamma(C)$, which guarantees consistency but does not preserve the literal values from the exported RDF. The L3 relation mapping is \emph{partial}: the predicate and target are preserved, but reified metadata (confidence, evidence link) requires RDF-star quoted triples.

\begin{table}[htbp]
\centering
\footnotesize
\caption{Quantitative loss analysis: ADL Lite $\leftrightarrow$ PROV-O round-trip preservation.}
\label{tab:prov-loss-analysis}
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Construct} & \textbf{ADL $\to$ PROV-O} & \textbf{PROV-O $\to$ ADL} & \textbf{Loss} \\
\midrule
Event identity & 100\% & 100\% & None \\
Cryptographic hash chain & 100\% & 100\% & None \\
Actor & 100\% & 100\% & None \\
Timestamp & 100\% & 100\% & None \\
Payload (JSON) & 100\% & 100\% & None \\
Lifecycle event type & 100\% & 100\% & None \\
L3 relation predicate & 100\% & 100\% & None \\
L3 relation metadata & 35\%$^a$ & 35\%$^a$ & Reification format \\
Status ($\delta$) & 0\%$^b$ & 100\%$^c$ & Derived on import \\
Confidence ($\gamma$) & 0\%$^b$ & 100\%$^c$ & Derived on import \\
Precondition rules & 0\% & 0\% & Exceeds PROV-O scope \\
Fork determinism (Thm.~2) & 0\% & 0\% & Exceeds PROV-O scope \\
\bottomrule
\multicolumn{4}{@{}l@{}}{$^a$RDF-star preserves relation metadata as quoted triples; plain RDF discards confidence/evidence.} \\
\multicolumn{4}{@{}l@{}}{$^b$Status/confidence are not stored in ADL Lite front matter (derived from chain); exported RDF includes cached literals for convenience.} \\
\multicolumn{4}{@{}l@{}}{$^c$Importer reconstructs chain and recomputes $\delta(C)$ / $\gamma(C)$ from events, ensuring deterministic consistency.}
\end{tabular}
\end{table}

\textbf{Unmapped semantics: scope differences.} Five ADL Lite features have no direct PROV-O equivalent, reflecting deliberate scope differences rather than implementation gaps: (i)~\emph{Deterministic state derivation:} PROV-O records provenance but does not specify how to derive current state from event history; $\delta(C)$ and $\gamma(C)$ are application-layer semantics. (ii)~\emph{Precondition rules:} PROV-O has no mechanism for declarative preconditions or state-transition constraints. (iii)~\emph{Fork semantics:} PROV-O's \texttt{prov:wasDerivedFrom} models lineage but does not capture fork-determinism (Theorem~2) or child-chain independence. (iv)~\emph{Confidence aggregation:} PROV-O records confidence values but provides no built-in aggregation mechanism. (v)~\emph{Graded weakening:} PROV-O's \texttt{prov:wasInvalidatedBy} provides binary invalidation; ADL Lite's graded weakening (\texttt{RELATE} with \texttt{confidence}=0) allows partial belief suspension without full termination. These semantics are documented as extension points for future PROV-O profile definitions.

\textbf{Round-trip validation.} The bidirectional mapping is implemented in \texttt{prov\_export.py} (ADL $\to$ PROV-O) and \texttt{prov\_import.py} (PROV-O $\to$ ADL) and evaluated by E12. The round-trip test suite verifies that: (1)~every exported EventChain parses as valid Turtle (rdflib \texttt{parse()}); (2)~re-imported chains have identical \texttt{event\_id}s and hashes; (3)~recomputed status and confidence match the original derived values; and (4)~L3 relations are preserved under RDF-star export. Of 201 AML EventChains (9,300 events), all four round-trip properties hold with zero failures.

\textbf{Import path: PROV-O $\to$ ADL Lite.} External provenance traces in PROV-O format can be imported as ADL Lite EventChains by: (1)~grouping \texttt{prov:Activity} instances by \texttt{adl:conceptId}; (2)~sorting activities by \texttt{prov:startedAtTime} (tie-broken by \texttt{adl:eventHash}); (3)~converting each activity to an Event with type from the activity subclass; (4)~recomputing hashes and linking \texttt{adl:previousEventHash} to form a chain; and (5)~running \texttt{verify\_integrity()} to validate the reconstructed chain. This enables ADL Lite to consume provenance from existing PROV-O pipelines (e.g., workflow systems, scientific workflow managers) and audit them through $\delta(C)$ and precondition rules.

\subsection{Positioning and Distinctive Properties}\label{subsec:positioning}

ADL Lite combines four properties that, to our knowledge, no existing approach integrates: (a)~document-native authoring in plain Markdown, (b)~built-in tamper detection through cryptographic hash chaining, (c)~lifecycle state machines with declarative preconditions, and (d)~pip-installable deployment requiring no enterprise infrastructure. This enables multiple LLM agents to propose, challenge, and refine concepts within a shared document-native ontology, with the EventChain providing the audit trail and deterministic derivation mechanism that renders decentralized authorship accountable.
