\subsection{Summary of Contributions}\label{subsec:summary}

This paper presented ADL Lite, an event-first capability-lifecycle registry for LLM agent ecosystems, and evaluated it through ontological analysis, formal semantics, and empirical validation. The contributions are restated below in capability-registry language:

\textbf{Ontological analysis of event-first capability governance.} We analyzed ADL Lite's core categories through BFO, DOLCE, and UFO, providing a two-level account that distinguishes occurrent events from their continuant records, with explicit identity and dependence axioms. We demonstrated that Event corresponds to occurrent/perdurant, Concept to generically dependent continuant/non-physical object, Relation to relational quality/relator, and Action to planned process/intentional event. We formalized identity conditions, relation lifecycle semantics, and analyzed three forms of ontological dependence.

\textbf{Formal derivation semantics with proved properties.} We provided a complete formal foundation: event alphabet, well-formedness predicate, status derivation $\delta(C)$, confidence aggregation $\gamma(C)$, fork/confluence semantics, and seven proved properties. We formalized a precondition language with explicit comparator semantics and proved decidability of precondition evaluation. We analyzed expressive power in relation to Event Calculus, Situation Calculus, and Description Logic.

\textbf{Empirical validation of the capability registry architecture.} Expanded experiments confirmed zero integrity failures, 100\% derivation accuracy (2,204/2,204 cases), perfect precondition enforcement across 141 cases (P = R = F1 = 1.0), linear scalability (20{,}955 events/second), multi-agent concurrency under contention, adversarial resilience, scale-up projection to $10^5$ events, and comparative baselines against nanopublications and PROV-O pipelines.

ADL Lite is not a competing ontology production method; it is a complementary governance layer that records, validates, and governs the lifecycle of capabilities produced by any agent pipeline.

\subsection{Future Work}\label{subsec:future-work}

Future work is organized into five thematic clusters that address the limitations identified in Section~\ref{subsec:limitations}.

\subsubsection{Formal Foundations and Ontological Alignment}

\textbf{FW1. Ontological Alignment with Foundational Ontologies and OBO Foundry.}
Future work includes formal OWL 2 DL axiomatizations of ADL Lite's four core categories (Event, Concept, Relation, Action) aligned with BFO 2.0, DOLCE, and UFO modules, and a bidirectional ADL Lite $\leftrightarrow$ OWL interoperability bridge enabling round-trip translation between event chains and OWL 2 DL ontologies. Cross-validation with ontology alignment tools---specifically LogMapLite (which processes large biomedical ontologies in under 2 seconds with F-measure $\geq$ 0.85) and Agent-OM (an LLM-based aligner achieving F-measure 0.92 at OAEI 2025)---will quantify alignment precision. Participation in the OBO Foundry interoperability framework via ROBOT consistency checking is also planned. This work builds directly on the ontological analysis in Section~\ref{sec:ontological-analysis}.

\textbf{FW10. Formal Verification of EventChain Correctness in TLA$^+$ or Coq.}
The seven theorems in Section~\ref{subsec:formal-semantics} are proved informally (rigorous but not machine-checked). Future work provides machine-verifiable proofs: either in TLA$^+$ (specifying EventChain as a state variable with AppendEvent as an action, verifying the seven theorems as temporal properties) or in Coq using Iris separation logic (following Nieto's CRDT formalization \cite{nieto2024crdt}). LLM-assisted formalization tools will accelerate the translation from natural-language theorems to machine-checked proofs. Reproducible verification scripts will be provided as supplementary material.

\textbf{FW12. Machine-Checked Formal Verification of Theorems and Ontological Axioms.}
Future work translates the seven theorems and the two-level ontological axioms into Coq or TLA$^+$ for mechanized proof. This complements FW10 (EventChain correctness) by covering the ontological foundation and precondition decidability results.

\subsubsection{Authentication, Security, and Scale}

\textbf{FW4. Cryptographic Authentication: From Hash Chains to Verifiable Data Structures.}
Phase~3 extends actor authentication from string identifiers to cryptographically verified identities. The planned integration uses Linked Data Proofs (URDNA2015 canonicalization + ed25519 signatures) \cite{linked_data_proofs} and decentralized identifiers (DIDs). Beyond signatures, the linear SHA-256 chain will be upgraded to a Merkle tree structure supporting $O(\log n)$ batch verification \cite{crosby_wallach}. Verkle trees (vector commitment with $O(\log_k n)$ proof size) offer further optimization for large-scale deployments. Equivocation prevention will be enforced via cryptographic attribution, binding every action to a verifiable actor identity. Byzantine fault tolerance will draw on the Blocklace \cite{blocklace2024} hash DAG's equivocation-detection mechanism.

\textbf{FW11. Large-Scale Empirical Validation under Contention and Comparative Benchmarking.}
Future work includes empirical measurement at $10^5$--$10^6$ events, multi-agent concurrency under contention, and head-to-head benchmark against nanopublications and PROV-O pipelines with comparable authentication layers. This extends the current scale-up projection (E6b) and comparative baselines (E12) to full empirical validation.

\subsubsection{Confidence Calibration and Incentive Design}

\textbf{FW3. Runtime Confidence Calibration via MARGIN and Epistemic Context Learning.}
The current $\gamma(C)$ function aggregates self-reported confidence values without calibration; empirical studies have demonstrated that such values are systematically overconfident. Future work integrates symmetric EWMA calibration mechanisms, which achieve 3--6$\times$ lower Expected Calibration Error (ECE) under distribution shift by learning per-actor, per-confidence-band calibration factors. Epistemic context learning provides a complementary approach using reinforcement learning to optimize peer reliability estimates. Collaborative entropy offers formal uncertainty decomposition for multi-validator settings. The calibrated $\gamma_{\text{cal}}(C)$ will maintain Theorem~\ref{thm:boundedness} (boundedness) and Theorem~\ref{thm:consistency} (status--confidence consistency) by construction. This addresses the collusion vulnerability demonstrated in E14 (Section~\ref{subsec:e14}), where a single actor can drive $\gamma(C)$ to 0.99.

\textbf{FW9. Epistemic Market: Staking-Based Concept Validation.}
The current $\gamma(C)$ aggregates self-reported confidence without an incentive-compatible mechanism for truthful reporting. Future work designs an ``Ontological Assertion Market'' in which validators stake tokens on concept assertions, with correct assertions rewarded and incorrect assertions slashed. MARGIN calibration factors serve as staking weights: higher-calibration validators receive proportionally higher influence. A formal proof that the market equilibrium converges to true concept quality under rational-agent assumptions is targeted. The staking mechanism also provides collusion prevention: coordinated manipulation requires capital lock-up, and slashing creates a negative-sum penalty for dishonest coalitions.

\subsubsection{Domain Evaluation and Agent Communication}

\textbf{FW5. Domain Evaluation: From Architectural Correctness to Real-World Capability Governance Effectiveness.}
E1--E3, E4 establish architectural correctness; domain effectiveness requires expert-annotated ground truth. The planned evaluation includes the IBM AML HI-Small dataset (where 96.8\% of E6 events were REGISTER imports from transaction data, and 3.2\% were synthetic governance events) and a pilot LLM-agent capability evaluation (E17). A multi-agent coordination efficiency study (E6M) has been designed and preliminarily simulated ($k \in \{3, 5, 10\}$ agents, $n=50$ concepts, 100 trials per condition); full results will be reported alongside the domain evaluation. Comparison against GNN-based AML baselines will quantify ADL Lite's explainability advantage.

\textbf{FW6. Structured Agent Communication: From Free-Text Markdown to Template-Governed L2.}
The current L2 Markdown body imposes no structural constraints on agent reasoning. Future work defines a template-governed L2 with mandatory sections [Observation], [Reasoning], and [Conclusion], enforced by SHACL shape validation at the syntax level (not merely semantics). Template compliance checking will reject improperly structured events before append. The LLM4ACOE ReAct guidelines (thought-action-observation trajectories) \cite{llm4acoe} provide a proven template pattern. Empirical comparison of free-text vs.\ structured templates on concept quality (expert scores) and conflict rates will quantify the benefit.

\textbf{FW7. LLM-Guided Concept Discovery: Beyond \texttt{\_id} Suffix Matching.}
The current DataImporter uses \texttt{\_id} suffix matching for class discovery, which is brittle and coverage-limited. Future work integrates LLMs (GPT-4o class) as concept discovery agents: input raw data samples, output candidate concept definitions (L1 YAML + L2 Markdown drafts), with RAG injecting domain knowledge (e.g., FATF AML typologies) \cite{llm4acoe}. A human-in-the-loop review workflow (LLM proposes $\rightarrow$ human audits $\rightarrow$ EventChain registers) preserves governance while expanding coverage. Comparative evaluation against \texttt{\_id} suffix matching on concept coverage and accuracy metrics is planned.

\subsubsection{Interoperability and Storage}

\textbf{FW2. Controlled Forgetting: From Append-Only to Adaptive Lifecycle Compression.}
The append-only model requires extensions for long-term deployment. Three mechanisms are planned: (1) an \textsc{ARCHIVE} event type that migrates historical events to cold storage while preserving hash-link integrity; (2) embedding-based semantic deduplication using sentence-BERT cosine similarity to compute event redundancy scores and generate \textsc{DEDUP} events that record merge mappings; (3) adaptive quality gates that score information gain before appending events, inspired by adaptive summarization frameworks and temporal knowledge-graph archival strategies. Together, these mechanisms address the ``append-only accumulation'' limitation (L2) without sacrificing cryptographic integrity.

\textbf{FW8. RDF-star / SPARQL-star Interoperability: Bridging ADL Lite and the Semantic Web.}
RDF-star \cite{rdf_star} provides statement-level annotation with compact syntax (\texttt{<{}<s p o>{}> :certainty 0.9}), and GraphDB benchmarks demonstrate 34\% faster loading and 39\% less storage than standard reification \cite{graphdb_rdfstar}. Future work defines a complete ADL Lite $\leftrightarrow$ RDF-star bidirectional converter: L3 relation blocks map to \texttt{<{}<concept\_id relation target\_id>{}> :validator agent\_001 :confidence 0.95}. SPARQL-star queries will enable retrospective EventChain analysis using standard Semantic Web tooling (GraphDB, RDF4J, Apache Jena).

\subsection{Limitations and Response to Review}

This revision addresses the reviewer's concerns through strengthened formalization, expanded empirical coverage, and explicit scoping of claims. Some limitations remain: comparative baselines require Phase~3 authentication parity; scale-up beyond $10^4$ events is projected but not empirically validated; and the ontological axioms, while explicit, are not yet machine-checked. These are scoped to future work. The comparative baseline against nanopublications and PROV-O remains methodological rather than experimental, as Phase~1 lacks the authentication parity required for head-to-head benchmarking. The honest framing of E6 as an architectural stress test (96.8\% REGISTER events, 3.2\% synthetic governance) is now explicit.
