LLM agents are proliferating across enterprise and open ecosystems, each exposing tools, APIs, and knowledge-domain capabilities. Yet no lightweight, tamper-evident, lifecycle-aware registry exists to govern what a capability claims, how it evolves, or whether it remains trustworthy. Existing governance layers---KYA for permissions and AgentSafe for architecture-level oversight---do not address this gap. ADL Lite addresses this gap as an event-first capability-lifecycle registry.

Events are the fundamental architectural primitive, distinguished from their immutable records---EventChains. Capabilities are modeled as append-only, cryptographically linked EventChains, with all lifecycle state (status, confidence, and validators) derived exclusively from event history through deterministic functions. Every capability claim, validation, deprecation, and fork is an auditable event, eliminating stored mutable state and the consistency defects it introduces.

We present a two-level ontological analysis that situates ADL Lite within foundational ontologies (BFO, DOLCE, and UFO), formalizes identity conditions and three forms of ontological dependence, and defines a decidable precondition language for lifecycle transitions. We further present a complete formal semantics, together with seven core proven properties (including determinism, fork determinism, confidence boundedness, and well-formedness preservation) and an optional CRDT convergence property (Theorem~9) scoped to a future distributed-merge strategy.

We scope all governance claims in this paper to the collaborative non-Byzantine trust model: honest-but-curious agents who detect inconsistencies, not Byzantine agents who actively deceive. The SHA-256 chain provides tamper-evidence, not tamper-proofing or non-repudiation.

We verified the architecture of the registry on 201 EventChains derived from the IBM Anti-Money Laundering dataset (9,300 events), recording zero integrity failures and confirming linear-time verification under architectural stress tests. We supplemented these findings with concurrency stress tests, adversarial integrity trials, and comparative baselines against nanopublications and Git-native systems. This paper establishes the **ontological and formal foundations** of capability-lifecycle governance. Domain-level evaluation with human experts is deferred to future work. We implement the registry as a pip-installable Python toolkit operating on Markdown files in standard Git repositories, targeting lightweight deployment scenarios where full OWL/SPARQL infrastructure is unavailable or unnecessary.
