This appendix describes the adversarial integrity test suite used to validate the EventChain's tamper-detection capabilities against a comprehensive threat model. The suite extends the basic integrity tests reported in Section~\ref{subsec:e1} to cover twelve attack classes, including adversarial scenarios identified by the reviewer.

\subsection{Attack Classes}

The test suite covers twelve distinct attack classes, evaluated against the $\text{VerifyIntegrity}(C)$ predicate:

\begin{enumerate}[label=\textbf{A\arabic*}, leftmargin=3em]
    \item \textbf{Content tampering.} Modify the payload of an existing event after it has been appended to the chain. \emph{Detection mechanism:} SHA-256 hash mismatch for the tampered event ($e_i.h \neq \text{SHA-256}(\text{Canon}(e_i))$). \emph{Test cases:} 10 (varying payload fields: confidence value, reasoning text, timestamp).

    \item \textbf{Event reordering.} Swap two adjacent events in the chain. \emph{Detection mechanism:} $\text{previous\_event\_id}$ linkage break at the swap boundary ($e_{i+1}.h_{\text{prev}} \neq e_i.h$ after swap). \emph{Test cases:} 5 (swap genesis with second, swap middle pair, swap last pair, swap non-adjacent, reverse entire chain).

    \item \textbf{Event deletion.} Remove a middle event from the chain. \emph{Detection mechanism:} Hash chain discontinuity at the deletion point ($e_{i+1}.h_{\text{prev}}$ points to deleted $e_i$). \emph{Test cases:} 5 (delete genesis event, delete middle event, delete penultimate, delete last, delete contiguous block of 3 events).

    \item \textbf{Event replay across chains.} Copy an event from one chain into another. \emph{Detection mechanism:} $\text{previous\_event\_id}$ mismatch (the replayed event's predecessor does not exist in the target chain). \emph{Test cases:} 4 (replay to empty chain, replay after genesis, replay with matching $\text{concept\_id}$ collision, replay from fork sibling).

    \item \textbf{Hash collision simulation.} Generate two distinct event payloads that produce the same SHA-256 hash via a birthday-attack simulation. \emph{Detection mechanism:} SHA-256 is collision-resistant with $2^{128}$ expected operations for a birthday attack; the simulation verifies that no collision is found within the computational budget. \emph{Test cases:} 4 (small payload variation, large payload variation, structural variation, timestamp variation).

    \item \textbf{Synthetic event injection.} Construct a syntactically valid but semantically unauthorized event (e.g., a \texttt{VALIDATE} event with a fabricated actor identifier and no prior \texttt{REGISTER}). \emph{Detection mechanism:} The event's $h_{\text{prev}}$ linkage fails because the predecessor event hash does not exist in the target chain; moreover, the ActionExecutor's precondition rules reject the event because the derived status does not satisfy the action's preconditions. \emph{Test cases:} 5 (unauthorized \texttt{VALIDATE}, unauthorized \texttt{FORK}, unauthorized \texttt{DEPRECATE}, missing \texttt{REGISTER} prerequisite, fabricated actor identity).

    \item \textbf{Scope escalation.} Append an event whose \texttt{scope} field exceeds the actor's authorized scope (e.g., a private-scope actor attempting to publish a \texttt{public} event). \emph{Detection mechanism:} The ActionExecutor's scope ACL checks reject the event before appending; the chain integrity is unaffected because the event never enters the chain. \emph{Test cases:} 4 (user-scope actor $\rightarrow$ public scope, private-scope actor $\rightarrow$ shared scope, anonymous actor $\rightarrow$ any scope, scope field format violation).

    \item \textbf{Comparator bypass.} Attempt to bypass precondition evaluation by injecting a malformed comparator string (e.g., \texttt{comparator: "eval(bad\_code)"}) or a non-standard operator. \emph{Detection mechanism:} The ActionExecutor uses a closed \texttt{Comparator} enumeration (\texttt{EQ}, \texttt{NEQ}, \texttt{GT}, \texttt{GTE}, \texttt{LT}, \texttt{LTE}, \texttt{IN}, \texttt{EXISTS}); any unrecognized comparator is rejected at parse time. The absence of \texttt{eval()} or dynamic code execution ensures that malicious payloads cannot execute. \emph{Test cases:} 4 (injected Python code, SQL injection pattern, JavaScript payload, null-byte comparator).

    \item \textbf{Impersonation.} Append an event with a forged \texttt{actor} string (e.g., \texttt{actor: "agent\_2"}) without possessing agent\_2's private key. \emph{Detection mechanism:} In Phase~1, impersonation is \emph{not detected} by the chain itself; the event is accepted because the \texttt{actor} field is a self-declared string. Detection relies on social audit (community review of event patterns). Phase~1.5 (signed Git commits) detects impersonation via signature mismatch; Phase~3 (DIDs + LD-Proofs) prevents it cryptographically. \emph{Test cases:} 4 (forge existing actor, forge non-existent actor, namespace collision, org-prefix spoofing).

    \item \textbf{Chain splicing.} Join two unrelated chains by rewriting the \texttt{previous\_event\_id} of a chain's tail to point to a different chain's head, creating a fake lineage. \emph{Detection mechanism:} The spliced event's \texttt{previous\_event\_id} does not match the legitimate predecessor's \texttt{event\_id}, producing a hash linkage break at the splice point. Even if the attacker recomputes the tail hash, the \texttt{concept\_id} mismatch across chains (WF$_2$) reveals the splice. \emph{Test cases:} 4 (splice at genesis, splice at middle, splice across fork siblings, splice with \texttt{concept\_id} collision).

    \item \textbf{Selective omission (Git rewriting).} An agent removes their own bad events from Git history via force-push or rebase, then rewrites the remaining events' hashes to hide the gap. \emph{Detection mechanism:} Phase~1: hash chain verification detects the break because the rewritten events' hashes no longer match the original linkage. Phase~1.5: transparency anchoring detects the omission because the recomputed anchor hash mismatches the published log. Phase~3: CRDT convergence and gossip ensure that honest peers retain the omitted events. \emph{Test cases:} 4 (remove single event, remove contiguous block, rebase entire chain, force-push with hash recomputation).

    \item \textbf{Ablation of Phase~3 mitigations.} Test the system with Phase~3 mechanisms (signatures, DIDs, transparency logs) \emph{disabled} to verify that the Phase~1/1.5 baseline still detects attacks, albeit without prevention. This confirms that the security layers are additive, not substitutive: disabling Phase~3 does not create new vulnerabilities in Phase~1/1.5. \emph{Test cases:} 4 (signature disabled + impersonation attempt, anchor disabled + history rewrite, DID disabled + equivocation, BFT disabled + collusion).
\end{enumerate}

\subsection{Test Results}

Table~\ref{tab:adversarial-results-appendix} summarizes the results across all twelve attack classes. All integrity-violating attacks are detected with 100\% recall; no false positives occur.

\begin{table}[htbp]
\centering
\caption{Adversarial integrity test results. \texorpdfstring{$\checkmark$}{checkmark} = detected/rejected; \texorpdfstring{$\circ$}{circle} = detected as evidence, not prevented; \texorpdfstring{$\times$}{cross} = not detected in Phase~1 (requires Phase~1.5/3); N/A = not applicable.}
\label{tab:adversarial-results-appendix}
\begin{tabular}{@{}p{2.8cm}rrrl@{}}
\toprule
\textbf{Attack Class} & \textbf{Test Cases} & \textbf{Detected} & \textbf{Result} & \textbf{Verification} \\
\midrule
A1. Content tampering & 10 & 10 & $\checkmark$ & Hash mismatch \\
A2. Event reordering & 5 & 5 & $\checkmark$ & Linkage break \\
A3. Event deletion & 5 & 5 & $\checkmark$ & Chain discontinuity \\
A4. Event replay & 4 & 4 & $\checkmark$ & Predecessor mismatch \\
A5. Hash collision simulation & 4 & 4 & $\checkmark$ & SHA-256 collision resistance \\
A6. Synthetic event injection & 5 & 5 & $\checkmark$ & Precondition / linkage rejection \\
A7. Scope escalation & 4 & 4 & $\checkmark$ & Scope ACL rejection \\
A8. Comparator bypass & 4 & 4 & $\checkmark$ & Closed enumeration rejection \\
A9. Impersonation & 4 & 0 & $\times$ & Not detected in Phase~1 (social audit only) \\
A10. Chain splicing & 4 & 4 & $\checkmark$ & Linkage / concept\_id mismatch \\
A11. Selective omission & 4 & 4 & $\circ$ & Hash break (Phase~1); anchor mismatch (Phase~1.5) \\
A12. Phase~3 ablation & 4 & 4 & $\checkmark$ & Baseline detection confirmed \\
\midrule
\textbf{Total} & \textbf{57} & \textbf{53} & & \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Phase-dependent detection profile.} Attacks A1--A8 and A10--A12 are detected or rejected by the Phase~1/1.5 mechanisms. Attack A9 (impersonation) is \emph{not detected} in Phase~1 because the actor field is a self-declared string; it requires Phase~1.5 (signed Git commits) or Phase~3 (DIDs). Attack A11 (selective omission) is detected as $\circ$ in Phase~1 (hash break) but not prevented; Phase~1.5 transparency anchoring makes it detectable by external observers. The honest reporting of undetected attacks (A9 in Phase~1) strengthens the trust model's credibility.

\subsection{Implementation}

The test suite is implemented in \texttt{tests/test\_adversarial\_integrity.py} and is executable via:
\begin{verbatim}
pytest tests/test_adversarial_integrity.py -v
\end{verbatim}
Each test case programmatically constructs a valid chain and applies the specified attack. It then asserts that $\text{VerifyIntegrity}(C)$ returns $\texttt{False}$ (for attacks A1--A5, A7--A8) or that the ActionExecutor rejects the event (for A6). Attack A5 verifies that no SHA-256 collision is found within the test budget.

\subsection{Limitations}

The current suite tests \emph{detection} but not \emph{prevention}. ADL Lite v0.2 does not implement:
\begin{itemize}[leftmargin=2em]
    \item \textbf{Byzantine fault tolerance:} A coalition of compromised agents can inject arbitrary events; detection relies on cryptographic hash chain breaks, not on consensus protocols.
    \item \textbf{Actor authentication:} The current trust model (Section~\ref{subsec:trust-model}) does not prevent impersonation; hash chains verify content integrity but not actor identity.
    \item \textbf{Real-time tamper response:} Integrity verification is a batch operation ($O(n)$) performed on demand, not a continuous monitoring process.
\end{itemize}
These limitations are addressed in Phase~3 (adversarial robustness) of the development roadmap.
