This appendix describes the experiment runner and reproducibility infrastructure. All experiments reported in this paper are executable via a unified runner: \texttt{python -m experiments.runner all}. The runner discovers experiments via the \texttt{@register} decorator in \texttt{experiments/registry.py}, executes each experiment, and generates a JSON summary.

Individual reproduction scripts are provided in \texttt{scripts/reproduce/}:
\begin{itemize}[leftmargin=2em]
    \item \texttt{reproduce.sh}: Runs all 21 experiments (E1--E21).
    \item \texttt{reproduce\_prov.sh}: Validates PROV-O export for all example concepts.
    \item \texttt{reproduce\_shacl.sh}: Validates exported Turtle against SHACL shapes.
    \item \texttt{reproduce\_sign.sh}: Verifies Ed25519 LD-Proof generation and verification.
    \item \texttt{reproduce\_rdfstar.sh}: Exports RDF-star quoted triples for all concepts.
    \item \texttt{reproduce\_adversarial.sh}: Runs adversarial integrity stress tests.
    \item \texttt{reproduce\_all.sh}: Master orchestrator that runs all sub-scripts and prints a unified PASS/FAIL summary.
\end{itemize}

Hardware specifications for all reported timings are documented in \path{docs/experiments/HARDWARE_SPECS.md}: Apple M2 (8P+4E cores), 16~GB RAM, Python 3.10, macOS 14.

\subsection{Reproducibility Package}

The complete reproducibility package is available at \url{[GitHub/DOI link]}. It contains:
\begin{enumerate}[leftmargin=2em]
    \item \texttt{requirements.txt}: Pinned versions of all dependencies (Pydantic, PyYAML, networkx, pytest, etc.).
    \item \texttt{experiments/}: Full experiment suite (E1--E21) with runner infrastructure.
    \item \texttt{data/aml/}: Data pipeline scripts to download the IBM AML HI-Small dataset and transform it to EventChains.
    \item \texttt{benchmarks/}: Head-to-head comparison drivers (E19: ADL Lite vs. nanopub vs. PROV-O vs. Git-only).
    \item \texttt{tests/}: Adversarial integrity test suite (A1--A12) with expected pass/fail outcomes.
    \item \texttt{Dockerfile}: One-command reproduction environment.
    \item \texttt{REPRODUCIBILITY.md}: Step-by-step instructions.
\end{enumerate}

\textbf{Double-blind review replication package.} During the anonymous review period, the artifact is available via an anonymous GitHub repository (\path{anonymous-4-open-science/adl-lite-repro}) and a persistent Zenodo deposit (DOI: 10.5281/zenodo.XXXXXX) with the author identity redacted. The repository contains the complete source code, all experiment scripts, the pre-processed AML dataset, and the LaTeX source of this paper. The Docker image \path{anonymous-4-open-science/adl-lite-repro} on Docker Hub reproduces all experiments in Table~\ref{tab:results} with one command: \texttt{docker run --rm anonymous-4-open-science/adl-lite-repro}. After acceptance, the repository will be transferred to the authors' institutional GitHub organization with a permanent DOI redirect. The replication package excludes only the (empty) \texttt{.adl/} directory that stores per-user calibration profiles and API keys, which are created automatically on first run.

\subsection{Docker Environment}

The provided \texttt{Dockerfile} builds a reproducible environment based on \texttt{python:3.10-slim} with all dependencies pre-installed. The image includes:
\begin{itemize}[leftmargin=2em]
    \item Python 3.10 with pinned dependencies from \texttt{requirements.txt}
    \item The IBM AML HI-Small dataset (pre-downloaded and cached)
    \item All experiment scripts and benchmark drivers
    \item A pre-built LaTeX environment (TeX Live 2024) for paper compilation
\end{itemize}

To reproduce all experiments:
\begin{verbatim}
docker build -t adl-lite-repro .
docker run --rm adl-lite-repro python -m experiments.runner all
\end{verbatim}

To reproduce a single experiment (e.g., E6):
\begin{verbatim}
docker run --rm adl-lite-repro python -m experiments.runner E6 --verbose
\end{verbatim}

To reproduce the head-to-head benchmark (E19):
\begin{verbatim}
docker run --rm adl-lite-repro python -m experiments.e19_governance_benchmark
\end{verbatim}

Expected runtime on Apple M2 (8P+4E cores), 16~GB RAM: ~5 minutes for E1--E21, ~15 minutes for E19 (includes nanopub and PROV-O setup), ~2 minutes for adversarial suite (A1--A12).

\subsection{Synthetic Event Injection Strategy (E6)}

The IBM AML HI-Small dataset provides raw transaction records, not capability-lifecycle events. To evaluate the EventChain architecture, we synthetically injected governance events that simulate a realistic multi-agent capability-lifecycle workflow. The injection strategy is as follows.

\textbf{Injection ratio.} Of the 9,300 total events across 201 chains, 96.8\% (9,000 events) are \textsc{Register} events derived directly from the raw transaction records (one \textsc{Register} per account). The remaining 3.2\% (300 events) are synthetically injected governance events distributed as shown in Table~\ref{tab:synthetic-events}.

\begin{table}[htbp]
\centering
\small
\caption{Synthetic event injection distribution for E6 (300 events across 201 chains).}
\label{tab:synthetic-events}
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Event Type} & \textbf{Count} & \textbf{Percentage} & \textbf{Confidence Distribution} \\
\midrule
\textsc{Validate} & 180 & 60.0\% & $\mathcal{U}[0.50, 0.95]$ \\
\textsc{Evidence} & 60 & 20.0\% & N/A (no confidence field) \\
\textsc{Relate} & 30 & 10.0\% & N/A (predicate-only) \\
\textsc{Deprecate} & 20 & 6.7\% & N/A \\
\textsc{Fork} & 10 & 3.3\% & N/A \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Injection logic.} For each account chain, we injected events according to the lifecycle transition matrix (Table~\ref{tab:lifecycle-matrix}): (i)~every chain receives at least one \textsc{Validate} event after the genesis \textsc{Register}; (ii)~10\% of validated chains receive a \textsc{Deprecate} event; (iii)~5\% of validated chains receive a \textsc{Fork} event; (iv)~\textsc{Evidence} and \textsc{Relate} events are interleaved at random indices between lifecycle events. Confidence values for \textsc{Validate} events are sampled uniformly from $[0.50, 0.95]$ to reflect realistic validator uncertainty, with a small fraction ($\approx$5\%) at $1.0$ to test the clamp boundary. The synthetic events are not intended to model real AML expert behavior; they are architectural stress tests that exercise the full lifecycle transition graph.

\textbf{Validation of injection realism.} The uniform confidence distribution is a conservative lower bound on validator diversity: real experts would likely show higher variance and domain-specific clustering. The 60/20/10/6.7/3.3 split is inspired by open-source software governance patterns (many validations, fewer deprecations, rare forks), not by AML domain data. Domain-level evaluation with human AML experts (E5) is in progress to replace these synthetic assumptions with empirical distributions.
