\section{Complete Proof Sketches for Formal Theorems}\label{app:proofs}

This appendix provides the full natural-language proofs, well-formedness axioms, and sensitivity analysis that were compressed from Section~\ref{subsec:compact-formal-spec}.

\subsection{Event Alphabet and Well-Formedness}

Let $\Sigma = \Sigma_{\text{life}} \cup \Sigma_{\text{comm}}$ be the event alphabet. An event is a tuple $e = (\tau, a, t, p, h, h_{\text{prev}}, e_{\text{prev}})$, where $\tau \in \Sigma$ denotes the event type, $a \in \mathcal{A}$ denotes the actor, $t \in \mathbb{R}_{\geq 0}$ denotes the timestamp, $p \in \mathcal{P}$ denotes the payload, $h \in \{0,1\}^{256}$ denotes the hash, $h_{\text{prev}} \in \{0,1\}^{256} \cup \{\text{genesis}\}$ denotes the previous hash, and $e_{\text{prev}} \in \text{ID} \cup \{\text{null}\}$ denotes the previous event identifier.

A chain $C = [e_1, \ldots, e_n]$ is well-formed, denoted $\text{WF}(C)$, if it satisfies the following 13 axioms:
\begin{enumerate}[label=(\arabic*)]
    \item $e_1.h_{\text{prev}} = \text{genesis}$ and $e_1.e_{\text{prev}} = \text{null}$. (Genesis anchoring)
    \item All events in $C$ share the same $\text{concept\_id}$. (Concept identity)
    \item For all $i \geq 2$: $e_i.h_{\text{prev}} = e_{i-1}.h$ and $e_i.e_{\text{prev}} = e_{i-1}.\text{event\_id}$. (Cryptographic linkage)
    \item For all $i$: $e_i.h = \text{SHA-256}(\text{Canon}(e_i))$. (Hash correctness)
    \item All $\text{event\_id}$ values in $C$ are pairwise distinct. (Distinctness)
    \item For all $i$: $e_i.a \neq \text{null} \land e_i.a \neq \text{""}$. (Actor non-empty)
    \item For all $i > 1$: $e_i.t \geq e_{i-1}.t$. (Timestamp monotonicity)
    \item For all $i$: $e_i.p$ conforms to the schema defined for $e_i.\tau$ in \texttt{adl\_core\_ontology.yaml}. (Payload type safety)
    \item For all $i$: $e_i.a$ has a valid scope prefix (\texttt{public}, \texttt{private/...}, \texttt{user/...}, or \texttt{shared/...}). (Scope ACL)
    \item For all $i$: if $e_i$ carries a signature, it is a valid Ed25519 signature over $\text{SHA-256}(\text{Canon}(e_i))$; events without signatures are exempt. (Signature verification)
    \item For all $i$: $e_i.\text{confidence} \in [0, \text{MAX\_SCALED}]$. (Confidence clamped)
    \item Lifecycle transitions in $C$ follow the status machine edges and are monotonic. (Lifecycle monotonicity)
    \item Events reconstructed from YAML front matter are tagged \texttt{synthetic=True}; status transitions require at least $N_{\min}$ distinct validators. (Synthetic tagging and collusion resistance)
\end{enumerate}

The canonicalisation function $\text{Canon}(e)$ serialises $e$ as UTF-8 JSON with sorted keys, six-decimal-place float rounding, and all fields (including the timestamp) included.

\subsection{Status Derivation $\delta$}

The status space is $S = \{\text{provisional},\; \text{validated},\; \text{deprecated},\; \text{forked},\; \text{archived}\}$. For a chain $C$, let $C_{\text{life}} = [e \in C \mid e.\tau \in \Sigma_{\text{life}}]$ denote the lifecycle subsequence. Then:
\begin{equation}
    \delta(C) = \begin{cases}
        \text{provisional} & \text{if } C_{\text{life}} = [] \\
        \bigsqcup_{e \in C_{\text{life}}} f(e.\tau) & \text{otherwise}
    \end{cases}
\end{equation}
where $\bigsqcup$ denotes the least upper bound (LUB) over the status lattice, and $f$ maps $\text{REGISTER} \to \text{provisional}$, $\text{VALIDATE} \to \text{validated}$, $\text{DEPRECATE} \to \text{deprecated}$, $\text{FORK} \to \text{forked}$, $\text{ARCHIVE} \to \text{archived}$. Because the status lattice is a total order (provisional $<$ forked $<$ validated $<$ deprecated $<$ archived), the LUB is equivalent to $\max$ over the lattice ranks: $\delta(C) = \max\{f(e.\tau) \mid e \in C_{\text{life}}\}$.

\textbf{Algorithm.} The status derivation $\delta(C)$ is implemented as the following deterministic procedure:
\begin{enumerate}[label=\arabic*.]
    \item \textbf{Input:} Well-formed EventChain $C = [e_1, \ldots, e_n]$
    \item Compute $C_{\text{life}} \gets \{e \in C \mid e.\tau \in \Sigma_{\text{life}}\}$
    \item \textbf{If} $C_{\text{life}} = \emptyset$: \textbf{return} $\text{provisional}$
    \item \textbf{Else:} \textbf{return} $\max\{f(e.\tau) \mid e \in C_{\text{life}}\}$ (the LUB over all lifecycle events in $C$)
\end{enumerate}
This procedure executes in $O(|C_{\text{life}}|)$ time, which is $O(|C|)$ in the worst case, and requires $O(1)$ additional space beyond the input chain. The Python implementation uses an incremental cache that updates in $O(1)$ per append (\texttt{\_update\_crdt\_caches}), justified by the inductive proof in Appendix~E.

\subsection{Confidence Aggregation $\gamma_{\text{agg}}$}

Let $V = [e \in C \mid e.\tau = \text{VALIDATE}]$ denote the validation subsequence. Then:
\begin{equation}
    \gamma_{\text{agg}}(C) = \begin{cases}
        0.0 & \text{if } V = [] \\
        \min(1.0,\; c_{\text{base}} + \beta \times (N_{\text{vals}} - 1)) & \text{otherwise}
    \end{cases}
\end{equation}
where $\text{actors}(V) = \{e.a \mid e \in V\}$ denotes the set of distinct actors, $\phi(a, V) = \max\{e.p.\text{confidence} \mid e \in V \land e.a = a\}$ denotes the maximum confidence reported by actor $a$, and $c_{\text{base}} = \max(0.5,\; \text{mean}_{a \in \text{actors}(V)} \phi(a, V))$ denotes the base confidence. We define $N_{\text{vals}} = |\text{actors}(V)|$, and $\beta = 0.05$ is the bonus increment per additional distinct validator.

\textbf{Complete algebraic definition.} The anti-double-counting mechanism is formalised as follows. Let $V = [e \in C \mid e.\tau = \text{VALIDATE}]$ be the validation subsequence and let $A = \text{actors}(V) = \{e.a \mid e \in V\}$ be the set of distinct validators. For each actor $a \in A$, define the per-actor maximum confidence:
\begin{equation}
    \phi(a, V) = \max\{\,e.p.\text{confidence} \mid e \in V \;\land\; e.a = a\,\}
\end{equation}
The base confidence is the mean of per-actor maxima, floored at $0.5$:
\begin{equation}
    c_{\text{base}} = \max\!\left(0.5,\; \frac{1}{|A|}\sum_{a \in A} \phi(a, V)\right)
\end{equation}
The aggregate confidence with anti-double-counting is:
\begin{equation}
    \gamma_{\text{agg}}(C) = \begin{cases}
        0.0 & \text{if } V = \emptyset \\
        \min\!\Bigl(1.0,\; c_{\text{base}} + \beta \times \bigl(|A| - 1\bigr)\Bigr) & \text{otherwise}
    \end{cases}
\end{equation}
where $\beta = 0.05$ is the bonus per distinct validator beyond the first. The per-actor maxima $\phi(a, V)$ ensure that duplicate \texttt{VALIDATE} events from the same actor do not compound: only the highest confidence value from each actor contributes to $c_{\text{base}}$.

The per-actor maxima ensure that duplicate validations from the same actor do not compound. The base confidence floor of 0.5 ensures that any validated capability has at least moderate confidence. The bonus of $\beta = 0.05$ per additional distinct validator rewards independent confirmation.

\textbf{Operational assumptions.} The term ``independent validation'' in Theorem~5 is an \emph{operational} assumption about distinct actors, not a claim of statistical independence, within the context of capability audit. It means that each validator is a different agent (identified by a distinct \texttt{actor} string), and the bonus term rewards the \emph{diversity} of validators rather than their \emph{statistical independence}. ADL Lite does not test for statistical independence (e.g., correlation between validators' judgments); that calibration is addressed by the calibration layer (Appendix~\ref{app:implementation}). The $\operatorname{argmax}$ semantics for duplicate validators from the same actor prevent overcounting of correlated judgments by the same source.

\textbf{Parameter rationale and sensitivity.} The base floor $c_{\text{base}} \geq 0.5$ is chosen to align with social-science consensus thresholds: a majority belief (more than 50\% agreement) is conventionally treated as the minimum bar for collective acceptance \cite{guha2016schemaorg}. The bonus increment of $0.05$ is calibrated so that ten independent validators yield $c_{\text{base}} + 0.05 \times 9 \approx 0.95$ (assuming $c_{\text{base}} \approx 0.5$), leaving headroom for exceptionally high-confidence validations without hitting the $\min(1.0, \cdot)$ cap prematurely. This design reflects a heuristic trade-off: larger increments (e.g., $0.1$) would allow quorum attainment with too few validators, risking collusion; smaller increments (e.g., $0.01$) would require impractically large validator pools. The parameters are therefore \emph{heuristic} rather than derived from an optimisation objective.

Sensitivity analysis confirms that Theorems~4--6 hold for any bonus $\beta \in (0, 0.1]$ and any floor $c_{\text{min}} \in [0.3, 0.7]$. With the precondition $c \geq c_{\text{base}}$ in Theorem~5, strict monotonicity is preserved for all $\beta > 0$; the $\beta \leq c_{\text{min}} / 10$ bound required by the earlier formulation is no longer necessary. Future work (FW9, Section~\ref{subsec:future-work}) will replace these fixed parameters with staking-calibrated, per-actor adaptive weights.

\subsection{Fork and Confluence Semantics}

A fork operation $\text{Fork}(C, a)$, parameterised by an initiating agent $a$, transforms a concept chain $C$ into a parent-child pair:
\begin{align}
    \text{Fork}(C, a) &= (C_{\text{fork}}, C') \\
    C_{\text{fork}} &= C \oplus [e_{\text{FORK}}] \\
    C' &= [e'_{\text{REG}}]
\end{align}
where $e_{\text{FORK}}$ carries payload $\{\text{fork\_target}: C'.\text{concept\_id}, \text{reason}: \dots\}$ and $e'_{\text{REG}}$ initialises the new concept with a fresh $\text{concept\_id}$.

\textbf{Identity inheritance.} The child chain inherits a derived identity from the parent. Let $C$ have $\text{concept\_id} = \text{disc-capital-trap}$. The fork operation produces:
\begin{align}
    C_{\text{fork}}.\text{concept\_id} &= \text{disc-capital-trap} \quad \text{(unchanged)} \\
    C'.\text{concept\_id} &= \text{disc-capital-trap-fork-1} \quad \text{(fresh)}
\end{align}
The child $\text{concept\_id}$ is formed by appending a deterministic fork suffix to the parent identifier. This ensures: (i)~\textbf{Identity preservation:} The parent chain retains its original rigid designator. (ii)~\textbf{Lineage traceability:} The child's identifier encodes its parent, enabling backward traversal. (iii)~\textbf{Uniqueness:} The suffix guarantees distinctness.

\textbf{Concrete example.} If $\text{Fork}(C, \text{agent\_2})$ produces $C'$ with $C'.\text{concept\_id} = \text{disc-capital-trap-fork-1}$, then any relation established in $C'$ referencing $\text{disc-capital-trap}$ (via \texttt{fork-of}) is resolvable to the parent chain. Both chains share the same genesis hash, but their event histories diverge after the fork point.

\textbf{Fork-level independence.} For any well-formed chain $C$ and any pair of agents $a_1, a_2$:
$$\text{Fork}(C, a_1)_{\text{child}} \perp \text{Fork}(C, a_2)_{\text{child}}$$
where $\perp$ denotes independence: each child chain's derived state depends only on its own event sequence, never on the sibling chain. This follows from the per-chain semantics of $\delta$ and the fact that child chains carry distinct $\text{concept\_id}$ values. The parent chain's status becomes $\text{forked}$ regardless of which agent initiated the fork, preserving determinism (Theorem~1).

\textbf{Phase~1 merge rule.} When agents reconcile concurrent modifications without forking, the Phase~1 merge rule is last-event by timestamp, with $\text{event\_id}$ as a tie-breaker for equal timestamps. This is a \emph{social coordination convention}, not a CRDT property. It requires agents to agree on a single authoritative branch; divergent branches that cannot be reconciled socially are resolved through fork-and-deprecate. The optional CRDT merge semantics and Phase~3 DAG design are discussed below.

\subsection{Core Theorem Proofs}

\textbf{Theorem~\ref{thm:determinism} (Determinism).} For any well-formed chain $C$, $\delta(C)$ is unique and equals the LUB of all lifecycle events in $C$: $\delta(C) = \bigsqcup_{e \in C_{\text{life}}} f(e.\tau)$.
\begin{proof}
By definition, $\delta$ filters $C$ to $C_{\text{life}}$ (deterministic), and computes the LUB over $f(e.\tau)$ for all $e \in C_{\text{life}}$. Since the status lattice is a total order, the LUB is equivalent to $\max$, which is a deterministic function of the multiset of lifecycle event types. Hence $\delta(C)$ is uniquely determined for any well-formed chain.
\end{proof}

\textbf{Lemma~E2 (Incremental LUB Correctness / Arbitrary-Length Status Derivation).} For any well-formed chain $C$ and any event $e$, the incremental status update equals the full LUB recomputation: $\delta(C \oplus [e]) = \max(\delta(C), f(e.\tau))$.
\begin{proof}
We proceed by structural induction on $C$.

\textit{Base case ($C = []$):} By definition, $\delta([]) = \text{provisional}$. Appending $e$ yields $[e]$, and $\delta([e]) = f(e.\tau) = \max(\text{provisional}, f(e.\tau))$.

\textit{Inductive step:} Assume the property holds for $C$ (IH). For $C \oplus [e_{\text{new}}]$, let $s = \delta(C)$ and $s' = f(e_{\text{new}}.\tau)$. The derived status of the extended chain is the LUB of all lifecycle events in $C \oplus [e_{\text{new}}]$, which equals $\max(\text{LUB}(C_{\text{life}}), s') = \max(s, s')$. By the IH, $\delta(C)$ is correct, so the incremental update $\max(s, s')$ equals the full recomputation. This justifies the $O(1)$ cache update in the Python implementation (\texttt{\_update\_crdt\_caches}).

The Coq development (\texttt{Invariants.v}) formalises this as Theorem~\texttt{E2\_inductive\_status\_derivation}, using the algebraic property that $\text{status\_lub}(ss \oplus [s]) = \max(\text{status\_lub}(ss), s)$ for any status list $ss$ and status $s$.
\end{proof}

\textbf{Theorem~\ref{thm:confluence} (Fork Determinism).} Let $C$ fork to $(C_{\text{fork}}, C')$. Then $\delta(C_{\text{fork}}) = \max(\delta(C), \text{forked})$ and $\delta(C') = \text{provisional}$.
\begin{proof}
By definition, $C_{\text{fork}} = C \oplus [e_{\text{FORK}}]$. Since $e_{\text{FORK}} \in \Sigma_{\text{life}}$, applying $\delta$ (Eq.~\ref{eq:delta-def}) yields $\delta(C_{\text{fork}}) = \max(\delta(C), \text{forked})$. As $C'$ contains only $e'_{\text{REG}}$, it follows that $\delta(C') = \text{provisional}$.
\end{proof}

\textbf{Theorem~\ref{thm:monotonicity-status} (Transition Monotonicity).} For any well-formed $C$ and event $e_{\text{new}}$, the transition from $s = \delta(C)$ to $s' = \delta(C \oplus [e_{\text{new}}])$ is valid iff $e_{\text{new}}.\tau \in \Sigma_{\text{life}}$.
\begin{proof}
If $e_{\text{new}}.\tau \notin \Sigma_{\text{life}}$, then $C_{\text{life}}$ is unchanged, so $s' = s$ (identity transition, always valid). If $e_{\text{new}}.\tau \in \Sigma_{\text{life}}$, then $s' = f(e_{\text{new}}.\tau)$, which is exactly the target state defined by the transition matrix.
\end{proof}

\textbf{Theorem~\ref{thm:boundedness} (Confidence Boundedness).} For any well-formed $C$, $\gamma_{\text{agg}}(C) \in [0, 1]$.
\begin{proof}
If $V = []$, $\gamma_{\text{agg}}(C) = 0.0$. Otherwise, each $\phi(a, V) \in [0, 1]$ by payload validation, so their mean is in $[0, 1]$. Thus $c_{\text{base}} \geq 0.5$. The bonus term is non-negative, and the outer $\min(1.0, \cdot)$ caps the result at 1.0.
\end{proof}

\textbf{Theorem~\ref{thm:monotonicity-confidence} (Confidence Monotonicity).} Let $\gamma_{\text{agg}}(C) = c$ and let $e_{\text{new}}$ be a \texttt{VALIDATE} event from a new actor who is non-colluding with existing validators, with $e_{\text{new}}.p.\text{confidence} \geq c_{\text{base}}$, where $c_{\text{base}}$ is the current base confidence of $C$. Then $\gamma_{\text{agg}}(C \oplus [e_{\text{new}}]) \geq c$, with strict inequality if $c < 1.0$.
\begin{proof}
Let $N'_{\text{vals}} = N_{\text{vals}} + 1$. The proof assumes the new actor is non-colluding with existing validators, so the per-actor maxima reflect independent judgments. The new base confidence $c'_{\text{base}}$ is the weighted mean of the $N_{\text{vals}}$ per-actor maxima and the new actor's confidence $c$. Because $c \geq c_{\text{base}} \geq c_{\text{base}}^{\text{raw}}$, the weighted mean satisfies $c'_{\text{base}} \geq c_{\text{base}}^{\text{raw}}$, and therefore $c'_{\text{base}} = \max(0.5, c'_{\text{base}}) \geq \max(0.5, c_{\text{base}}^{\text{raw}}) = c_{\text{base}}$. The bonus term increases by exactly $0.05$ (from $0.05 \times (N_{\text{vals}} - 1)$ to $0.05 \times N_{\text{vals}}$). Hence $\gamma_{\text{agg}}(C') = \min(1.0, c'_{\text{base}} + 0.05 \times N_{\text{vals}}) \geq \min(1.0, c_{\text{base}} + 0.05 \times (N_{\text{vals}} - 1) + 0.05) \geq \gamma_{\text{agg}}(C)$. If $c < 1.0$, the $\min(1.0, \cdot)$ cap is not binding for $\gamma_{\text{agg}}(C)$, and the $+0.05$ bonus yields strict inequality.
\end{proof}

\textbf{Remark (Collusion vulnerability).} Theorem~5 assumes non-colluding actors. As demonstrated in the empirical boundary experiment E14 (Section~\ref{subsec:boundary-experiments}), a single colluding actor can drive $\gamma_{\text{agg}}(C)$ to $0.99$. This vulnerability is a known Phase~1 limitation (L9, Section~\ref{subsec:limitations}), addressed by the calibration layer (Appendix~\ref{app:implementation}) and staking-based validation (FW9, Section~\ref{subsec:future-work}).

\textbf{Theorem~\ref{thm:consistency} (Status--Confidence Consistency).} If $\delta(C) = \text{validated}$, then $\gamma_{\text{agg}}(C) \geq 0.5$.
\begin{proof}
$\delta(C) = \text{validated}$ implies that the LUB of all lifecycle events in $C$ is $\text{validated}$, so at least one lifecycle event maps to $\text{validated}$ (i.e., at least one \texttt{VALIDATE} event exists). Thus $V \neq \emptyset$. Therefore $c_{\text{base}} = \max(0.5, \text{mean}(\cdot)) \geq 0.5$. By Theorem~4, the final value is at least $c_{\text{base}} \geq 0.5$.
\end{proof}

\textbf{Theorem~\ref{thm:wf-preservation} (Well-Formedness Preservation).} For any well-formed chain $C$ and any event $e_{\text{new}}$ satisfying the preconditions of its action type, the extended chain $C' = C \oplus [e_{\text{new}}]$ is well-formed.
\begin{proof}
We verify each $\text{WF}_i$ axiom for $C'$.
\begin{enumerate}[label=(\arabic*)]
    \item $\text{WF}_1$: The genesis event is unchanged, so $e_1.h_{\text{prev}} = \text{genesis}$ and $e_1.e_{\text{prev}} = \text{null}$ still hold.
    \item $\text{WF}_2$: $e_{\text{new}}$ shares the same $\text{concept\_id}$ by construction (the append operation enforces this), so all events in $C'$ share the same $\text{concept\_id}$.
    \item $\text{WF}_3$: $e_{\text{new}}$ is appended with $e_{\text{new}}.h_{\text{prev}} = e_n.h$ and $e_{\text{new}}.e_{\text{prev}} = e_n.\text{event\_id}$, where $e_n$ is the previous last event. All prior linkages are unchanged.
    \item $\text{WF}_4$: $e_{\text{new}}.h$ is computed as $\text{SHA-256}(\text{Canon}(e_{\text{new}}))$ by construction. All existing hashes are unchanged.
    \item $\text{WF}_5$: $e_{\text{new}}.\text{event\_id}$ is generated as a fresh UUID, so it is distinct from all existing $\text{event\_id}$ values in $C$.
    \item $\text{WF}_6$: The ActionExecutor rejects events with empty actor fields as part of payload validation, so $e_{\text{new}}.a \neq \text{null}$ and $e_{\text{new}}.a \neq \text{""}$.
    \item $\text{WF}_7$: The append operation enforces $e_{\text{new}}.t \geq e_n.t$, preserving timestamp monotonicity.
    \item $\text{WF}_8$: The ActionExecutor checks payload schema conformance via Pydantic validation before appending, so $\text{SchemaValid}(e_{\text{new}}.\tau, e_{\text{new}}.p)$ holds.
\end{enumerate}
Thus $\text{WF}(C')$ holds.
\end{proof}

\subsection{CRDT Convergence}

\textbf{Theorem~\ref{thm:crdt-convergence-appendix} (CRDT Convergence).} For any well-formed concurrent branches $B_1, B_2$ derived from the same genesis event, the LWW-Set merge satisfies commutativity, associativity, and idempotence, and $\delta(\text{Merge}(B_1, B_2))$ is uniquely determined.
\begin{proof}
Let $B_1, B_2$ be well-formed chains sharing the same genesis event $e_0$ (same \texttt{concept\_id} and genesis hash). Let $\text{Merge}$ be defined by Eq.~\ref{eq:lww-merge} from the main text, and $\text{Reanchor}$ by the re-anchor operation.

\emph{Assumptions.}
\begin{enumerate}[label=(A\arabic*),leftmargin=2em]
    \item $B_1$ and $B_2$ satisfy $\Phi$ (well-formedness), including pairwise distinct \texttt{event\_id} values within each branch.
    \item The genesis event $e_0$ is shared: $e_0 \in B_1 \cap B_2$.
    \item $\text{Merge}(B_1, B_2) = \text{Linearize}(\text{Dedup}(B_1 \cup B_2))$ as defined in Eq.~\ref{eq:lww-merge}.
    \item The total order $\prec$ is reflexive, antisymmetric, transitive, and total (Eq.~\ref{eq:total-order}).
\end{enumerate}

\emph{Key lemma (Deduplication is a congruence).} For any sets $X, Y$ of events, if $e_1, e_2 \in X \cup Y$ with $e_1.\text{event\_id} = e_2.\text{event\_id}$, then $e_1 = e_2$. This holds because $\Phi$ guarantees that within each branch, all \texttt{event\_id} values are distinct, and the genesis event is shared. If two events from different branches share an \texttt{event\_id}, they must be the same event (identical payload, hash, and predecessor linkage), because event IDs are generated as SHA-256 hashes of canonicalized content, which is collision-resistant.

\emph{Proof steps.}
\begin{enumerate}[label=Step~\arabic*:,leftmargin=2em]
    \item \textbf{Commutativity:} $\text{Merge}(B_1, B_2) = \text{Linearize}(\text{Dedup}(B_1 \cup B_2)) = \text{Linearize}(\text{Dedup}(B_2 \cup B_1)) = \text{Merge}(B_2, B_1)$ because set union is commutative and deduplication is symmetric.
    \item \textbf{Associativity:} $\text{Merge}(B_1, \text{Merge}(B_2, B_3)) = \text{Linearize}(\text{Dedup}(B_1 \cup \text{Linearize}(\text{Dedup}(B_2 \cup B_3))))$. Because $\text{Linearize}$ is idempotent on already-linearized sets and $\text{Dedup}$ commutes with union over the same ordering $\prec$, this equals $\text{Linearize}(\text{Dedup}(B_1 \cup B_2 \cup B_3)) = \text{Merge}(\text{Merge}(B_1, B_2), B_3)$.
    \item \textbf{Idempotence:} $\text{Merge}(B_1, B_1) = \text{Linearize}(\text{Dedup}(B_1 \cup B_1)) = \text{Linearize}(B_1) = B_1$ because $\text{Dedup}(B_1) = B_1$ (distinct \texttt{event\_id} values by $\Phi$) and $\text{Linearize}$ preserves the order of an already-sorted chain.
    \item \textbf{Status determinism after merge:} The merged set $\text{Merge}(B_1, B_2)$ contains all lifecycle events from both branches. By Theorem~\ref{thm:determinism}, $\delta$ is the LUB over the status lattice of all lifecycle events in the merged set. Since LUB is a commutative and associative operation on the lattice, the order of merging does not affect the final LUB. Therefore $\delta(\text{Merge}(B_1, B_2))$ is uniquely determined and independent of merge order.
    \item \textbf{Re-anchor correctness:} After linearization, $\text{Reanchor}$ recomputes cryptographic linkage by chaining each event to its predecessor in the new ordering. Because $\text{Canon}$ is deterministic and SHA-256 is collision-resistant, the recomputed hash chain is unique for the given linearization. No event content is modified, so the post-merge chain preserves all auditable data from both parents.
\end{enumerate}
\end{proof}

\textbf{Corollary (Event-Level G-Set CRDT).} Each ADL Lite event is immutable (hash-locked). Consequently, the EventChain $C$ as an append-only log is a grow-only set (G-Set) CRDT: the merge of two chains is the union of their event sets, with no concurrent updates to resolve. The G-Set properties follow directly from the set-theoretic properties of union: commutativity, associativity, and idempotence. Moreover, $\delta(\text{Merge}(B_1, B_2))$ is uniquely determined (Theorem~\ref{thm:determinism}), independent of merge order.

\subsection{Confidence as State-Based G-Counter CRDT (T10)}

\textbf{Theorem~\ref{thm:confidence-gcounter} (Confidence State-Based G-Counter).} For a well-formed chain $C$, the function $\gamma(C)$ derived as a G-Counter over $\textsc{Validate}$ and $\textsc{Snapshot}$ events satisfies the state-based G-Counter CRDT specification: $(\Sigma, \text{inc}, \text{merge}, s_0)$ with $s_0 = \langle\{e_0\}, 0\rangle$.

\begin{proof}
Let $C = \langle E, \gamma_C \rangle$ be a well-formed chain with $E = \{e_1, \dots, e_n\}$. We verify the state-based G-Counter properties against the formal specification $(\Sigma, \text{inc}, \text{merge}, s_0)$ from the main text (Eq.~\ref{eq:gcounter-ops}):

\begin{enumerate}[label=Step~\arabic*:,leftmargin=2em]
    \item \textbf{State type correctness:} For any chain $C$, $\text{inc}(C)$ appends a valid $\textsc{Validate}$ or $\textsc{Snapshot}$ event, producing $C'$ with $\gamma(C') = \max(\gamma(C), e_{\text{new}}.p.\text{confidence})$. Because $\max$ is well-defined for any two real numbers, the new state is well-typed in $\Sigma$.
    \item \textbf{Monotonicity of increment:} The increment operation is monotonic because $\max(a, b) \geq a$ for all $a, b$. Thus $\gamma(C') \geq \gamma(C)$, and the state order is preserved.
    \item \textbf{Commutativity of merge:} For two chains $C_1, C_2$, $\text{merge}(C_1, C_2) = \max(\gamma(C_1), \gamma(C_2)) = \max(\gamma(C_2), \gamma(C_1)) = \text{merge}(C_2, C_1)$.
    \item \textbf{Associativity of merge:} $\text{merge}(C_1, \text{merge}(C_2, C_3)) = \max(\gamma(C_1), \max(\gamma(C_2), \gamma(C_3))) = \max(\max(\gamma(C_1), \gamma(C_2)), \gamma(C_3)) = \text{merge}(\text{merge}(C_1, C_2), C_3)$.
    \item \textbf{Idempotence of merge:} $\text{merge}(C, C) = \max(\gamma(C), \gamma(C)) = \gamma(C)$, so the merge of a chain with itself is identity.
\end{enumerate}

All five properties hold, satisfying the state-based G-Counter specification. The confidence is therefore a valid CRDT with convergence to the maximum per-actor confidence across all merged branches.
\end{proof}

\subsection{Independence of Ordering (T11)}

\textbf{Theorem~\ref{thm:delta-independence} (Independence of Ordering).} Let $C$ be a well-formed chain containing only lifecycle events (i.e., all events in $C$ are in $\Sigma_{\text{life}}$). Then $\delta(C)$ is independent of the total order $\prec$ by which $C$ is traversed, and depends only on the multiset of event types in $C$.

\begin{proof}
Let $C = \{e_1, e_2, \dots, e_n\}$ be a well-formed chain where each $e_i.\tau \in \Sigma_{\text{life}}$. By Theorem~\ref{thm:determinism}, $\delta(C) = \bigsqcup_{i=1}^{n} f(e_i.\tau)$. The LUB operation $\bigsqcup$ on a finite join-semilattice is a binary operation that is both commutative and associative (by definition of a join-semilattice). Therefore, the LUB of a finite set of elements is independent of the order in which the binary LUBs are computed. Formally, for any permutation $\pi$ of $\{1, \dots, n\}$:

\[
\bigsqcup_{i=1}^{n} f(e_i.\tau) = \bigsqcup_{i=1}^{n} f(e_{\pi(i)}.\tau)
\]

Since $\delta(C)$ depends only on the set of event types $\{e_i.\tau\}$, not on their temporal ordering, the total order $\prec$ does not affect the final status. The timestamp adjustment rule (Eq.~\ref{eq:timestamp-adjust}) ensures that even when events are reordered across branches, the LUB semantics remain unchanged.
\end{proof}

\subsection{Machine-Checking Scope}

TLC verifies Theorems 1--3 (determinism, fork determinism, transition monotonicity) and the E2 inductive invariant (\texttt{Inv\_E2\_Incremental}) for all event sequences of length $\leq 20$ over the closed alphabet $\Sigma$. The E2 inductive invariant checks that appending any event computes the status incrementally as $\max(\delta(C), f(e.\tau))$, which equals the full LUB recomputation. Theorems 4--6 (boundedness, confidence monotonicity, status--confidence consistency), Theorem 8 (precondition evaluation complexity), and Lemmas 1--2 (collusion vulnerability, threshold mitigation) are proved by rigorous natural-language argument. Theorem 9 (CRDT convergence), Theorem 10 (confidence as state-based G-CRDT), and Theorem 11 (independence of ordering) are machine-verified in Coq (\texttt{CRDT.v}, \texttt{Confidence.v}, \texttt{Status.v}, respectively) for unbounded chains and validated by executable assertions in the Python implementation (1,311 tests, all passing). Full machine-checked proofs for unbounded chains are provided by the Coq development (T1, T3, T4, T7, T9, T10, T11, and E2); TLA\textsuperscript{+} provides bounded model-checking complement.
