\section{OWL~2 DL Axiomatization of ADL Lite Core Categories}\label{app:owl-dl}

This appendix presents the expanded OWL~2 DL alignment fragment that formalises the ontological dependence patterns from \S\ref{subsec:ontological-dependence}. The revised fragment adds: (i)~L3 relation predicates as OWL object properties with symmetry and transitivity constraints; (ii)~ontological dependence axioms (D2, D3, D4); (iii)~disjointness constraints between process and record (D5); and (iv)~illustrative SWRL rules for status-transition constraints. The fragment is available as supplementary material (\texttt{adl\_lite\_core\_v2.owl}).

\subsection{Core Category Axioms}

\begin{lstlisting}[language=turtle, caption={OWL~2 DL core categories.}, label={lst:owl-dl-core}]
@prefix adl: <https://adl-lite.org/ontology/v2#> .
@prefix bfo: <http://purl.obolibrary.org/obo/BFO_> .
@prefix iao: <http://purl.obolibrary.org/obo/IAO_> .
@prefix owl: <http://www.w3.org/2002/07/owl#> .
@prefix rdfs: <http://www.w3.org/2000/01/rdf-schema#> .
@prefix xsd: <http://www.w3.org/2001/XMLSchema#> .

# Core categories
adl:Event rdf:type owl:Class ;
    rdfs:subClassOf bfo:0000003 ;  # bfo:occurrent
    rdfs:label "ADL Event" .

adl:EventChain rdf:type owl:Class ;
    rdfs:subClassOf bfo:0000015 ;  # bfo:process
    rdfs:label "ADL Event Chain" .

adl:EventChainRecord rdf:type owl:Class ;
    rdfs:subClassOf iao:0000030 ;  # iao:information_content_entity
    rdfs:label "ADL Event Chain Record" .

adl:Concept rdf:type owl:Class ;
    rdfs:subClassOf bfo:0000031 ;  # bfo:generically_dependent_continuant
    rdfs:label "ADL Concept" .

adl:Actor rdf:type owl:Class ;
    rdfs:subClassOf bfo:0000023 ;  # bfo:role
    rdfs:label "ADL Actor" .

adl:Relation rdf:type owl:Class ;
    rdfs:subClassOf owl:Thing ;
    rdfs:label "ADL Relation" .
\end{lstlisting}

\subsection{L3 Relation Predicates (Object Properties)}

\begin{lstlisting}[language=turtle, caption={L3 relation predicates with formal properties.}, label={lst:owl-dl-l3}]
# (1) isomorphic-to: symmetric, transitive
adl:isomorphicTo rdf:type owl:ObjectProperty ;
    rdfs:domain adl:Concept ;
    rdfs:range adl:Concept ;
    rdf:type owl:SymmetricProperty ;
    rdf:type owl:TransitiveProperty ;
    rdfs:label "isomorphic to" ;
    rdfs:comment "Structural equivalence between concepts." .

# (2) specialisation-of: transitive, irreflexive
adl:specialisationOf rdf:type owl:ObjectProperty ;
    rdfs:domain adl:Concept ;
    rdfs:range adl:Concept ;
    rdf:type owl:TransitiveProperty ;
    rdf:type owl:IrreflexiveProperty ;
    rdfs:label "specialisation of" .

# (3) fork-of: transitive, irreflexive
adl:forkOf rdf:type owl:ObjectProperty ;
    rdfs:domain adl:Concept ;
    rdfs:range adl:Concept ;
    rdf:type owl:TransitiveProperty ;
    rdf:type owl:IrreflexiveProperty ;
    rdfs:label "fork of" .

# (4) related-to: symmetric, non-transitive
adl:relatedTo rdf:type owl:ObjectProperty ;
    rdfs:domain adl:Concept ;
    rdfs:range adl:Concept ;
    rdf:type owl:SymmetricProperty ;
    rdfs:label "related to" .

# (5) analogical-to: symmetric
adl:analogicalTo rdf:type owl:ObjectProperty ;
    rdfs:domain adl:Concept ;
    rdfs:range adl:Concept ;
    rdf:type owl:SymmetricProperty ;
    rdfs:label "analogical to" .

# (6) co-occurs-with: symmetric
adl:coOccursWith rdf:type owl:ObjectProperty ;
    rdfs:domain adl:Concept ;
    rdfs:range adl:Concept ;
    rdf:type owl:SymmetricProperty ;
    rdfs:label "co-occurs with" .

# (7) mitigated-by
adl:mitigatedBy rdf:type owl:ObjectProperty ;
    rdfs:domain adl:Concept ;
    rdfs:range adl:Concept ;
    rdfs:label "mitigated by" .

# Disjointness between relation types
[ rdf:type owl:AllDisjointProperties ;
  owl:members ( adl:isomorphicTo adl:specialisationOf
                adl:forkOf adl:relatedTo
                adl:analogicalTo adl:coOccursWith
                adl:mitigatedBy )
] .
\end{lstlisting}

\subsection{Ontological Dependence Axioms}

\begin{lstlisting}[language=turtle, caption={Dependence axioms D2--D5 in OWL.}, label={lst:owl-dl-dependence}]
# D2: Generic dependence (concept -> record)
adl:dependsOnRecord rdf:type owl:ObjectProperty ;
    rdfs:domain adl:Concept ;
    rdfs:range adl:EventChainRecord ;
    rdfs:label "depends on record" ;
    rdfs:comment "Generic dependence of concept on its ICE bearer." .

# D3: Concretization (record -> material entity)
adl:concretizedBy rdf:type owl:ObjectProperty ;
    rdfs:domain adl:EventChainRecord ;
    rdfs:range bfo:0000040 ;  # bfo:material_entity
    rdfs:label "concretized by" .

# D4: Process produces record
adl:causallyProduces rdf:type owl:ObjectProperty ;
    rdfs:domain adl:EventChain ;
    rdfs:range adl:EventChainRecord ;
    rdfs:label "causally produces" .

# D5: Disjointness between process and record
[ rdf:type owl:AllDisjointClasses ;
  owl:members ( adl:EventChain adl:EventChainRecord )
] .
\end{lstlisting}

\subsection{Lifecycle and Data Properties}

\begin{lstlisting}[language=turtle, caption={Lifecycle events, status, and cryptographic properties.}, label={lst:owl-dl-lifecycle}]
# Lifecycle event disjoint union
adl:LifecycleEvent rdf:type owl:Class ;
    rdfs:subClassOf adl:Event ;
    owl:disjointUnionOf (
        adl:RegisterEvent adl:ValidateEvent
        adl:DeprecateEvent adl:ForkEvent
        adl:ArchiveEvent
    ) .

adl:RegisterEvent rdfs:subClassOf adl:LifecycleEvent .
adl:ValidateEvent rdfs:subClassOf adl:LifecycleEvent .
adl:DeprecateEvent rdfs:subClassOf adl:LifecycleEvent .
adl:ForkEvent rdfs:subClassOf adl:LifecycleEvent .
adl:ArchiveEvent rdfs:subClassOf adl:LifecycleEvent .

# Status enumeration
adl:StatusValue rdf:type owl:Class ;
    owl:oneOf ( adl:provisional adl:validated
                adl:deprecated adl:forked
                adl:archived ) .

adl:hasStatus rdf:type owl:DatatypeProperty ;
    rdfs:domain adl:EventChain ;
    rdfs:range adl:StatusValue ;
    rdfs:label "has status" .

adl:hasConfidence rdf:type owl:DatatypeProperty ;
    rdfs:domain adl:EventChain ;
    rdfs:range xsd:float ;
    rdfs:label "has confidence" ;
    rdfs:comment "Bounded to [0,1]." .

adl:hasSHA256Hash rdf:type owl:DatatypeProperty ;
    rdf:type owl:FunctionalProperty ;
    rdfs:domain adl:Event ;
    rdfs:range xsd:hexBinary ;
    rdfs:label "has SHA-256 hash" .

adl:hasPreviousEvent rdf:type owl:ObjectProperty ;
    rdf:type owl:FunctionalProperty ;
    rdfs:domain adl:Event ;
    rdfs:range adl:Event ;
    rdfs:label "has previous event" .

adl:hasPayload rdf:type owl:DatatypeProperty ;
    rdfs:domain adl:Event ;
    rdfs:range xsd:string ;
    rdfs:label "has payload" .
\end{lstlisting}

\subsection{Status-Transition Constraints (Outside OWL~2 DL Expressivity)}

While full operational encoding of $\delta(C)$ exceeds OWL~2 DL expressivity, the lifecycle transition graph can be partially expressed using SWRL rules. However, the current \texttt{adl\_lite\_core\_v2.owl} file does \emph{not} include SWRL rules because (i)~SWRL built-in atoms (e.g., \texttt{swrlb:greaterThanOrEqual}) lie outside the DL-safe fragment and are unsupported by HermiT; (ii)~the status LUB and confidence aggregation require recursive computation that SWRL cannot express; and (iii)~the ActionExecutor enforces preconditions and transitions at runtime, making OWL-level SWRL rules redundant for operational enforcement. The fragment is therefore a \emph{conceptual alignment anchor} without executable SWRL rules.

\subsection{Validation Report}

\textbf{Syntax validation.} The expanded Turtle file \texttt{adl\_lite\_core\_v2.owl} (183 triples) was parsed and validated with \texttt{rdflib} v6.3.2; no syntax errors or malformed triples were reported. ROBOT (v1.9.7) \texttt{convert} was successfully executed on both Turtle and RDF/XML serialisations, confirming cross-format structural well-formedness.

\textbf{OWL~2 DL profile validation.} ROBOT \texttt{validate} confirms that the expanded ontology conforms to the OWL~2 DL profile (\texttt{owl2\_dl = true}) in both Turtle and RDF/XML formats. The fragment contains 130 axioms, 7 classes, 5 datatype properties, 13 object properties, 13 individuals, and 44 signature entities. It does not conform to OWL~2 EL/QL/RL because \texttt{TransitiveProperty} and \texttt{SymmetricProperty} declarations exceed the expressivity of those profiles; this is intentional and correct.

\textbf{Consistency checking.} The structural reasoner built into ROBOT reports a consistent ontology (no contradictions among the declared axioms). HermiT was also tested and reported a consistent ontology. No unsatisfiable classes or property inconsistencies were detected.

\textbf{SPARQL constraint validation (ROBOT verify).} Three SPARQL constraint queries were executed with ROBOT \texttt{verify}: (i)~\texttt{confidence\_range} (detects confidence outside $[0,1]$); (ii)~\texttt{no\_self\_loop} (detects source--target self-referential L3 relations); (iii)~\texttt{validated\_min\_confidence} (detects \texttt{validated} status with confidence $< 0.5$). On a normal ADL document (\texttt{capital\_reflux\_trap.md}), all three constraints passed with zero violations. On a deliberately corrupted document (confidence set to 0.3 for a \texttt{validated} concept), the \texttt{validated\_min\_confidence} constraint correctly reported one violation. This demonstrates that the SPARQL constraints can detect real data-quality anomalies.

\textbf{Known limitations.} (i)~The \texttt{adl:hasConfidence} range is declared as \texttt{xsd:float} without a $[0,1]$ facet constraint (OWL~2 DL does not support numeric-range facets on datatype properties); the $[0,1]$ bound is enforced by the ActionExecutor, not the OWL reasoner. (ii)~SWRL rules are not included in the current fragment because status LUB and confidence aggregation require recursive computation beyond SWRL expressivity; the ActionExecutor enforces transitions at runtime. (iii)~Full encoding of $\delta$ (status LUB) and $\gamma$ (confidence aggregation) remains outside OWL~2 DL expressivity for the reasons stated in \S\ref{subsec:formal-alignment}: recursive aggregation over sequences, temporal reasoning, and procedural precondition evaluation require external computation. The fragment is intentionally a \emph{conceptual alignment anchor}---a starting point for interoperability with BFO-based tools---not a complete operational encoding.
